---
title: "Continous Data - Notes"
author: "Wouter van Amsterdam"
date: 2018-01-08
output: html_document
---

<!-- The file analysis/chunks.R contains chunks that define default settings
shared across the workflowr files. -->
```{r read-chunk, include=FALSE, cache=FALSE}
knitr::read_chunk("chunks.R")
```

<!-- Update knitr chunk options -->
```{r knitr-opts-chunk, include=FALSE}
```

<!-- Insert the date the file was last updated -->
```{r last-updated, echo=FALSE, results='asis'}
```

<!-- Insert the code version (Git commit SHA1) if Git repository exists and R
 package git2r is installed -->
```{r code-version, echo=FALSE, results='asis'}
```

<!-- Add your analysis here -->
## General info

### Coordinator
Cas Kruitwagen

### Tutor
Jan van den Broek
j.vandenbroek@uu.nl

### Course info
Computer lab login
account: dooMID01 - dooMID32
passwd: Methdata\$\$

## Continous data

### Model types

independent variable | type of model
--- | ---
Grouping | ANOVA
Continuous | Regression
Both grouping and continous | Analysis of covariance

All models can be referred to as linear models.

### Notation

* observation $j$ in group $i$: $y_{ij}$
* group mean: $\bar{y}_i$
* overall mean: $\bar{y}$
* group effect: $\mu_i - \mu$
* residual: $\epsilon_{ij}$

For a good model: the model 'generated' the data


### ANOVA model

$$y_{ij} = \mu + (\mu_i - \mu) + \epsilon_{ij}$$

Where $y_{ij}$ are normally distributed with mean $\mu_i$ and variance $\sigma^2$.

Using **estimators**:

$$y_{ij} = \bar{y} + (\bar{y}_i - \bar{y}) + \epsilon_{ij}$$

In terms of **deviations**

$$y_{ij} - \bar{y} = (\bar{y}_i - \bar{y}) + \epsilon_{ij}$$

Total deviation = group deviation + residual

#### Get sums of squares

$$\sum_{i,j}{(y_{ij}-\bar{y})^2} = \sum_{i,j}{(\bar{y}_i-\bar{y})}+\sum_{i,j}\epsilon_{ij}^2$$

In words:

$$SS_{Total} = SS_{Group} + SS_{Residual}$$

The degrees of freedom:

* $df_{Total} = n-1$
* $df_{Groups} = n_{groups} - 1$
* $df_{Residual} = df_{Total} - df_{Groups} = n - n_{groups} - 2$

Mean sums of squares = **Variances**:

$$MS_. = SS_./df_.$$

If groups have the same means, than variance between groups is equal to 
residual variance

$$F = \frac{MS_{Groups}}{MS_{Residual}} \approx 1$$

Looking at summary tables: not just verifying if the group means are different,
 but if the data fits the model.
 
### Two-way ANOVA

One-way = grouping can only be done 1 way.

(Full)-factorial design: for more than 1 factors, all levels are present in all factors


$$y_{ijk} = \bar{y} + (\bar{y_i} - \bar{y}) + (\bar{y}_j - \bar{y}) + \epsilon_{ijk}$$

Now:

$$SS_{total} = SS_{treatment} + SS_{dose} + SS_{residual}$$

Also works when grouping variables are non independent

#### With some simulated data

```{r}
require(magrittr)
set.seed(2)
n = 100
eff1 = .1
eff2 = .2
resSD = 0.1
treatment = rbinom(n = n, size = 1, prob = .5)
dose = rbinom(n = n, size = 1, prob = .5)

y = eff1*treatment + eff2*dose + rnorm(n, sd = resSD)

aov(y~treatment+dose) %>% summary()

```

## Regression

Model

$$y_i = \beta_0+\beta_x * x_i + \epsilon_i$$

Assumes:

* linear dependence of $y$ on $x$.
* overall variance $\sigma^2$ (does not depend on $x$)

### Minimize sum of squares

$$\min_{\beta_0, \beta_x}{(y_i-(\beta_0+\beta_x * x_i))^2}$$

Solution by differentiating with respect to parameters and set to zero.

$$\beta_x = \frac{\sum_{i}(x_i-\bar{x})(y_i-\bar{y})}{\sum_i(x_i-\bar{x})^2}$$

And 

$$\beta_0 = \bar{y} - \beta_x*\bar{x}$$

### Get an ANOVA table:

$$y_i = \beta_0 + \beta_x *x + \epsilon_{i}$$

So 

$$y_i - \bar{y} = \beta_0 -\bar{y}+ \beta_x *x + \epsilon_{i}$$

$$= -\beta_x*\bar{x}+ \beta_x *x + \epsilon_{i}$$

$$= \beta_x*(x-\bar{x}) + \epsilon_{i}$$


Regression: 1 DF, no slope is no regression

#### Simulated data

```{r}
set.seed(1)
n = 100
x = runif(n)
b = 0.15
rs = 0.3
y = b*x + rnorm(n, sd=rs)

plot(x,y, xlim = c(0,1), ylim = c(floor(min(y)), ceiling(max(y))))

fit <- lm(y~x)
summary(fit)
abline(fit, lty=2)

```


## Analysis of co-variance (ANCOVA)

Given both a grouping variable and a regression variable

Assuming:

* equal slopes in the groups
* only intercepts are different
* $y_{ij}$ is normally distributed with mean $\alpha_i + \beta_x * x_{ij}$
* with $\alpha_i = \bar{y} + (\bar{y}_i-\bar{y})$

$$y_{ij} = \bar{y} + (\bar{y}_i-\bar{y}) + \beta_x*x_{ij} + \epsilon_{ij}$$

Continous independent variable is also called the co-variate

#### Simulate data

```{r}
set.seed(3)
n = 100

econt  = 0.1
egroup = 0.2
rs     = 0.3
ngroups= 4

xcont  = runif(n)
xgroup = sample(0:(ngroups-1), replace = T, size = n)

y = econt * xcont + egroup * xgroup + rnorm(n, sd = rs)

fit <- glm(y~xcont+factor(xgroup))

summary(fit)
aov(fit)
```



## Interaction

When intercepts are different between groups

Degrees of freedom for interaction: df group * df regression

#### Simulate data

```{r}
set.seed(3)
n = 100

econt  = 0.1
egroup = 0.2
rs     = 0.3
ngroups= 4

xcont  = runif(n)
xgroup = sample(0:(ngroups-1), replace = T, size = n)

y = econt * xcont + egroup * xgroup + rnorm(n, sd = rs)

fit <- glm(y~xcont+factor(xgroup)+xcont:factor(xgroup))

summary(fit)
aov(fit)

```

When a continous variable is only included as an interaction term, 
it is called nesting. So the continous variable is nested in the 
group variable.

## Session information

<!-- Insert the session information into the document -->
```{r session-info}
```
