---
title: "Estimating with uncertainty"
author: "Wouter van Amsterdam"
date: 2017-10-24
output: html_document
---

<!-- The file analysis/chunks.R contains chunks that define default settings
shared across the workflowr files. -->
```{r read-chunk, include=FALSE, cache=FALSE}
knitr::read_chunk("chunks.R")
```

<!-- Update knitr chunk options -->
```{r knitr-opts-chunk, include=FALSE}
```

<!-- Insert the date the file was last updated -->
```{r last-updated, echo=FALSE, results='asis'}
```

<!-- Insert the code version (Git commit SHA1) if Git repository exists and R
 package git2r is installed -->
```{r code-version, echo=FALSE, results='asis'}
```

<!-- Add your analysis here -->
## Estimating with uncertainty
Tutor

Standard deviation
Standard error of the mean

Descriptive statistics
Inferential statistics


```{r, message = F, eval = F}
require(GGally)
data("mtcars")
fit1 <- lm(mpg~., data = mtcars)
fit2 <- lm(mpg~wt+am, data = mtcars)
summary(fit1)
summary(fit2)
ggcoef(fit1)

str(broom::tidy(fit1, conf.int = T, conf.level = .95))

# rbindlist
lapply(list(fit1, fit2), function(x)
{broom::tidy(fit1, conf.int = T, conf.level = .95)}
  )


```

**Estimation therory**

* Unbiasedness: $E(estimator) = parameter$; e.g. sample mean, median
* Consistency: When sample size goes to infinity, the variance goes to zero $\lim_{n\to\infty} Var(estimate)=0$; sample mean, sample median both consistent
* Efficiency: smallest variance of all available unbiased estimators; sample mean is most efficient, more efficient than median.

For variance:
population variance

$$Var_{population} = \frac{1}{n}\sum_{i}^{n}{(x_i-\mu)^2}$$

$$Var_{sample} = \frac{1}{n-1}\sum_{i}^{n}{(x_i-\mu)^2}$$

Sample variance is an unbiased estimator for the population variance.


Standard error of an estimator is the stnadard deviation of the estimator's sampling distribution.

**Standard error of estimator**
$$SE = \frac{\sigma}{\sqrt{n}}$$
**Estimator of standard error**
$$SE = \frac{s}{\sqrt{n}}$$
**Rough approximation of 95% confidence interval**
$$mean \pm 2*SE$$

#### Central limit theorem
If $Y_{1}$, $Y_{2}$, ..., $Y_{n}$ are independent, identically distributed random variables, 
from any distribution, with mean $\mu$ and variance $\sigma^2$ then, if $n$ is large enough,
 the mean of these varibles $\bar{Y}$, follows approximately a normal distribution with 
 mean $\mu$ and variance $\frac{\sigma^2}{n}$.

If we standardize $Y$, then $$\frac{Y-\mu}{\frac{\sigma}{\sqrt{n}}}$$ follows a standard normal distribution. 
* Approximately 2/3 of the *standardized* sample means should be within -1 and 1
* Approximately 95% of the *standardized* sample means should be within -2 and 2

If $\sigma^2$ is unknown, and the sample variance is estamated with $s^2$, than 
$$\frac{Y-\mu}{\frac{s}{\sqrt{n}}}$$ follows the Student's t-distribution.

The $100*(1-\alpha)\%$ confidence interval is given by:
$$\bar{Y} - t_{1-\frac{\alpha}{2}, n-1}*\frac{s}{\sqrt{n}} < \mu < \bar{Y} + t_{1-\frac{\alpha}{2}, n-1}*\frac{s}{\sqrt{n}}$$

#### Combining samples
##### Equal variance of groups, unpaired measurements
Difference of mean
$$SE_{difference} = s_{pooled}*\sqrt{\frac{1}{n_{1}}+\frac{1}{n_{2}}}$$

With

$$s_{pooled} = \frac{(n_{1}-1)*s_{1}^2 + (n_{2}-1)*s_{2}^2}{n_{1}+n_{2}-2}$$

Choose critical value from T distribution with $n_{1} + n_{2} - 2$ degrees of freedom.

95% confidence interval:
$$\bar{Y_{1}}-\bar{Y_{2}} \pm t_{0.025, n_{1}+n_{2}-2}*SE_{difference}$$

##### Unequal variance, unpaired measurements
$$t = \frac{\bar{Y_{1}} - \bar{Y_{2}}}{s_{\bar{\Delta}}}$$
Where
$$s_{\bar{\Delta}} = \sqrt{\frac{s_{1}^2}{n_{1}} + \frac{s_{2}^2}{n_{2}}}$$

Then, for the T-distribution, the degrees of freedom is estimated by the Welch-Satterthwaite equation
$$df = \frac{(\frac{s_{1}^2}{n_{1}} + \frac{s_{2}^2}{n_{2}})^2}{(\frac{s_{1}^2}{n_{1}})^2/(n_{1}-1) + (\frac{s_{2}^2}{n_{2}})^2/(n_{2}-1)}$$

## Remarks
Example of old vs young persons with balance for calculating pooled variance, 
because it is unlikely that variances are equal between the groups.

## Session information

<!-- Insert the session information into the document -->
```{r session-info}
```
