---
title: "Count data and Poisson Regression"
author: "Wouter van Amsterdam"
date: 2018-01-12
output: html_document
---

<!-- The file analysis/chunks.R contains chunks that define default settings
shared across the workflowr files. -->
```{r read-chunk, include=FALSE, cache=FALSE}
knitr::read_chunk("chunks.R")
```

<!-- Update knitr chunk options -->
```{r knitr-opts-chunk, include=FALSE}
```

<!-- Insert the date the file was last updated -->
```{r last-updated, echo=FALSE, results='asis'}
```

<!-- Insert the code version (Git commit SHA1) if Git repository exists and R
 package git2r is installed -->
```{r code-version, echo=FALSE, results='asis'}
```

<!-- Add your analysis here -->

## Intro

Tutor: Jan van den Broek

### Profile likelihood

Marginal likelihood of a parameter

### Poisson distribution

$$\ln(\mu) = linear\ model$$

$$P(Y_i = y_i) = \frac{e^{-\mu}\mu^{y_i}}{y_i!}$$

In rare disease situation, no offset necessary
Grouped binomial data will not work, due to numerical 
problems


$$L(\mu) = \prod_i{\frac{e^{-\mu}\mu^{y_i}}{y_i!}}$$

This gives

$$l(\mu) = \sum_i{[-\mu + y_i\ln(\mu)-ln(y_i!)]}$$

### With offset
With offset for population this is interpreted as a risk:

$$\ln(\frac{\mu}{population}) = linear\ model$$

With offset for person time, it is interpreted as rates.

## Exponential family

* normal
* gamma: concentrations
* log-normal: concentrations / normal
* inversed gaussian

* exponential
* chi-squared
* beta
* Dirichlet
* Bernoulli
* categorical
* Poisson
* Wishart
* inverse Wishart
* geometric

### With restrictions

* binomial (fixed number of trials)
* multinomial (fixed number of trials)
* negative binomial (fixed number of trials)

### GLM

* distribution from exponential family
* linear predictor
* there is a link function from linear predictor to distribution


Remark: only for normal distribution the variance is independent of the mean.

When there is a limited range (like strictly positive), the normal distribution is not appropriate

## On science

Chamberlain's method of multiple working hypothesis

Instead of a single hypothesis.


Likelihood connects models with data.
All things that make biological sense can be used.


## Session information

<!-- Insert the session information into the document -->
```{r session-info}
```
