---
title: "Analysis of longitudinal data"
author: "Wouter van Amsterdam"
date: 2018-01-17
output: html_document
---

<!-- The file analysis/chunks.R contains chunks that define default settings
shared across the workflowr files. -->
```{r read-chunk, include=FALSE, cache=FALSE}
knitr::read_chunk("chunks.R")
```

<!-- Update knitr chunk options -->
```{r knitr-opts-chunk, include=FALSE}
```

<!-- Insert the date the file was last updated -->
```{r last-updated, echo=FALSE, results='asis'}
```

<!-- Insert the code version (Git commit SHA1) if Git repository exists and R
 package git2r is installed -->
```{r code-version, echo=FALSE, results='asis'}
```

<!-- Add your analysis here -->

## Intro

Tutor: Cas Kruitwagen

## Independent and dependent sources of variation

E.g. multiple measurements in time for a single subject

### Paired T-test

For 2 time-points

### Split-plot ANOVA

Target: divide variation (sums of squares): e.g. time and persons

Gives estimate of error between persons, and within persons over time

Here there is not really an interaction between persons and time,
it is used as a random effect.



**Regression models ignoring the time dependency tend to overestimate
the standard errors of time-varying covariates, 
and underestimate the standard errors of time-invariant covariates.**

#### Assumptions

  1. Residuals are normally distributed
  2. Covariance matrices for different levels of between subject factors are equal. (extension of homoscedasticity, variances and covariances at different 
  time points are assumed to be equal for all 'cells'); sphericity
  3. Compound symmetry: variances are equal at all time points. 
  Correlations are equal for any two time points

For 3 timepoints, the covariance matrix of a single group is:

$$
\left(\begin{array}{ccc} 
var_1 & cov_{1,2} & cov_{1,3}\\
cov_{2,1} & var_2 & cov_{2,3}\\
cov_{3,1} & cov_{3,2} & var_3
\end{array}\right)
$$ 

Assumption 2: covariance matrix any form, but 'symmetric positive definite'

Robust to variance ratio < 2, 3 between groups

Assumption 3:

$$
\sigma^2\left(\begin{array}{ccc} 
1 & \rho & \rho\\
\rho & 1 & \rho\\
\rho & \rho & 1
\end{array}\right)
$$ 

Assumption 3 is strict, it assumes that the correlation between 
close-in-time observations is equal to the correlation between 
far-away-in-time observations


### Linear Mixed Effects models (MLE)

Use 
  - fixed effects (like diet and time, what you can reproduce by design)
  - random effects (like participants), assumed to follow a known distribution


Advantages over split-plot ANOVA

  1. LME can handle missing value
  2. Variance need not be the same at each time point (can be modelled separately)
  3. Correlations between various time points may vary (can be modelled,
  e.g. correlation between observations decreases when time differences increases)


### With data

```{r}
require(dplyr)
data("ChickWeight")
str(ChickWeight)
```

```{r}
length(unique(ChickWeight$Chick))
unique(ChickWeight$Time)
```

#### Take only 2 timepoints

Get only 2 timepoints

```{r}
chick <- ChickWeight %>% filter(Time %in% c(0, 2))
head(chick)
```

T-test with only time

```{r}
t.test(weight~Time, data = chick)
```

```{r}
boxplot(weight~Diet+Time, data = chick)
```

```{r}
require(ggplot2)
chick %>%
  group_by(Time, Diet) %>%
  summarize(weight = mean(weight)) %>%
  ggplot(aes(x = Time, y = weight, col = Diet)) + geom_line()
```

Time * diet
```{r}
summary(aov(weight~Time*Diet, data = chick))
```

Time * Diet and chicks

```{r}
summary(aov(weight~Time*Diet + Error(Chick), data = chick))
```

#### All timepoints

```{r}
ChickWeight %>%
  group_by(Diet, Time) %>%
  summarize(weight = mean(weight)) %>%
  ggplot(aes(x = Time, col = Diet, y = weight)) + geom_line()
```

```{r}
aov(weight~Time*Diet + Error(Chick), data = ChickWeight) %>%
  summary()
```

Use factor on time to eliminate linear assumption

```{r}
aov(weight~factor(Time)*Diet + Error(Chick), data = ChickWeight) %>%
  summary()
```

```{r}
fit <- lm(weight~Time*Diet, data = ChickWeight)
summary(fit)
```


```{r}
require(tidyr)
mat <- ChickWeight %>% 
  filter(Diet == 1) %>% 
  data.table::dcast(Chick~Time, value.var = "weight") %>%
  select(-Chick) %>%
  as.matrix()
```

#### LME

Generate an intercept for each subject

```{r}
require(nlme)
fit <- lme(fixed = weight~Diet*factor(Time), 
           random = ~1|Chick, data = ChickWeight)
anova(fit)
```


## Session information

<!-- Insert the session information into the document -->
```{r session-info}
```
