---
title: "Assignments week 3"
author: "Wouter van Amsterdam"
date: 2017-11-06
output: html_document
---

<!-- The file analysis/chunks.R contains chunks that define default settings
shared across the workflowr files. -->
```{r read-chunk, include=FALSE, cache=FALSE}
knitr::read_chunk("chunks.R")
```

<!-- Update knitr chunk options -->
```{r knitr-opts-chunk, include=FALSE}
```

<!-- Insert the date the file was last updated -->
```{r last-updated, echo=FALSE, results='asis'}
```

<!-- Insert the code version (Git commit SHA1) if Git repository exists and R
 package git2r is installed -->
```{r code-version, echo=FALSE, results='asis'}
```

<!-- Add your analysis here -->

## Day 9 Multiple regression
### Exercises with R
> *Introduction*
> Below you will find a worked example that helps you understand how to perform model building in multiple regression and analysis of covariance in R.

> We are interested in the diastolic blood pressure (y) for people on two different treatments (group). Within the two treatments different dosages were given. First type in the data (or Copy-Paste):

```{r}
y <- c(87,86.5,89,88.5,87.5,88,86.5,87,85,86,85,83)
dose <- c(5,6,7,8,9,10,5,6,7,8,9,10)
group <- c(0,0,0,0,0,0,1,1,1,1,1,1)
```

Let???s take a look at a plot of the data: 

```{r}
interaction.plot(dose, group, y , mean, ylab = "Bloodpressure")
```


> Use `help(interaction.plot)` to see how it works. R uses ":" to denote an interaction so group:dose is the R interaction term in a model. 

> It looks like there may be an interaction in the data: group 0 has higher blood pressure levels than group 1 in the higher dosages. If there is an explanatory variable in the data file that is categorical (other than 0-1), then you should tell R this by using the function factor(). So factor(group) tells R that group is not a numeric variable but that its numbers should be used as group labels. To fit an ANOVA model you can use either of the following:

```{r}
model.an <- lm(y~group) 
model.an <- glm(y~group, family = gaussian)
```

> In the second statement, the `family=gaussian` part may be left out since the gaussian (normal) is the default (???GLM??? stands for generalized linear model, of which ANOVA, linear regression, and logistic regression models are special cases). The result of the ANOVA model will be stored in model.an. In the second case, this will be an object of glm-type because you used glm to create it. To see what is in it use `names(model.an)` and if you want to see something specific use for instance `model.an$coefficients`. To get the table with the estimates: 

```{r}
summary(model.an)
```

> The function `drop1(fit, test = "F")` looks at the variables in the model ???fit???, then leaves out the terms one by one and calculates the F-test for every term if it were to be left out. Of course, in `model.an` there is only one variable so you just get one test. Note that the p-value of the F-test is exactly the same as the p-value of the t-test in the model summary.
To fit the model without the interaction: 

```{r}
model.anc <- glm(y~group + dose, family = gaussian)
summary(model.anc) 
drop1(model.anc, test = "F") 
```

> The column "Deviance" contains the residual sums of squares for different models. The first line gives the residual sums of squares if none of the terms is dropped so for the model with both group and dose in it. The second line gives the residual sum of squares for the model without group so for the model with only dose in it. The difference in these residual sums of squares gives the sum of squares for the group: 29.310-12.976 = 16.334. In the same way the sum of squares for dose can be obtained (14.583-12.976 = 1.607). For dose and group the F-values and the p-values are shown. With this information an ANOVA table could be constructed.

This somewhat elaborate method is simplified by using the Anova function in the library car:

```{r}
library(car) #Note: you might have to install this library first, using Packages, Install packages 
Anova(model.anc, test.statistic = "F")
```


> The model with an interaction can be fitted as: (or the exact same model can be given by:)

```{r}
model.int <- glm(y~group + dose + group:dose, family=gaussian)
model.int <- glm(y~group*dose, family = gaussian)
```

> Note that if you now use the drop1 function, only the interaction will be evaluated for possible dropping:

```{r}
drop1(model.int, test = "F")
Anova(model.int, test.statistic = "F")
```

> R does this because it makes little sense to drop a main effect while the interaction is still in the model; generally, one first checks whether the interaction can be removed.
The interaction term is statistically significant, so the trend in blood pressure over the dosages is different for the two treatment groups.

**this should be somewhere else, as `full` does not exist yet
> Checking Multicollinearity (used on day 10)
The variance inflation factor can be obtained using the vif() function in the car package. The argument to the vif() function is a model you have already fit.

> vif(full)
      temp  factories population       wind   rainfall   daysrain 
  3.763553  14.703175  14.340318   1.255460   3.404904   3.443932

> To get the tolerance instead, you can invert the VIF:
> 1/vif(full)

#### A note on automatic variable selection in R
> R does not have the same type of forward, backward and stepwise selection procedures as SPSS. The add1 and drop1 can be used to examine variables and decide which variable should be added/dropped next. The actual adding and dropping is not done automatically and needs to be done by the analyst, so a new model is fitted and again checked for variables that can be added/dropped. Note that the add1 and drop1 functions both give Akaike???s Information Criterion (AIC, to be treated during Modern Methods in Data Analysis) by default; an F-test can be obtained by using the option test="F" in the command. 

#### 5.
> This is a repeat of exercise #1, but now in R. Compare the results with those obtained in SPSS. 
The dataset with SO2 data from the lecture is available in the dataset so2.RData. Try to repeat the findings from the lecture notes using R. How will you do the variable selection in R?

```{r}
load(epistats::fromParentDir("data/so2.RData"))
str(so2)
```

First some histograms of the numeric variables

```{r}
# find out which variables in so2 are numeric
num_vars <- colnames(so2)[sapply(so2, is.numeric)]

# how many are there
length(num_vars)

par(mfrow = c(2, 4))
for (variable in num_vars) {
  hist(so2[[variable]], main = variable, xlab = variable)
}
  
par(mfrow = c(1, 1))
```

Then the pairwise scatterplots

```{r}
pairs(so2)
```

Correlation plot of all numeric variables (this can be done nicely with the package `corrplot`)
```{r}
cor_matrix <- cor(so2[, num_vars])
corrplot::corrplot(cor_matrix, method = "number")
```

Now for the model building. We can use the `step` function for variable selection. 
Note that it uses the Akaike Information criterion for variable selection.
NB we must remove the `city` variable, as this is merely a label of the observations, and is unique for each row.
```{r}
model_vars <- setdiff(colnames(so2), "city") # take all colnames of so2, remove 'city' from these names
fit0 <- lm(SO2~1, data = so2[, model_vars])
fit_full <- lm(SO2~., data = so2[, model_vars]) # use ~. to include all variables except for SO2

steps_forward  <- step(fit0, scope = list(lower = fit0, upper = fit_full), direction = "forward")
steps_backward <- step(fit_full, data = so2[, model_vars], direction = "backward")
steps_stepwise <- step(fit0, scope = list(upper = fit_full), data = so2[, model_vars], direction = "both")
```

The function by default prints all the steps. I do not know how to stop this behaviour

Look at the final models:

```{r}
summary(steps_forward)
summary(steps_backward)
summary(steps_stepwise)
```

#### 6. Cigarettes
> This is a repeat of exercise #2, but now in R. Compare the results with those obtained in SPSS. 
The workspace cigarette.RData contains a dataset cigarettte with data on carbon monoxide, tar and nicotine contents and weight of 25 brands of cigarettes.
We want to predict the carbon monoxide contents using the other 3 variables.
a.	Make a scatter plot matrix of the 4 variables, and formulate which variables you expect to predict (part of) carbon monoxide content.

NB the `RData` file did not seem to contain any data, so we imported the SPSS file with package `foreign`
```{r}
epistats::fromParentDir("data/cigarette.RData")
cigarette <- foreign::read.spss(epistats::fromParentDir("data/cigarette.sav"))
cigarette <- as.data.frame(cigarette)
# save file:
# save(cigarette, file = epistats::fromParentDir("data/cigarette2.RData"))
pairs(cigarette)
```

Nicotine looks highly correlated with tar and carbonmonoxide. Carbonmonoxide 
looks highly correlated with tar too.

> b.	Now make a correlation matrix of the 4 variables, and check your expectations from a.

```{r}
corrplot::corrplot(cor(cigarette), method = "number")
```

> c.	Build a regression model with all 3 predictor variables. Are all variables significant? Are regression coefficients what you would expect? Can you think of an explanation?

```{r}
fit_full <- lm(carbonmonoxide~., data = cigarette)
summary(fit_full)
```

Only tar is significant. If the explanatory variables were independent, we would 
expect that nicotine was alsa correlated with carbonmonoxide. However, due 
to coliniearity, the effect of nicotine vanishes when tar is included in the model.

> d.	Using backward selection reduce the model from c until it contains only significant variables. Which variable(s) are in the final model? Which proportion of the variation in carbon monoxide content is explained by this model?

Let's do manual backward selection

```{r}
fit_full <- lm(carbonmonoxide~., data = cigarette)
summary(fit_full)
```

Use `drop1` to determine which variable to drop first. 
Remove the coefficient with the highest p-value

```{r}
drop1(fit_full, test = "F")
fit_1 <- lm(carbonmonoxide~tar+nicotine, data = cigarette)
drop1(fit_1, test = "F")
```

Drop the next variable

```{r}
fit_2 <- lm(carbonmonoxide ~ tar, data = cigarette)
drop1(fit_2, test = "F")
```

Now we cant remove anymore predictors, because 'all' are significant.
Check assumptions of homoscedasticity and normal distribution of residuals:

```{r}
plot(fit_2, which = c(1,2))
```

Residuals look pretty OK. Homoscedasticity is a little hard to judge, but at least there is no clear funnel shape.

> e.	Based on the backward selection model, what is the predicted carbon monoxide content of a cigarette with tar = 13.0, nicotine = 1.0 and weight = 1.0?  What is its 95% prediction interval, and how do you interpret this?
	(Use the R function predict to obtain predictions for new data based on the model:
	
```{r}

new <- data.frame(tar=13.0, nicotine=1.0, weight=1.0)
predict(fit_2, newdata=new, interval="prediction", level=0.95)

```

#### 7.
> This is a repeat of exercise #4, but now in R. Compare the results with those obtained in SPSS. 
The variables in the study of 38 stream sites in New York state by Lovett et al. (2000) fell into two groups measured at different spatial sites ??? watershed variables (elevation, stream length and area) and chemical variables for a site averaged across sampling dates (averaged over 3 years). We use only the chemical variables. The data are given in the data file stream.RData

> STREAM 	name of the stream (site) from which observations were collected
MAXELEV 	maximum elevation of stream (m above sea level)
SAMPELEV 	site elevation (m above sea level)
LENGTH 	length of stream
AREA 	area of watershed
NO3 	concentration (mmol/L) of nitrogen oxide ions 
TON 	concentration (mmol/L) of total organic nitrogen 
TN 	concentration (mmol/L) of total nitrogen 
NH4 	concentration (mmol/L) of ammonia ions 
DOC 	concentration (mmol/L) of dissolved oxygen 
SO4 	concentration (mmol/L) of sulphur dioxide ions 
CL 	concentration (mmol/L) of chloride ions 
CA 	concentration (mmol/L) of calcium ions 
MG 	concentration (mmol/L) of magnesium ions 
H 	concentration (mmol/L) of hydrogen ions 

> Which of the chemical variables can predict  the maximum elevation of the stream?
Lovett et al. have used the log of the variables DOC, CL and H in their analyses. Can you imagine why they did it and is it necessary? 

```{r}
load(epistats::fromParentDir("data/stream.RData"))
str(stream)
```

All variables are numeric, except for `STREAM`, which is the name of the site.
Lets remove this variable to make our lives easier

```{r}
streams <- stream$STREAM
stream$STREAM <- NULL

car::scatterplotMatrix(stream[, c(1:7)], diagonal = "histogram")
car::scatterplotMatrix(stream[, c(8:14)], diagonal = "histogram")

```

It looks like `LENGTH` and `AREA` are tightly correlated, like `NO3` and `TN`,
 also `SO4` and `MG`. Note that many scatterplots were not included due to the limited 
 plot area.
 
We can formalize this by sorting the correlations

```{r}
# create correlation matrix
cor_matrix <- cor(stream)

# to remove the uninformative diagonal, and duplicity, retain only upper triangle
cor_matrix[lower.tri(cor_matrix, diag = T)] <- NA

# to analyze this, 'melt' the data to a conveniant format
cor_melted <- data.table::melt(cor_matrix, value.name = "correlation")

# remove the NA values
cor_melted <- cor_melted[!is.na(cor_melted$correlation),]
head(cor_melted)

# add a column with absolute correlation
cor_melted$abs_cor <- abs(cor_melted$correlation)

# sort by that column
cor_melted[order(cor_melted$abs_cor, decreasing = T), ][1:10,]
```


Zoom in on only `DOC`, `CL` and `H`
```{r}
car::scatterplotMatrix(stream[, c("MAXELEV", "DOC", "CL", "H")], diagonal = "histogram")
```

These 3 variables are right skewed, which is probably why they were log-transformed

Look at the transformed variables

```{r}
require(dplyr)
stream %>%
  mutate(DOC = log(DOC),
         CL = log(CL),
         H = log(H)) %>%
  select(c(MAXELEV, DOC, CL, H)) %>%
  car::scatterplotMatrix(diagonal = "histogram")
```

The distributions look a little nicer now. However, for linear regression,
normality of the independent variables is not assumed, only a linear relation 
between the independent variables and the dependent variable.

To answer the question which variables predict the elevation, lets use the 
`step` function with stepwise selection. And lets keep the transformed variables

```{r}
stream$DOC = log(stream$DOC)
stream$CL  = log(stream$CL)
stream$H   = log(stream$H)

fit0 = lm(MAXELEV ~ 1, data = stream)
fit_all = lm(MAXELEV ~., data = stream)

fit_step <- step(fit0, scope = list(upper = fit_all), data = stream, direction = "both")
```

Look at the final model:

```{r}
summary(fit_step)
```

Note that the `CL` variable remains included, even though we would drop this 
variable when using the p-values based on the F-statistic.

Of the variable pairs that were tightly correlated, in none of the cases 
both variables were included.

Check assumptions of the model

```{r}
plot(fit_step, which = c(1,2))
```


Residuals are a little skewed, not too much heteroscedasticity.


## Day 10 model building issues
### Exercises with R
> See the notes at the beginning of Exercises with R on day 9, especially how to check for multicollinearity.

#### 7.	
> This is a repeat of exercise #1, but now in R.
(Note that where SPSS calculates the “tolerance” for the excluded variables, R calculates it for the variables in the model.)
a.	Build a regression model with all 3 predictors. What is the tolerance for weight in this model? (Use the command 1/vif(), see previous day.)

```{r}
fit <- lm(carbonmonoxide~., data = cigarette)
1/vif(fit)
```

> b.	Now do a multiple regression with weight as dependent variable and tar and nicotine as predictors. What is the R2 of this model? How does it relate to the Tolerance?

```{r}
fit_weight <- lm(weight~tar+nicotine, data = cigarette)
summary(fit_weight)
```

The $R^2$ from this fit is the same as the tolerance from `tar` from above.

> c. 	Now what is the tolerance of nicotine in the model with only tar and nicotine? Check this, analogously to what you did in b. 

```{r}
fit2 <- lm(carbonmonoxide~tar+nicotine, data = cigarette)
1/vif(fit2)
```

The tolerance is hardly changed, so it looks like `nicotine` was pretty 
independent of `weight` in our sample.

#### 8.	
> This is a repeat of exercise #2, but now in R.
An indicator of a tree’s production capacity is the canopy. We want to examine whether there is a relation between the canopy and the production for fruit trees, and whether fertilizer has an influence on the relation. To answer this question, data on the canopy and production were collected from 14 available trees (6 fertilized and 8 unfertilized). The data are stored in crownarea.RData .
a.	Test whether the coefficient for the slope in the two groups (fertilized & unfertilized) is the same; i.e. are the lines for the two groups parallel or not?

```{r}
load(epistats::fromParentDir("data/crownarea.RData"))
str(crownarea)
```

The variable classes are pretty uncommon for R. Probably due to transferring 
data from SPSS to R. Let's polish them up.

```{r}
crownarea$FERTILIZ <- factor(crownarea$FERTILIZ)
crownarea$PRODUCTI <- as.numeric(crownarea$PRODUCTI)
crownarea$CROWNARE <- as.numeric(crownarea$CROWNARE)
```



```{r}
fit <- lm(PRODUCTI~CROWNARE*FERTILIZ, data = crownarea)
summary(fit)
```

There does not seem to be a statistically significant interaction between 
`FERTILIZ` and `CROWNARE`. Let's check in a plot:

```{r}
require(ggplot2)
ggplot(crownarea, aes(x = CROWNARE, y = PRODUCTI, col = FERTILIZ)) + 
  geom_point() + 
  geom_smooth(method = "lm")

```

The intercepts look different, but the slopes look the same (so no interaction)

> b.	Can the relation between canopy and production be described by one and the same line for both groups (fertilized & unfertilized)?

```{r}
fit2 <- lm(PRODUCTI~CROWNARE + FERTILIZ, data = crownarea)
summary(fit2)
```

Although there seems to be no significant interaction, the intercept for both 
groups is pretty different, and significant in the model fit.

> c.	Interpret the results.

Fertilization type seems to have an additive effect to canopy. For every level 
of canopy, fertilization status adds (or substracts, based on the reference category)
 the same amount of production.

> 9.	This is a repeat of exercise #3, but now in R. 
Read in the data file lowbirth.dat using the command

```{r}
lb <- read.table(epistats::fromParentDir("data/lowbirth.dat"), header = T)
str(lb)
```

> We are only going to use the following variables:
age 		age of the mother in years,
ht		history of hypertension (1 = yes, 0 = no) and 
bwt 		birth weight in grams

> a.	Fit a model with bwt as dependent and with ht and age as independent variables. Also include the interaction between age and ht in the model.

First off, lets recode `ht` as a factor variable.

```{r}
lb$ht <- factor(lb$ht)
fit <- lm(bwt ~ age*ht, data = lb)
summary(fit)
plot(fit, which = c(1,2))
```

Assumptions of homoscedasticity and normality of residuals seem ok.

Plot the data:
```{r}
require(ggplot2)
ggplot(lb, aes(x = age, y = bwt, col = ht)) + 
  geom_point() + 
  geom_smooth(method = "lm")
```



> b.	Give an interpretation of the interaction term.

It looks like age itself is not a significant predictor, while hypertension is.
However, in the group with hypertension, age seems to be an important factor. 
For participants with hypertension, increasing age is associated with a lower 
birthweight.

> c.	Check whether you need the interaction term in the model.

```{r}
summary(fit)
```

From the summary we see that the interaction term is significant. 
Let's also compare $R_adj^2$ between this model and a model without the interaction.

```{r}
summary(lm(bwt~age+ht, data = lb))
```

The model with interaction term looks better

> d.	Can the model be reduced further? Give the final model, the one that cannot be reduced further, and interpret this model.

We cannot reduce further. The interaction needs to be included, so also the 
intercepts for hypertension groups and the $\beta$ for age.

#### 10.
> FEV (forced expiratory volume) is an index of pulmonary function that measures the volume of air expelled after one second of constant effort. The dataset fev.txt contains FEV measurements on 654 children, aged 6-22, who were seen in the Childhood Respiratory Disease Study in 1980 in East Boston, Massachusetts. The question of interest is whether a child’s smoking status has an effect on his FEV.

> a.	Read the text file into R, using fev <- read.table(choose.files()). Results of earlier analyses indicate that logfev is a better measure to use in linear regression. Make a new variable fev$logFEV <- log(fev$FEV).

```{r}
fev <- read.table(epistats::fromParentDir("data/fev.txt"), header = T)
fev$logFEV <- log(fev$FEV)
str(fev)
```


> b.	Make a box plot of logfev by smoking status. What do you notice? What might be an explanation for this unexpected result? Test the difference in mean logfev between smokers and non-smokers using a t-test.

```{r}
boxplot(logFEV~Smoker, data = fev)
```

FEV looks higher in current smokers. Maybe this is mediated by age? 
Older children tend to smoke more often.

```{r}
t.test(logFEV~Smoker, data = fev)
```

> c.	Use linear regression to model logfev as a function of smoking status (note that the variable smoker cannot be used in regression; make a new variable smoke that takes the value 1 if the child is a current smoker and 0 if not); compare the result with part b).

```{r}
fit <- lm(logFEV~Smoker, data = fev)
summary(fit)
```

> d.	We suspect that age and height are distorting the relation between smoking and FEV. Look at some graphs of logfev, age, height and smoking. Are age and height related to FEV? To smoking?

```{r}
suppressWarnings(
  car::scatterplotMatrix(fev[, c("logFEV", "Age", "Height")], diagonal = "histogram")
)

par(mfrow = c(1,3))
boxplot(logFEV~Smoker, data = fev)
boxplot(Age~Smoker, data = fev)
boxplot(Height~Smoker, data = fev)
par(mfrow = c(1,1))
```

Or in a single call to `ggpairs` from the `GGally` package:

```{r, warning = F}
GGally::ggpairs(fev[, c("logFEV", "Age", "Height", "Smoker")])
```

> e.	Fit the following models:
i	logfev  ~ smoke
ii	logfev ~ smoke + age
iii	logfev  smoke + age + height
What is the interpretation of the coefficients in the third model? What happens to the coefficient for smoke and its standard error when age and height are added to the model?

```{r}
fit0 <- lm(logFEV~Smoker, data = fev)
fit1 <- lm(logFEV~Smoker + Age, data = fev)
fit2 <- lm(logFEV~Smoker + Age + Height, data = fev)
summary(fit0)
summary(fit1)
summary(fit2)
```

The interpretation of the coefficients:

* For equal Age and Height, non-smokers have 0.05 more ml FEV
* For equal smoking status and height, FEV increases with 0.02 ml for 
every year in age
* For equal smoking status and age, FEV increase with 0.04ml for each unit 
increase in height (probably inches, since it's the USA)

> f.	Since we saw an indication of non-parallel lines for smokers and non-smokers in the relation between logfev and age, fit a fourth model, including the interaction of age and smoking:
iv	logfev = smoke, age, height, agesmoke
What happens now to the coefficient for smoke and its standard error?

```{r}
fit3 <- lm(logFEV~Smoker + Age + Height + Age*Smoker, data = fev)
summary(fit3)
```

The coefficient for smoke goes from positive to negative, but the standard error 
increases and it is no longer significant.

> g.	Now let’s try a backward regression. Make one block with all the variables in model iv, plus sex. Choose Method=Backward to allow SPSS to drop non-significant variables. What do you get? Is this a logical model?

```{r}
fit_full <- lm(logFEV~Age*Smoker + Height + Sex, data = fev)
drop1(fit_full, test = "F")
```

It looks like the interaction term can be disregarded

```{r}
fit_a <- lm(logFEV~Age + Smoker + Height + Sex, data = fev)
drop1(fit_a, test = "F")
```

Now we would keep all variables. So FEV increases with age, 'not smoking', 
height and male sex. Seems like a reasonable model. Let's check assumptions:

```{r}
plot(fit_a, which = c(1,2))
```

Nice homoscedasticity, residuals maybe a little skewed

```{r}
hist(fit_a$residuals)
```

#### 11.
> Data from a study of the effect of three different methods of instruction on reading  comprehension in children are stored in the file readingtestscores.dat. 66 participants (22 in each group) were given a reading comprehension test before and after receiving the instruction. The following variables were collected:

> Subject: Subject number 
Group: Type of instruction that student received (Basal, DRTA, or Strat) 
PRE1: Pretest score on first reading comprehension measure 
PRE2: Pretest score on second reading comprehension measure 
POST1: Posttest score on first reading comprehension measure 
POST2: Posttest score on second reading comprehension measure 
POST3: Posttest score on third reading comprehension measure

> a.	The dataset is a flat text file with fixed widths. You can read it into R by using the read.table() function. 

It looks like the values are actually separated by a tab. Which can be read by 
using `sep = "\t"`.

```{r}
comp <- read.table(epistats::fromParentDir("data/readingtestscores.dat"), header = T, sep = "\t")

str(comp)
```

> b.	Make a boxplot of the posttest score on the second reading comprehension measure for the three types of instruction. Do you think there are statistically significant differences between the groups? 

```{r}
boxplot(POST2~Group, data = comp)
```

It looks like a pretty big difference, I would expect statistical significance.

> c.	Fit a model to examine whether there are differences in the mean posttest score on the second reading comprehension measure for the three types of instruction.

```{r}
fit <- lm(POST2~Group, data = comp)
summary(fit)
```

This summary containts the F-test, equivalent to performing an ANOVA. 
It also comes up with a post-hoc test for testing difference between a baseline 
group and the other groups.

> d.	Test whether or not there are differences between the groups. Which hypothesis are you testing?

Since we have no prior hypothesis on which group should score better or worse, 
it is best to do an ANOVA with proper post-hoc tests that incorporate adjustment 
for multiple comparisons among all groups. 
$H_0$: all groups equal, $H_1$: not all groups are equal.


```{r}
TukeyHSD(aov(POST2~Group, data = comp))

```


So DRTA and Basal are equavalent. Strat is better than both of these

> e.	This study is an example of a randomized, pre-post design. There are several ways to analyze such a design (analyzing only the post measurement, as we did above; analyzing the difference between pre and post; or using analysis of covariance), but “ANCOVA analysis” (a linear regression model with post scores at the outcome, and group and pre-intervention score as explanatory variables) has a number of advantages* over other analyses. Fit a model to examine whether there are differences in the mean posttest score on the second reading comprehension measure for the three types of instruction, controlling for the posttest score on the same test. Use the drop1() function to get p-values for the two variables in the model.

Probably what is meant in the question is 'controlling for the pre-test score'

```{r}
fit2 <- lm(POST2 ~ Group + PRE2, data = comp)
summary(fit2)
```


> f.	Get the model coefficients from the model in (e) and interpret them.

```{r}
coef(fit)
```

For equal `PRE2` score, the group with Strat instruction had 2.90 higher score 
on the second post-test evaluation. 

> g.	Get a summary of the model frm (c), and compare the mean squared error (called the “Residual standard error” in the R output) to that of the model in (e). What do you notice?

```{r}
summary(fit)
summary(fit2)
```

It looks like the the second model, including pre-test scores, fits the data 
better (since the residual standard error is lower).

> *The interested reader can take a look at Vickers AJ and Altman DG. Analysing controlled trials with baseline and follow up measurements. British Medical Journal 2001;323:1123 http://www.bmj.com/content/323/7321/1123.full.pdf+html 

## Day 11 logistic regression
### Exercises with R
#### Introduction
> If there is an exposure (explanatory variable) in the data file that is a categorical variable where the categories are denoted by numbers, other than a 0-1, then you should tell this to R by using the function factor(). So factor(group) tells R that group is not a numeric variable but that it numbers should be used as group labels.
To fit the logistic regression model use:
`model <- lm(data~y+x,family=binomial)`. 
> model will store the result. To see what is in it use names(model) and if you want to see something specific use for instance model$coefficients. To get the tables from the test  summary(model). The deviance residuals can be obtained by residuals(model) and the fitted values by fitted.values(model).
Now you can leave out 1 variable from the model and look at the differences in AIC's. One can do this by fitting all the different models and then comparing them. One can also use the function drop1(fit). This function looks at the terms in fit, then leaves the terms out one by one and calculates for every term left out the AIC of the model. The command drop1(fit,test="Chisq") calculates the likelihood ratio test for every term left out. Then you can fit a model by leaving out the variable with the least influence and then start the procedure all over again using this last model as a starting point. etc. For the AIC this can also be done at once: step(model).

#### A note on model-building in R
> R does not have the same type of forward, backward and stepwise selection procedures as SPSS. The add1 and drop1 can be used to examine variables and decide which variable should be added/dropped next. The actual adding and dropping needs to be done by the analyst, so a new model is fitted and again checked for variables that can be added/dropped. Note that the add1 and drop1 functions both give the AIC by default; an F-test (for linear models) or a chi-square test of the deviance can be obtained by using an extra option in the command.
The step function was developed for automatic selection of variables using the drop1 and add1 functions repeatedly. However, please note that the AIC is used for model selection and not the deviance criterion (LR selection in SPSS). Using the scope command a “lower” and “upper” model are given and R uses a stepwise procedure (based on the AIC) to find the best model. In order to ‘trick’ R into using the chi-square test for deviance, we can give a criterion of 3.84 (instead of the default 2); note that this is equivalent to a chi-square test with 1 degree of freedom. The step function will then choose models that improve a model “significantly” based on the deviance test. Note that the critical value 3.84 is only appropriate for continuous and dichotomous explanatory variables, but not for explanatory variables with more than 2 categories (because then the degrees of freedom of the chi-square test is greater than 1 and the critical value of the chi-square test will be higher).

#### 6.
> This is a repeat of exercise #1, but now in R. The aim of this exercise is to redo the steps from this morning’s lecture, but for a different data, namely `perulung.RData`, containing lung function data of 636 Peruvian children. We wish to know whether the probability of having respiratory problems is different for boys and girls. 
a. `load(“perulung.RData”)`
```{r}
load(epistats::fromParentDir("data/perulung.RData"))
```

> The data frame d (containing data on FEV, age, height, gender and lung problems of Peruvian children) is now added to your workspace. Type names(d) to see the contents of the data frame. 

Even better: `str()` will give you the names and classes (= data type) and first few observations
```{r}
str(d)
```

> b.	Make a contingency table for problems vs. gender.

```{r}
table(d$problems, d$gender)
```

> c.	Use prop.table on the result to get the proportion of babies with low birth weight for both smoke categories.

```{r}
lb <- read.table(epistats::fromParentDir("data/lowbirth.dat"), header = T)
table(lb$low, lb$smoke)
prop.table(table(lb$low, lb$smoke))
```

> d.	To calculate odds ratios and relative risks you can use package epicalc. Install the package and load it (using library). (Should this not work here, you could try this at home or calculate OR and RR directly by typing in the appropriate numbers.)
e.	Using chisq.test and prop.test on the contingency table you can test whether OR=1.

```{r}
chisq.test(table(lb$low, lb$smoke))
epiR::epi.2by2(table(lb$low, lb$smoke), method = "case.control")
```

> f.	Do a logistic regression in R using the commands

```{r}
low.glm <- glm(problems ~ fev, family = binomial, data = d)
summary(low.glm)
coefficients(low.glm)
```

What is the logistic regression equation?

$$Ln(\frac{p}{1-p}) = \beta_0 + \beta_1X_1 = 1.4595 - 1.7247*fev$$

> `exp(coefficients(low.glm))` will give you odds and OR. What is the odds ratio for FEV?

```{r}
exp(coefficients(low.glm))
```

> g. Calculate yourself the estimated probability of suffering from lung problems for children with an fev value of 1.5.

```{r}
log_odds = 1.4595 - 1.7247 * 1.5
odds = exp(log_odds)
prob = 1/(1+odds)
prob
```


> h.	Now have R do this for you using the predict command (See: How to in R and help(predict.glm)). Note: in logistic regression by default you will get predictions on the log odds scale. If you want predictions for probabilities use type=”response” in the predict command.

Compare with using the `predict` function

```{r}
predict(low.glm, newdata = data.frame(fev = 1.5), type = "response")
```

This is 1-prob. It is not clear to me why it would give me this. 

> i.	Run a multiple logistic regression with explanatory variables fev, age, height and gender. Assign the outcome to the object full.logreg

```{r}
full.logreg <- glm(problems~fev+age+height+gender, data = d, family = binomial)
```

> j.	Give an estimate of the log odds of having lung problems for a child with fev=1.5, age=7, height=130 and gender=0. What if this child has age 8, 9 or 10?

```{r}
predict(full.logreg, 
        newdata = data.frame(
          fev = rep(1.5,4), 
          age = 7:10, 
          height = rep(130, 4), 
          gender = rep(0,4)),
        type = "response")
```

> k.	Repeat part l), but now estimate the probability of having lung problems.

This is probably 1 minus the answer of j.

> 7.	This is a repeat of exercise #2, but now in R. The ICU dataset consists of a sample of 200 subjects who were part of a much larger study on survival of patients following admission to an adult intensive care unit (ICU).  The major goal of this study was to develop a logistic regression model to predict the probability of survival to hospital discharge of these patients and to study the risk factors associated with ICU mortality. A number of publications have appeared which have focused on various facets of the problem. The reader wishing to learn more about the clinical aspects of this study should start with Lemeshow, Teres, Avrunin, and Pastides (1988). 
Data were collected at Baystate Medical Center in Springfield, Massachusetts. The data are given in the dataset icu.RData.
To keep the exercise simple only the most important risk factors will be considered.
Identification Code	id number	Id
Vital Status	0 = lived; 1 = died	sta
Age	Years	age
Service at ICU Admission	0 = medical; 1 = surgical	ser
Cancer Part of Present Problem	0 = no; 1 = yes	can
History of Chronic Renal Failure	0 = no; 1 = yes	crn
Infection Probable at ICU Admission	0 = no; 1 = yes	inf
CPR Prior to ICU Admission	0 = no; 1 = yes	cpr
Systolic Blood Pressure at ICU Admission	mm Hg	sys
Type of Admission	0 = elective; 1 = emergency	typ
PO2 from Initial Blood Gases	0 = > 60; 1 = < 60	po2
Bicarbonate from Initial Blood Gases	0 = > 18; 1 = < 18	bic
Creatinine from Initial Blood Gases	0 = < 2.0; 1 = > 2.0	cre
Level of Consciousness at ICU Admission	0 = no coma or stupor; 1 = deep stupor; 2 = coma	loc

> a.	Fit two models, one with risk factor age and one with risk factor sys.

```{r}
load(epistats::fromParentDir("data/icu.RData"))
str(ICU)

fit1 <- glm(STA~AGE, data = ICU, family = binomial)
fit2 <- glm(STA~SYS, data = ICU, family = binomial)
```

> b.	Give for both models an interpretation of the estimate of the OR with its 95% CI for the specific risk factor at hand.

```{r}
summary(fit1)
summary(fit2)

exp(MASS:::confint.glm(fit1))
exp(MASS:::confint.glm(fit2))

```

The 95% confidence interval for the variable `AGE` is strictly greater than 1, 
so with increasing age, the risk of death is greater. For `SYS` it is the other 
way around.

> c.	Which model fits the data best, and why?

The model with `SYS` is best, as it has the lowest AIC.

> d.	Fit a model with each of the other risk factors separately. Notice that all these risk factors are categorical so you need the ‘factor’ option. 

Lets do this in an R-programming way that repeats the same steps for each of 
the values in `independent_vars`

```{r}
independent_vars <- c("AGE", "SER", "CAN", "CRN", "INF", "CPR", 
                      "SYS", "TYP", "PO2", "BIC", "CRE", "LOC")
fits <- lapply(independent_vars, function(var) {
  form <- reformulate(termlabels = var, response = "STA")
  fit  <- glm(form, family = binomial, data = ICU)
  return(fit)
  })
```

Now `fits` is a list with 12 glm-fits, each item of this list contains 1 fit. 
The fit can be extracted with `fits[[i]]`, where i is between 1 and 12.

> e.	Which risk factors can explain the vital status? Which one is the most important one?

Lets get the AIC for each of the fits

```{r}
aics <- sapply(fits, function(fit) {
  result <- summary(fit)
  result$aic
})
data.frame(independent_vars, aics)
```

Now we see that `CPR` has the lowest AIC

> f.	So far every risk factor has been examined separately. Can vital status also explained by 2 or more risk factors? Build a model using the ‘step’ procedure. Discuss the final model.

```{r}
fit0    <- glm(STA~1, family = binomial, data = ICU)
fit_all <- glm(reformulate(termlabels = independent_vars, response = "STA"),
               family = binomial, data = ICU)
fit_step_aic <- step(fit0, scope = list(upper = fit_all), data = ICU, 
                 direction = "both")
fit_step_chi <- step(fit0, scope = list(upper = fit_all), data = ICU, 
                 direction = "both", test = "Chisq")
```

```{r}
summary(fit_step_aic)
```

Both the AIC and the chi-squared method converged to the same model.

The final model includes 4 variables:

* Level of consciousness
* Type of admission
* Age
* Cancer part of the problem
* Systolic blood pressure

The directions of all these variables make sense: risk is increased for higher levels,
except for systolic blood pressure, as having a lower blood pressure (in an ICU setting) 
is worse.

Even though variables could be added and dropped, they only got added.

> g.	In the final model there is one risk factor that was not significant in the separate analyses. Can you explain this?

First check which one this was. In each fit object, when calling `summary`,
the p-value for the coefficients is in the 4th column of row 2 of the coefficients table

```{r}
summary(fits[[1]])$coefficients
pvals <- sapply(fits, function(fit) summary(fit)$coefficients[2,4])
data.frame(independent_vars, pvals)
```

Here, `CAN` has a p-value of 1. It must have collinearity with at least 1 
of the other included variables, which can act as confounders. 
We can do a quick scan for colleniarity with a pairwise distribution plot

```{r, warning = F}
included_terms <- attr(fit_step_aic$terms, "term.labels")
included_terms

require(GGally)
GGally::ggpairs(ICU[, included_terms])
```

From a quick look at this graph, it looks like `CAN` and `TYP` are related.

> Now we will try to fit the best logistic curve by eye to the ICU data with categorized age (agecat) as only explanatory variable.

> h.	Make a table of sta against agecat (with sta as column variable), and put the result in an object called tab.

```{r}
tab <- table(factor(ICU$STA), factor(ICU$AGECAT))
tab
```

> i.	Have R calculate the proportion of people dying in the ICU for each age category, using the object tab.

```{r}
prop_tab <- prop.table(tab, margin = 2)
```

> j.	Plot these proportions against the age category.

```{r}
y = prop_tab[2,]
x = colnames(prop_tab)
plot(x, y, ylim = c(0,1))
```

> k.	Try to fit the best logistic curve to these points (see previous exercise, part e, with different values for b0 and b1).

The reference to 'previous excercise, part e' is unclear, but we will try to solve it.
A logistic line follows the equation:

$$p(x) = \frac{1}{1+e^{-(\beta_0 + \beta_1*x)}}$$

The probability of death for age category 20 is low, so lets choose a low intercept,
 say $\beta_0 = -4$. At the top of our distribution, the probability approaches 50%.
 Lets set $\beta_1$ such that at age 100, $\beta_0 + \beta_1 * 100 = 0$. So $\beta_1 = 0.04$

```{r}
x_levels <- 20:90
y_levels <- 1/(1+exp(-1*(-4 + 0.04*x_levels)))
plot(x,y)
lines(x_levels, y_levels)
```

Not too bad

> l.	Do a logistic regression on these data to get the best parameter estimates.

There is some ambituity here, we could use AGECAT as a (sort-of) continous variable,
with very discrete values. But since the researchers wanted to discretize age 
into categories, we might wan't to treat `AGECAT` as a factor variable. 
Based on the provided answers, treat AGECAT as 'continous', at least such that
 90-80 = 50-40.

```{r}
fit_age <- glm(STA~AGECAT, family = binomial, data = ICU)
coefs <- (coef(fit_age))
exp(coefs)
```

m.	Plot the points again, and add the logistic curve with the parameters from e).

```{r}
x_vals <- 20:90
y_vals <- 1/(1+exp(-1*(coefs[1] + coefs[2] * x_vals)))
plot(x,y)
lines(x_vals, y_vals)
```

#### 8.
> This is a repeat of exercise #3, but now in R. In the years 1995 through 1998 a research was done among 1243 Dalmatian puppies. It was determined whether or not they were deaf in at least one ear. The research question was if deafness was related to pigmentation. In order to answer this it was measured whether or not there were many spots on the skin, whether or not they had a spot on the head and whether or not the dog had blue eyes. In addition one wants to determine if there was heredity involved. In order to look at this the family history score was determined. This is a method to cope with litter  effects (heredity): for every puppy it was determined how many brothers or sisters were deaf. Call this number m. Then from the whole dataset the fraction of dogs that are deaf can be determined. This fraction is multiplied by litter size - 1. This is then the expected number of deaf brothers or sisters when there are no differences between litters. The family history score is now defined as fhs = m - fraction * (litter size - 1)
Whether or not the puppy is deaf (0=no, 1=yes)	deaf
The number of spots on the skin (1=light, 2=moderate, 3=heavy)	spot
Whether or not the dog had blue eyes (0=no, 1=yes)	blueeye
Whether or not the dog had an spot on the head (0=no, 1=yes)	headspot
Gender (0=male, 1=female)	gender
Family history score 	fhs

> a.	Read in the data file dalmatian.csv.

```{r, read_dal}
dal <- read.csv(epistats::fromParentDir("data/dalmatian.csv"))
dal <- read.table(epistats::fromParentDir("data/dalmatian.csv"), header = T,
                  sep = ",")
str(dal)
```

> b.	Explain how the variable fhs deals with the heredity.

```{r}
hist(dal$fhs)
```

`fhs` is the difference between the observed number of deaf puppies in a litter, 
and the expected number of deaf puppies in the litter, given that all litters 
are equal, so there is no effect of heredity. A puppy with `fhs` smaller than 0 was born 
in a litter with fewer deaf siblings than expected, so in a 'protected' litter.

> c.	Fit the logistic regression model for deaf with fhs as an exposure variable.

```{r}
fit <- glm(deaf~fhs, data = dal, family = binomial)
```

d.	Compare the AIC of the previous model with the model that only contains a constant. Also calculate the likelihood ratio and interpret the results.

```{r}
fit0 <- glm(deaf~1, data = dal, family = binomial)
summary(fit)
summary(fit0)
```

The model with `fhs` has a lower AIC, so fits the data better.

```{r}
lmtest::lrtest(fit0)
lmtest::lrtest(fit)
lmtest::lrtest(fit0, fit)
```

> e.	Which model fits the data best, and why?

The model with the explanatory variable, since this one has a higher log-likelihood

#### 9.
> This is a repeat of exercise #4, but now in R. Low birth weight is an outcome that has been of concern to physicians for years. This is due to the fact that infant mortality rates and birth defect rates are very high for low birth weight babies. A woman's behavior during pregnancy (including diet, smoking habits, and receiving prenatal care) can greatly alter the chances of carrying the baby to term and, consequently, of delivering a baby of normal birth weight.
The variables, listed below, have been shown to be associated with low birth weight in the obstetrical literature. The goal of the current study was to ascertain if these variables were important in the population being served by the medical center where the data were collected. So one wants to identify risk factors associated with giving birth to a low birth weight baby (weighing less than 2500 grams). 
Data were collected on 189 women, 59 of which had low birth weight babies and 130 of which had normal birth weight babies.
The four variables, that are thought to be of importance, are age, weight of the subject at her last menstrual period, race, and the number of physician visits during the first trimester of pregnancy.
Source: Hosmer and Lemeshow, Applied Logistic Regression, Wiley, (1989).
Identification Code	id
Birth Weight in Grams 	bwt
Low Birth Weight (0 = bwt  2500g; 1 = bwt < 2500g), 	lowb
Age of the Mother in Years 	age
Weight in Pounds at the Last Menstrual Period 	lwt
Race (1 = White, 2 = Black, 3 = Other) 	race
Smoking Status During Pregnancy (1 = Yes, 0 = No) 	smoke
History of Premature Labor (0 = None 1 = One, etc.) 	ptl
History of Hypertension (1 = Yes, 0 = No) 	ht
Presence of Uterine Irritability (1 = Yes, 0 = No) 	ui
Number of Physician Visits During the First Trimester (0,1, 2, …) 	ftv
a.	Read in the dataset lowbirth.dat

```{r}
lb <- read.table(epistats::fromParentDir("data/lowbirth.dat"), header = T)
str(lb)
```

> b.	Fit three models, one with exposure age, one with exposure smoke and one with the exposure ht.

```{r}
fit_0     <- glm(low~1  ,   data = lb, family = binomial)
fit_age   <- glm(low~age,   data = lb, family = binomial)
fit_smoke <- glm(low~smoke, data = lb, family = binomial)
fit_ht    <- glm(low~ht,    data = lb, family = binomial)
```

> c.	Give for all 3 models an interpretation of the estimate of  for the specific exposure at hand.

```{r}
exp(coef(fit_age))
exp(coef(fit_smoke))
exp(coef(fit_ht))
```

Higher age reduces the risk of low birth weight, smoking increases the risk, 
hypertension also increases the risk. The exponents of the coefficients are 
the odds-ratios comparing those with the exposure against those without.

> d.	Compare all 3 models to the model with only an intercept in it using AIC. Calculate for each the likelihood ratio and give an interpretation.

Compare AICs
```{r}
summary(fit_0)$aic
summary(fit_age)$aic
summary(fit_smoke)$aic
summary(fit_ht)$aic
```

Each model is a little better than the null model (i.e. has lower AIC)

LR tests:

```{r}
lmtest::lrtest(fit_0, fit_age)
lmtest::lrtest(fit_0, fit_smoke)
lmtest::lrtest(fit_0, fit_ht)
```

Ony the smoke and ht models are significantly better than the null-model, 
according to the LR-test

> e.	Compare the 3 models with each other by comparing AIC. Also calculate likelihood ratios.

```{r}
lmtest::lrtest(fit_age, fit_smoke)
lmtest::lrtest(fit_age, fit_ht)
lmtest::lrtest(fit_smoke, fit_ht)
```

> f.	Which model fits the data best, and why?

The model with smoke has the lowest Log-likelihood, which is significantly 
different from th eother models

> g.	Fit the logistic regression model with exposures age, lwt, race, smoke, ptl, ht, ui and ftv.

```{r}
fit_full <- glm(low~age+lwt+race+smoke+ptl+ht+ui+ftv, data = lb, family = binomial)
summary(fit_full)
```

> h.	Find out which exposures can be left out using AIC.

```{r}
drop1(fit_full)
```

Dropping age and ftv would lower AIC, so they can be left out.

> i.	Discuss the final model.

```{r}
fit_final <- glm(low~lwt+race+smoke+ptl+ht+ui, data = lb, family = binomial)
summary(fit_final)
drop1(fit_final)
```

Is this the same as using the step function?

```{r}
fit_step <- step(fit_0, scope = list(upper = fit_full), data = lb, direction = "both")
```

```{r}
sort(attr(fit_final$terms, "term.labels"))
sort(attr(fit_step$terms, "term.labels"))
```

Yes it is the same as using stepwise selection, which considers some extra models.

> j.	Write a short report about your findings. Include the statistical model you used, the method you used to reduce this model and the final results (estimate and standard errors).

Left for now

#### 10. Optional 	This is a repeat of exercise #5, but now in R. The aim of this exercise is to familiarize you with the logistic functions.

> a.	Make a sequence of values for x by giving the commands 

```{r}
x <- seq(-2.5,2.5,by=0.1)
x 	# The content of x is now printed to your screen.
```

> b.	Define y1 as 

```{r}
y1 <- exp(1+3*x)/(1+exp(1+3*x))
```

> c.	Then run the command 

```{r}
plot(x,y1,type="l") 
```

> (The subcommand type “l” connects the points with a line.) This results in the graph of the logistic function with coefficients b0 = 1 and b1 = 3.

> d.	Define y2 using b0 = 1 and b1 = 2, and y3 using b0 = 1 and b1 = -2

```{r}
y2 <- exp(1+2*x)/(1+exp(1+2*x))
y3 <- exp(1-2*x)/(1+exp(1-2*x))
```

e.	Add these to the existing plot using lines(x, y2, lty=2) and similar for y3.

```{r}
plot(x,y1,type="l") 
lines(x,y2,lty = 2) 
lines(x,y3,lty = 2) 
```

> f.	Draw a horizontal line at y=0.5 using the command abline(h=0.5)

```{r}
plot(x,y1,type="l") 
lines(x,y2,lty = 2) 
lines(x,y3,lty = 2)
abline(h = 0.5)
```

> Do you understand how this abline command works?.

It takes a slope and intercept (or vertical / horizontal value) and draws a line. 
`h = 0.5` means a horizontal line at y = 0.5

> g.	Read off from the graphs at which x-values the horizontal line from f) intersects each of the curves. How does the answer connect to the values of b0 and b1? What would be the interpretation of those x-values in a logistic regression?

From the formula one can read that $p = 0.5$ when $\beta_0 + \beta_1x = 0$. 
So for y1, x = -1/3, for y2, x = -1/2, for y3, x = 1/2. This matches the plot













## Session information

<!-- Insert the session information into the document -->
```{r session-info}
```
