---
title: "Assignments for applied Bayesian statistics, part 2"
author: "Wouter van Amsterdam"
date: 2018-05-02
output: html_document
---

<!-- The file analysis/chunks.R contains chunks that define default settings
shared across the workflowr files. -->
```{r read-chunk, include=FALSE, cache=FALSE}
knitr::read_chunk("chunks.R")
```

<!-- Update knitr chunk options -->
```{r knitr-opts-chunk, include=FALSE}
```

<!-- Insert the date the file was last updated -->
```{r last-updated, echo=FALSE, results='asis'}
```

<!-- Insert the code version (Git commit SHA1) if Git repository exists and R
 package git2r is installed -->
```{r code-version, echo=FALSE, results='asis'}
```

<!-- Add your analysis here -->


```{r, 'load_packages'}
library(dplyr)
library(data.table)
library(magrittr)
library(purrr)
library(here) # for tracking working directory
library(ggplot2)
library(epistats)
library(broom)
library(rjags)
library(Bain)
```

# Day 3

Run tutorial from bain

```{r, 'step1'}
# THIS IS BFTUTORIAL.R

# FURTHER EXPLANATION OF THE CODE USED TO RUN BAIN CAN BE FOUND IN THE APPENDIX
# OF A TUTORIAL ON TESTING HYPOTHESES USING THE BAYES FACTOR 

#***************
#TUTORIAL STEP 1
#***************
# Load libraries. These libraries contain functions such as lm and Bain that will be used in 
# this R code. Each time you reopen BFTutorial.R you have to execute this step.

library(MASS)
library(Bain)
library(psych)

#***************
#TUTORIAL STEP 2
#***************
# Read Data. If you open monin.txt and holbar.txt in a text editor, you can see
# that variable labels have been inserted between " " in the first line of the file
# (this is called a header).

monin<-read.table(here("data", "monin.txt"),header=TRUE)
monin$group<-factor(monin$group)       # this command tells R that group is a factor             
holubar<-read.table(here("data", "holubar.txt"),header=TRUE)
holubar$gr <- factor(holubar$gr)       # this command tells R that gr is a factor

#***************
#TUTORIAL STEP 3
#***************
# Compute descriptives for the Monin data. The dependent variable name is attract, 
# the factor name is group (see also TutorialStep 2).

descrip <- describeBy(monin$attract,monin$group,mat=TRUE)
print(descrip)
```

```{r}
#***************
#TUTORIAL STEP 4
#***************
# Compute the Bayes factor for the Monin data. 
# If you return to BFTutorial.R after closing it, you always have to rerun 
# Steps 1 and 2, and lines 44-60.

prepmonin <-  lm(attract~group-1, data=monin)
varm <- (summary(prepmonin)$sigma)**2
estm<-coef(prepmonin)

sampm<-table(monin$group)

cov1m <- varm/sampm[1]
cov2m <- varm/sampm[2]
cov3m <- varm/sampm[3]
cov1m <- matrix(cov1m,1,1)
cov2m <- matrix(cov2m,1,1)
cov3m <- matrix(cov3m,1,1)

covm<-list(cov1m,cov2m,cov3m)

ERr1<-matrix(c(1,-1,0,0,0,1,-1,0),nrow=2,ncol=4, byrow = TRUE)
IRr1<-matrix(0,0,0)

resmonin<-Bain(estimate=estm,Sigma=covm,grouppara=1,jointpara=0,n=sampm,ERr1,IRr1)
```

Add hypothesis of m1 > m2 > m3

```{r}
# ERr2<-matrix(c(1,-1,0,0,
#                0,1,-1,0),nrow=2,ncol=4, byrow = TRUE)
ERr2<-matrix(0,0,0)
IRr2<-matrix(c(1,-1,0,0,
               0,1,-1,0),nrow=2,ncol=4, byrow = TRUE)

resmonin2<-Bain(estimate=estm,Sigma=covm,grouppara=1,jointpara=0,n=sampm,ERr2,IRr2)

```

Back to hypothesis m1 = m2 = m3

Add 2 outliers such that H1 is supported

Check means

```{r}
means <- monin %>%
  group_by(group) %>%
  summarize_at(vars(attract), funs(mean, n()))
means

means <- means$mean

```

Add outliers such that mean 1 and mean 2 are also 0.02

```{r}
monin_outliers <- monin %>%
  rbind(c(19 * (means[3] - means[1]), 1)) %>%
  rbind(c(19 * (means[3] - means[2]), 2))

monin_outliers %>%
  group_by(group) %>% summarize(mean(attract))
```

Prepare priors again with new data

```{r}

prepmonin <-  lm(attract~group-1, data=monin_outliers)
varm <- (summary(prepmonin)$sigma)**2
estm<-coef(prepmonin)

sampm<-table(monin$group)

cov1m <- varm/sampm[1]
cov2m <- varm/sampm[2]
cov3m <- varm/sampm[3]
cov1m <- matrix(cov1m,1,1)
cov2m <- matrix(cov2m,1,1)
cov3m <- matrix(cov3m,1,1)

covm<-list(cov1m,cov2m,cov3m)

ERr3 <- matrix(c(-1, 1, 0, 0,
                 0, -1, 1, 0),
               ncol = 4, byrow = T)
IRr3 <- matrix(0,0,0)

resmonin<-Bain(estimate=estm,Sigma=covm,grouppara=1,jointpara=0,n=sampm,ERr3,IRr3)


```

Formulate hyoptheses:

- H1, mu1 = mu2, mu1 > mu3
- H2, mu2 > mu1 > mu3
- H0, mu1 = mu2 = mu3
- Hu, mu1, mu2, mu3

Evaluate on other data

```{r}
holubar %>%
  group_by(gr) %>%
  summarize_at(vars(at), funs(mean, n()))
```


```{r}
hol_lm <- lm(at ~ gr, data = holubar)
hol_var <- summary(hol_lm)$sigma^2
hol_estm <- coef(hol_lm)
hol_counts <- table(holubar$gr)


hol_vcov1 <- matrix(hol_var / hol_counts[1])
hol_vcov2 <- matrix(hol_var / hol_counts[2])
hol_vcov3 <- matrix(hol_var / hol_counts[3])

hol_vcovs <- list(hol_vcov1, hol_vcov2, hol_vcov3)

ERr1 <- matrix(c(1, -1, 0, 0), ncol = 4)
IRr1 <- matrix(c(1, 0, -1, 0), ncol = 4)
ERr2 <- matrix(0, 0, 0)
IRr2 <- matrix(c(-1, 1, 0, 0,
                 1, 0, -1, 0), ncol = 4, byrow = T)
ERr3 <- matrix(c(1, -1, 0, 0,
                 0, 1, -1, 0), ncol = 4, byrow = T)
IRr3 <- matrix(0, 0, 0)

res_hol <- Bain(estimate = hol_estm, Sigma = hol_vcovs, 
                grouppara = 1, jointpara = 0, n = hol_counts,
                ERr1, IRr1, ERr2, IRr2, ERr3, IRr3)
```



```{r}
#***************
#TUTORIAL STEP 5
#***************
# Compute the Bayes factor for "all" the hypotheses for the Monin data
# Note that estm, covm and sampm have already been computed in Step 4.

ERr1<-matrix(c(1,-1,0,0,
               0,1,-1,0),nrow=2,ncol=4, byrow = TRUE)
IRr1<-matrix(0,0,0)

ERr2<-matrix(c(1,-1,0,0),nrow=1,ncol=4, byrow = TRUE)
IRr2<-matrix(0,0,0)

ERr3<-matrix(c(1,0,-1,0),nrow=1,ncol=4, byrow = TRUE)
IRr3<-matrix(0,0,0)

ERr4<-matrix(c(0,1,-1,0),nrow=1,ncol=4, byrow = TRUE)
IRr4<-matrix(0,0,0)

resmonin<-Bain(estimate=estm,Sigma=covm,grouppara=1,jointpara=0,n=sampm,ERr1,IRr1,ERr2,IRr2,ERr3,IRr3,ERr4,IRr4)

#***************
#TUTORIAL STEP 6
#***************
# Bayesian updating. Stepwise increasing the sample size for the Monin data

# WARNING: INCREASE THE SIZE OF THE PLOT WINDOW IN THE LOWER RIGHT HAND SCREEN
# OTHERWISE YOU ARE BOUND TO RECEIVE AN ERROR MESSAGE

# create an object BFmonin in which the Bayes factor for different sample sizes 
# will be collected.
BFmonin<-1

# start with an initial sample size of 2+0=2 persons per group an incease by 1 
# until 2+17=19 persons per group

for (i in 0:17){

  # collect the subset of persons selected in a matrix  
  subdata <- matrix(0, nrow = 6+3*i, ncol = 2)  # in total 6 initial persons in 3 groups
                                                # for which 2 variables have been recorded
  subdata <- monin[c(1:(2+i),20:(21+i),39:(40+i)),1:2] # the first person in the first group is 1
                                                       # the first person in the second group is 20
                                                       # the first person in the third group is 39
  
  submonin <-  lm(attract~group-1, data=subdata)
  subvarm <- (summary(submonin)$sigma)**2
  subestm<-coef(submonin)
  
  subsampm<-c(2+i,2+i,2+i)
  subcov1m <- subvarm/subsampm[1]
  subcov2m <- subvarm/subsampm[2]
  subcov3m <- subvarm/subsampm[3]
  subcov1m <- matrix(subcov1m,1,1)
  subcov2m <- matrix(subcov2m,1,1)
  subcov3m <- matrix(subcov3m,1,1)
  
  subcovm<-list(subcov1m,subcov2m,subcov3m)
  
  ERr1<-matrix(c(1,-1,0,0,0,1,-1,0),nrow=2,ncol=4, byrow = TRUE)
  IRr1<-matrix(0,0,0)
  
  resmonin<-Bain(estimate=subestm,Sigma=subcovm,grouppara=1,jointpara=0,n=subsampm,ERr1,IRr1,print = FALSE)
  
  # collect BF0a from the object resmonin and store in BFmonin
  BFmonin[1+i]<-resmonin[[4]][[3]]
}

# create a vector containing the sample size per group and plot this versus BF0a 
NperGroup<-c(2,3,4,5,6,7,8,9,10,11,12,13,14,15,16,17,18,19)
plot(NperGroup,BFmonin,ylab="BF0a",xlab="N per Group")
lines(NperGroup,BFmonin)
```

Bayes factor keeps changing after addition of more participants

```{r}
#***************
#TUTORIAL STEP 7
#***************
# Bayesian updating. Posterior Model Probabilities for the Holubar Data

# Create objects containing all 5 PMPs and each separate PMP
pall<-1
p1<-1
p2<-1
p3<-1
p4<-1
p5<-1

for (i in 0:17){
  
  subdata <- matrix(0, nrow = 6+3*i, ncol = 2)
  subdata <- monin[c(1:(2+i),20:(21+i),39:(40+i)),1:2]
  
  submonin <-  lm(attract~group-1, data=subdata)
  subvar  <- (summary(submonin)$sigma)**2
  subest <-coef(submonin)
  
  subsamp<-c(2+i,2+i,2+i)
  
  subcov1 <- subvar/subsamp[1]
  subcov2 <- subvar/subsamp[2]
  subcov3 <- subvar/subsamp[3]
  subcov1 <- matrix(subcov1,1,1)
  subcov2 <- matrix(subcov2,1,1)
  subcov3 <- matrix(subcov3,1,1)
  
  subcov<-list(subcov1,subcov2,subcov3)
  
  ERr1<-matrix(c(1,-1,0,0,
                 0,1,-1,0),nrow=2,ncol=4, byrow = TRUE)
  IRr1<-matrix(0,0,0)
  
  ERr2<-matrix(c(1,-1,0,0),nrow=1,ncol=4, byrow = TRUE)
  IRr2<-matrix(0,0,0)
  
  ERr3<-matrix(c(1,0,-1,0),nrow=1,ncol=4, byrow = TRUE)
  IRr3<-matrix(0,0,0)
  
  ERr4<-matrix(c(0,1,-1,0),nrow=1,ncol=4, byrow = TRUE)
  IRr4<-matrix(0,0,0)
  
  resmonin<-Bain(estimate=subest,Sigma=subcov,grouppara=1,jointpara=0,n=subsamp,ERr1,IRr1,ERr2,IRr2,ERr3,IRr3,ERr4,IRr4)

  pall <- resmonin[[5]][[9]]
  pall <- as.numeric(levels(pall))[pall]
  
  p1[i+1]<-pall[1]
  p2[i+1]<-pall[2]
  p3[i+1]<-pall[3]
  p4[i+1]<-pall[4]
  p5[i+1]<-pall[5]
}
NperGroup<-c(2,3,4,5,6,7,8,9,10,11,12,13,14,15,16,17,18,19)

# create a legenda for the plot
leg <- c("H0","Ha1","Ha2","Ha3","Ha")

# give each line a different appearance
lty <- c(1,2,3,4,5)

plot(NperGroup,p1,type = "l",ylab = "posterior probabilities",xlab="N per Group",ylim=(0:1.0))
lines(NperGroup,p2,lty=2)
lines(NperGroup,p3,lty=3)
lines(NperGroup,p4,lty=4)
lines(NperGroup,p5,lty=5)
legend(x = 17, y = 0.95,legend = leg, lty =lty)
```

## Session information

<!-- Insert the session information into the document -->
```{r session-info}
```
