---
title: "Bayesian statistics assignments day 4, regression, mediation and moderation"
author: "Wouter van Amsterdam"
date: 2018-05-03
output: html_document
---

<!-- The file analysis/chunks.R contains chunks that define default settings
shared across the workflowr files. -->
```{r read-chunk, include=FALSE, cache=FALSE}
knitr::read_chunk("chunks.R")
```

<!-- Update knitr chunk options -->
```{r knitr-opts-chunk, include=FALSE}
```

<!-- Insert the date the file was last updated -->
```{r last-updated, echo=FALSE, results='asis'}
```

<!-- Insert the code version (Git commit SHA1) if Git repository exists and R
 package git2r is installed -->
```{r code-version, echo=FALSE, results='asis'}
```

<!-- Add your analysis here -->


```{r}
library(dplyr)
library(data.table)
library(magrittr)
library(purrr)
library(here) # for tracking working directory
library(ggplot2)
library(epistats)
library(broom)
library(rjags)
```

# Day 4

## Linear regression

```{r}
# This code runs a Bayesian linear regression analysis using the package rjags.
# To start, import the data set LinearRegressionCubic.csv.

MyData <- read.csv(file=here("data", "LinearRegressionCubic.csv"), header=TRUE, sep=",")
N<-dim(MyData)[1]
str(MyData)
```

Define priors and likelihood in bugs language

```{r}
# writing out the .txt file with the model

modelstring <- as.character("
model{
beta.0 ~ dnorm(0, .001); # prior for the intercept
beta.1 ~ dnorm(0, .001); # prior for b1
beta.2 ~ dnorm(0, .001); # prior for b2
beta.3 ~ dnorm(0, .001); # prior for b3
tau.e ~ dgamma(.5, .5); # prior for the error precision for Y

sigma2.e<-1/tau.e
sigma.e<-sqrt(sigma2.e)

# Conditional probability of the data
# A regression model

for(i in 1:N){
y.prime[i] <- beta.0 + beta.1*x[i]+ beta.2*x[i]*x[i]+ beta.2*x[i]*x[i]*x[i]; # predicted value of Y
y[i] ~ dnorm(y.prime[i], tau.e); # conditional distribution of y
}
}
") # closes the model as string

model.file.name <- "bayes_4_Linear Regression.txt"
write(x=modelstring, file=here("analysis", model.file.name), append=FALSE)
```

Compile the model and run simulations

```{r}
library('rjags')

jags <- jags.model(here("analysis", model.file.name),
                   data = list('x' = MyData$X,
                               'y' = MyData$Y,
                               'N' = N),
                   n.chains = 3)

out=coda.samples(jags, variable.names=c("beta.0","beta.1","beta.2", "beta.3", "tau.e"),
                       n.iter=100)

summary(out)
```

Check model conversion

```{r}
# Coda
library(coda)

model.as.mcmc.list <- as.mcmc.list(out)

gelman.diag(model.as.mcmc.list)
gelman.plot(model.as.mcmc.list)
plot(model.as.mcmc.list, trace=TRUE, density=FALSE)
```

Gelman numbers are above 1.1

Does not look like conversion, run additional samples

```{r}
# Running additional iterations
out2=coda.samples(jags, variable.names=c("beta.0","beta.1","beta.2", "beta.3", "tau.e"),
                 n.iter=2000)

model.as.mcmc.list <- as.mcmc.list(out2)

gelman.diag(model.as.mcmc.list)
gelman.plot(model.as.mcmc.list)
plot(model.as.mcmc.list, trace=TRUE, density=FALSE)

summary(out2)
```

Looks like conversions after about 2000 iterations

Let's burn in with 2000 and then draw samples


```{r}
jags <- jags.model(here("analysis", model.file.name),
                   data = list('x' = MyData$X,
                               'y' = MyData$Y,
                               'N' = N),
                   n.chains = 3)


update(jags, 2000)

out3=coda.samples(jags, variable.names=c("beta.0","beta.1","beta.2", "beta.3", "tau.e"),
                 n.iter=2000)

model.as.mcmc.list <- as.mcmc.list(out3)
plot(out3, density = F)
```



Conversion is still not perfect.

Let's center X

```{r}
jags_scaled <- jags.model(here("analysis", model.file.name),
                   data = list('x' = as.numeric(scale(MyData$X, center = T, scale = T)),
                               'y' = MyData$Y,
                               'N' = N),
                   n.chains = 3)


update(jags_scaled, 2000)

out4=coda.samples(jags_scaled, variable.names=c("beta.0","beta.1","beta.2", "beta.3", "tau.e"),
                 n.iter=2000)

model.as.mcmc.list_scaled <- as.mcmc.list(out4)
plot(out4, density = F)
```

Looks better



```{r}
# After the chains have converged
# combine draws from chains
draws.to.analyze.as.one.list <- 
  as.mcmc(do.call(rbind,model.as.mcmc.list_scaled))

#obtain mean, medians, and quantiles
summary.stats <- summary(draws.to.analyze.as.one.list)
summary.stats

#obtain highest-posterior density interval
HPD.interval <- HPDinterval(draws.to.analyze.as.one.list, 
                            prob=.95)
HPD.interval

```

No indication of cubic effect, only linear effect

## Mediation analysis

Load data and define in buggs language

```{r}
# This code runs a Bayesian mediation analysis using the package rjags.
# To start, import the data set waterconsumption.csv.

MyData <- read.csv(file=here("data", "waterconsumption.csv"), header=TRUE, sep=",")
N<-dim(MyData)[1]

modelstring <- as.character("
model {
############################################
  # Prior distributions
############################################
  beta.0.m ~ dnorm(1, .001); # prior for the intercept for M
  beta.0.y ~ dnorm(1, .001); # prior for the intercept for Y

  a ~ dnorm(5, .11); # prior for a
  b ~ dnorm(5, .11); # prior for b

  cp ~ dnorm(0, .11); # prior for c?
  tau.e.M ~ dgamma(.5, .5); # prior for the error precision for M
  tau.e.Y ~ dgamma(.5, .5); # prior for the error precision for Y
  
  ab <-a*b

############################################
  # Conditional probability of the data
  # A regression model
############################################
  for(i in 1:N){
    m.prime[i] <- beta.0.m + a*(x[i]-70.18); # predicted value of M, predictor is mean-centered
    y.prime[i] <- beta.0.y + cp*(x[i]-70.18) + b*(m[i]-3.06) ; # predicted value of Y, predictor is mean-centered
    m[i] ~ dnorm(m.prime[i], tau.e.M); # conditional distribution of m
    y[i] ~ dnorm(y.prime[i], tau.e.Y); # conditional distribution of y
  }  
}
") # closes the model as string

#############################################################################################################################
# Write out the BUGS code to a file
#############################################################################################################################
model.file.name <- here("analysis", "bayes_4_Single Mediator Model.txt")
write(x=modelstring, file=model.file.name, append=FALSE)
```

Compile model and sample

```{r}
library(rjags)

mediation.model <- jags.model(model.file.name,
                   data = list('x' = MyData$x,
                               'm' = MyData$m,
                               'y' = MyData$y,
                               'N' = N),
                 n.chains = 3)
out=coda.samples(mediation.model,
                 variable.names=c("a", "ab","b", "cp", "beta.0.m", 
                   "beta.0.y", "tau.e.M", "tau.e.Y"),
                 n.iter=10000)

summary(out)
```

Plot convergence and posterior distributions

```{r}
library(coda)

model.as.mcmc.list <- as.mcmc.list(out)
draws.to.analyze.as.one.list <- 
  as.mcmc(do.call(rbind,model.as.mcmc.list)) 

summary(draws.to.analyze.as.one.list)

# diagnostics of convergence
gelman.diag(model.as.mcmc.list)
gelman.plot(model.as.mcmc.list)
plot(model.as.mcmc.list, trace=TRUE, density=FALSE)
plot(model.as.mcmc.list, trace=FALSE, density=TRUE)
```

No signs of non-convergence

For parameters a, ab, and b, we have a clear sign that they are greater than 0
For cp, a parameter value of 0 false within the 95% credible interval

There is a clear indication of mediation, and not of a direct effect


```{r}
# running additional iterations
out2=coda.samples(mediation.model,
                 variable.names=c("a", "ab","b", "cp", "beta.0.m", 
                                  "beta.0.y", "tau.e.M", "tau.e.Y"),
                 n.iter=10000)

model.as.mcmc.list <- as.mcmc.list(out2)
draws.to.analyze.as.one.list <- 
  as.mcmc(do.call(rbind,model.as.mcmc.list)) 

#obtain mean, medians, and quantiles
summary.stats <- summary(draws.to.analyze.as.one.list)
summary.stats

#obtain highest-posterior density interval
HPD.interval <- HPDinterval(draws.to.analyze.as.one.list, 
                            prob=.95)
HPD.interval

```

### Repeat with different priors:

First center predictors in the data (X and M) without scaling to unit variance

```{r}
scale2 <- function(x, ...) as.numeric(scale(x, ...))
df <- MyData %>%
  mutate_at(vars(x, m), scale2, center = T, scale = F)
```

Define 3 priors

1. Uniform [0,3] for both a & b
2. Uniform [-3,3] for both a & b
3. Normal (0, prec=.01) for both a & b

```{r}
model_string_upos <- as.character("
model {
############################################
  # Prior distributions
############################################
  beta.0.m ~ dnorm(1, .001); # prior for the intercept for M
  beta.0.y ~ dnorm(1, .001); # prior for the intercept for Y

  a ~ dunif(0, 3); # prior for a
  b ~ dunif(0, 3); # prior for b

  cp ~ dnorm(0, .11); # prior for c?
  tau.e.M ~ dgamma(.5, .5); # prior for the error precision for M
  tau.e.Y ~ dgamma(.5, .5); # prior for the error precision for Y
  
  ab <-a*b

############################################
  # Conditional probability of the data
  # A regression model
############################################
  for(i in 1:N){
    m.prime[i] <- beta.0.m + a*(x[i]); # predicted value of M, predictor is mean-centered in R
    y.prime[i] <- beta.0.y + cp*(x[i]) + b*(m[i]) ; # predicted value of Y, predictor is mean-centered in R
    m[i] ~ dnorm(m.prime[i], tau.e.M); # conditional distribution of m
    y[i] ~ dnorm(y.prime[i], tau.e.Y); # conditional distribution of y
  }  
}
") # closes the model as string

#############################################################################################################################
# Write out the BUGS code to a file
#############################################################################################################################
model_file_upos <- here("analysis", "bayes_4_mediation_uniform_positive.txt")
write(x=model_string_upos, file=model_file_upos, append=FALSE)
mediation.model_upos <- jags.model(model_file_upos,
                   data = list('x' = df$x,
                               'm' = df$m,
                               'y' = df$y,
                               'N' = N),
                 n.chains = 3)



model_string_unif <- as.character("
model {
############################################
  # Prior distributions
############################################
  beta.0.m ~ dnorm(1, .001); # prior for the intercept for M
  beta.0.y ~ dnorm(1, .001); # prior for the intercept for Y

  a ~ dunif(-3, 3); # prior for a
  b ~ dunif(-3, 3); # prior for b

  cp ~ dnorm(0, .11); # prior for c?
  tau.e.M ~ dgamma(.5, .5); # prior for the error precision for M
  tau.e.Y ~ dgamma(.5, .5); # prior for the error precision for Y
  
  ab <-a*b

############################################
  # Conditional probability of the data
  # A regression model
############################################
  for(i in 1:N){
    m.prime[i] <- beta.0.m + a*(x[i]); # predicted value of M, predictor is mean-centered in R
    y.prime[i] <- beta.0.y + cp*(x[i]) + b*(m[i]) ; # predicted value of Y, predictor is mean-centered in R
    m[i] ~ dnorm(m.prime[i], tau.e.M); # conditional distribution of m
    y[i] ~ dnorm(y.prime[i], tau.e.Y); # conditional distribution of y
  }  
}
") # closes the model as string

#############################################################################################################################
# Write out the BUGS code to a file
#############################################################################################################################
model_file_unif <- here("analysis", "bayes_4_mediation_uniform.txt")
write(x=model_string_unif, file=model_file_unif, append=FALSE)
mediation.model_unif <- jags.model(model_file_unif,
                   data = list('x' = df$x,
                               'm' = df$m,
                               'y' = df$y,
                               'N' = N),
                 n.chains = 3)



model_string_norm <- as.character("
model {
############################################
  # Prior distributions
############################################
  beta.0.m ~ dnorm(1, .001); # prior for the intercept for M
  beta.0.y ~ dnorm(1, .001); # prior for the intercept for Y

  a ~ dnorm(0, .01); # prior for a
  b ~ dnorm(0, .01); # prior for b

  cp ~ dnorm(0, .11); # prior for c?
  tau.e.M ~ dgamma(.5, .5); # prior for the error precision for M
  tau.e.Y ~ dgamma(.5, .5); # prior for the error precision for Y
  
  ab <-a*b

############################################
  # Conditional probability of the data
  # A regression model
############################################
  for(i in 1:N){
    m.prime[i] <- beta.0.m + a*(x[i]); # predicted value of M, predictor is mean-centered in R
    y.prime[i] <- beta.0.y + cp*(x[i]) + b*(m[i]) ; # predicted value of Y, predictor is mean-centered in R
    m[i] ~ dnorm(m.prime[i], tau.e.M); # conditional distribution of m
    y[i] ~ dnorm(y.prime[i], tau.e.Y); # conditional distribution of y
  }  
}
") # closes the model as string

#############################################################################################################################
# Write out the BUGS code to a file
#############################################################################################################################
model_file_norm <- here("analysis", "bayes_4_mediation_normal.txt")
write(x=model_string_norm, file=model_file_norm, append=FALSE)
mediation.model_norm <- jags.model(model_file_norm,
                   data = list('x' = df$x,
                               'm' = df$m,
                               'y' = df$y,
                               'N' = N),
                 n.chains = 3)




```

Combine models in a data frame for easy manipulation

```{r}
models <- data.frame(model_prior = c("uniform_positive", "uniform", "normal"))
models %<>% 
  mutate(model = list(uniform_positive = mediation.model_upos, 
                                  uniform = mediation.model_unif,
                                  normal = mediation.model_norm))

```


Burnin 10000 and then sample 10000

```{r}
nburnin = 10000
nsample = 10000

params = c("a", "ab","b", "cp", "beta.0.m", 
                                  "beta.0.y", "tau.e.M", "tau.e.Y")

burnin_and_sample <- function(model, params, nburnin, nsamples) {
  update(model, nburnin)
  samples = coda.samples(model, n.iter = nsamples, variable.names = params)
  return(samples)
}

models %<>%
  mutate(samples = map(model,
                        ~burnin_and_sample(.x, params, nburnin = nburnin, nsamples = nsample)))


```

Check convergence for the samples

```{r}
plot(models$samples[[1]], density = F)
```

Looks good

```{r}
plot(models$samples[[2]], density = F)
```

Looks good

```{r}
plot(models$samples[[3]], density = F)
```

Looks good

Let's grab the HDR confidence intervals for the parameters

```{r}
hpds <- models$samples %>%
  # map(samples_to_mcmc) %>%
  map(HPDinterval) %>%
  map(as.data.frame) %>%
  map(~mutate(.x, parameter = params)) %>%
  rbindlist(idcol = "model")
hpds
```

Plot parameter credible intervals

```{r}
hpds %>%
  ggplot(aes(ymin = lower, ymax = upper, x = parameter, col = model)) + 
  geom_errorbar(position = "dodge", width = .25) + 
  coord_flip() + theme_minimal()
```

No real differences for the different priors

Specifically, none of the confidence bounds for a and b overlap 0, 
so uniform positive did not enforce this too much




## Moderation analysis

Load data

```{r}
# This code runs a Bayesian moderation analysis using the package rjags.
# To start, import the data set Moderationdata.csv.

MyData <- read.csv(file=here("data", "Moderationdata.csv"), header=TRUE, sep=",")
N<-dim(MyData)[1]
```

Define model and compile

```{r}
# writing out the .bug file with the model

modelstring <- as.character("
model{
beta.0 ~ dnorm(0, .001); # prior for the intercept
beta.1 ~ dnorm(0, .001); # prior for regression coefficient for x
beta.2 ~ dnorm(0, .001); # prior for regression coefficient for z
beta.3 ~ dnorm(0, .001); # prior for regression coefficient for intercation xz
tau.e ~ dgamma(.5, .5); # prior for the error precision for Y

sigma2.e<-1/tau.e
sigma.e<-sqrt(sigma2.e)

# Conditional probability of the data
# A regression model

for(i in 1:N){
y.prime[i] <- beta.0 + beta.1*x[i] + beta.2*z[i] + beta.3*x[i]*z[i] ; # predicted value of Y
y[i] ~ dnorm(y.prime[i], tau.e); # conditional distribution of y
}
}
") # closes the model as string

model.file.name <- here("analysis", "bayes_4_Moderation.txt")
write(x=modelstring, file=model.file.name, append=FALSE)

library('rjags')

jags <- jags.model(model.file.name,
                   data = list('x' = MyData$x,
                               'z' = MyData$z,
                               'y' = MyData$y,
                               'N' = N),
                   n.chains = 3)
```

Run some samples

```{r}
out=coda.samples(jags, variable.names=c("beta.0","beta.1","beta.2", "beta.3", "tau.e"),
                       n.iter=10000)

summary(out)
```

Analyze convergence

```{r}
# Coda
library(coda)

model.as.mcmc.list <- as.mcmc.list(out)

gelman.diag(model.as.mcmc.list)
gelman.plot(model.as.mcmc.list)
plot(model.as.mcmc.list, trace=TRUE, density=FALSE)
plot(model.as.mcmc.list, trace=FALSE, density=TRUE)
```

Convergence is not bad

```{r}
# Run additional iterations
out2=coda.samples(jags, variable.names=c("beta.0","beta.1","beta.2", "beta.3", "tau.e"),
                 n.iter=10000)
model.as.mcmc.list <- as.mcmc.list(out2)
# combine draws from chains
draws.to.analyze.as.one.list <- 
  as.mcmc(do.call(rbind,model.as.mcmc.list))

#obtain mean, medians, and quantiles
summary.stats <- summary(draws.to.analyze.as.one.list)
summary.stats

#obtain highest-posterior density interval
HPD.interval <- HPDinterval(draws.to.analyze.as.one.list, 
                            prob=.95)
HPD.interval
```



### Plot region of significance

For which value of the mediator is there a significant effect of the predictor

We can use out2

```{r}
length(out2)
dim(out2[[1]])
colnames(out2[[1]])

```

out2 is a list of length 3, 1 for each chain
- for each chain, there are 50000 values, which are 10000 samples for 5 parameters

this is a numeric matrix that can easily be converted to a data frame

We can combine these into a (very long) data frame

Let's us data.table, which is suited for large datasets

```{r}
out2 %>%
  map(as.data.frame) %>%
  rbindlist(idcol = "chain") -> samples
str(samples)
```

Pick a grid of values of the moderator, and 
calculate the 'simple slope' of the predictor for each value

```{r}
table(MyData$z)
```

Z values are binary here, so it doesn't make a lot of sense to create a 
region of significance here, but we can do it

```{r}
z_mean <- mean(MyData$z)
z_values <- seq(0, 1, length.out = 100)
```

Now for each z_value, we will calculate the value of the simple slope for each sample

Add a column for sample number for ease of tracking

```{r}
samples[, sample_id:=.I]
```


```{r}
simple_slopes <- samples[, list(z = z_values, 
                                simple_slope = map_dbl(z_values, ~beta.1 + .x * beta.3)),
                         by = "sample_id"]

slopes_and_samples <- merge(samples, simple_slopes, by = "sample_id",
                            all.x = T, all.y = T)
```

```{r}
dim(slopes_and_samples)
head(slopes_and_samples)
```

Now we have the simple slope for 100 values of z for all samples

We can compute the HPDinterval for each z value

```{r}
hpds <- slopes_and_samples[, list(
  simple_slope = median(simple_slope),
  hpd = list(HPDinterval(as.mcmc(.SD)))),
                           by = "z"]
```

Now for each value of z we have the HPD interval of all parameters

```{r}
hpds$hpd[[1]]
```

Grab the upper and lower value for the simple_slope from the hpd intervals and plot

```{r}
hpds %<>%
  mutate(lower = map_dbl(hpd, ~.x["simple_slope","lower"]),
         upper = map_dbl(hpd, ~.x["simple_slope", "upper"]))

hpds %>%
  ggplot(aes(x = z)) +
  geom_line(aes(y = simple_slope)) + 
  geom_line(aes(y = upper), lty = 2) + 
  geom_line(aes(y = lower), lty = 2) + 
  geom_hline(aes(yintercept = 0), lty = 3) + 
  theme_minimal()
```



## Session information

<!-- Insert the session information into the document -->
```{r session-info}
```
