---
title: "Bayes hypothesis testing"
author: "Wouter van Amsterdam"
date: 2018-05-02
output: html_document
---

<!-- The file analysis/chunks.R contains chunks that define default settings
shared across the workflowr files. -->
```{r read-chunk, include=FALSE, cache=FALSE}
knitr::read_chunk("chunks.R")
```

<!-- Update knitr chunk options -->
```{r knitr-opts-chunk, include=FALSE}
```

<!-- Insert the date the file was last updated -->
```{r last-updated, echo=FALSE, results='asis'}
```

<!-- Insert the code version (Git commit SHA1) if Git repository exists and R
 package git2r is installed -->
```{r code-version, echo=FALSE, results='asis'}
```

<!-- Add your analysis here -->

# Define hypotheses

Formulate meaningfull hypotheses based on theory


Bayes factor:

Balance fit with complexity of 2 hypothesis

Bayes factor: the support of the data for the informative hypothesis, 
is x times larger than the support for the uninformative hypothesis

Fit: scale 0 to 1 (1 is perfect)
Complexity: e.g. number of possible orderings of means (mu1 > mu2 > mu3 is 1 / 6 possible permutations)

Possible set of hypothesis

- Hi
- Hi` (competing hypothesis)
- Hc (complementing hypothesis)

Interpretation is in terms of odds (can be turned into conditional hypothesis probabilities, conditional on the data)

Possible to compute:

BF informative vs uninformative
BF informative vs complementing

Bayes factor

1. is a measure of support (also for null-hypothesis)
2. bayes factor can be indecisive (around 1)
3. can update: collect more data
4. can compar more than two hypotheses
5. something is going on, and we do know what

Drawbacks

- not controlling overall type 1 and 2 errors (only conditional errors)
- no cut-off value for decisions (need to think of what level of evidence you want)
- multiple operalizations of a single hypothesis possible (subjective, pre-formulate and publish)

Posterior model probability

- odds to probability -> Bayesian error probability (variant on type 1 and 2 errors)
- evaluates both fit and complexity
- not a classical probability
- add up to 1 for all hypotheses tested
- sensitive to number of hypotheses (only include relevant hypotheses)
- they are the probability of the hypothesis, given the data (not the probability of the data, given the hypothesis)

Don't develop model and perform hypothesis testing on the same data


Bayesian model update: study synthesis

Can combine PMPs from different studies

### Technical detials (in Bain)

Based on multi-normal distribution approximations of the posterior distribution

Prior distribution based on (part of) the data, should be centered at 0,
uninformative, and of the same shape as the posterior distribution

Complexity of hypothesis is the proportion of the prior distribution in 
agreement with the hypothesis

Fit is proportion of posterio distribution in agreement with the hypothesis






Bayesian model averaging / ensembling; (selecting subsets of predictors)

Utrecht Applied Data Science

6. Gu. Mulder, Approximate adjusted fractional Bayes factors





## Session information

<!-- Insert the session information into the document -->
```{r session-info}
```
