---
title: "Bayes Regression"
author: "Wouter van Amsterdam"
date: 2018-05-03
output: html_document
---

<!-- The file analysis/chunks.R contains chunks that define default settings
shared across the workflowr files. -->
```{r read-chunk, include=FALSE, cache=FALSE}
knitr::read_chunk("chunks.R")
```

<!-- Update knitr chunk options -->
```{r knitr-opts-chunk, include=FALSE}
```

<!-- Insert the date the file was last updated -->
```{r last-updated, echo=FALSE, results='asis'}
```

<!-- Insert the code version (Git commit SHA1) if Git repository exists and R
 package git2r is installed -->
```{r code-version, echo=FALSE, results='asis'}
```

<!-- Add your analysis here -->

# Bayes regressions

Dr. Milica Miocevic


Priors for regression coefficients

usually normal

sometimes uniform (e.g. when the range of Y is 0 to 9)


For sigma: inverse gamma

Non-intuitive parameterization of inverse gamma

Usefull based on prior data:

$$\sigma_{\epsilon}^2=IG(\nu_0/2, \nu_0\sigma_0^2/2)$$

Higher sample size -> more informative


Use gamma for precision is parameterized with the same $\alpha$ and $\beta$


For hierarchical models: even diffuse prios can have an influence on the posterior
(especially variance parameters)


Gelman and Shirley (2011) recommendations on convergence:

1. use at least 3 chains with dispersed starting values
2. PSR < 1.1 seems like convergence
3. Burn-in up to half of iterations


Potential scale reduction factor (PSRF) = Gelman statistic:

- function of ratio of between variability over within-chain variability


Credibility interval

- from quantiles: equal tail interval
- for asymmetrical posteriors: highest posterior density interval (narrowest possible interval that contains 95% of the density)


Interpretation: there is 95% probability that when people report 0 as their 
thirst, they will drink between x and y dl of water

In frequentialist analysis: 95% of intervals constructed following this research
setup will contain the true value (but this doesn't say anything about the 
interval at hand)


Cool think: credibility intervalas have probabilistic interpretation 


## Moderation analysis

- simple intercept = combine intercept and main effect of moderator
- simple slope = combine beta of determinant and beta of moderator

Usually pick values for contuous moderator of +1 and -1 SD, and report 
simple intercepts and slopes

## Mediator model

model for mediator based on predictor
model of x and moderator on y

Other possibility is Structural Equation Modeling with variance-covariance matrix

Product of 2 normally distributed variables is not normally distributed

With Bayes it is no problem, just sample from posterior distribution and
compute product


Multicollinearity in the models can lead to non-convergence
mean centering can help


Diffuse priors vs Bayesian: Miocevic MacKinon Levy 2016


Additional options in Bayesian:

- calculate from posterior the probability that the coefficient is larger 
than a certain cut-off







## Session information

<!-- Insert the session information into the document -->
```{r session-info}
```
