---
title: "Bootstrapping with recipes"
author: "Wouter van Amsterdam"
date: 2018-01-15
output: html_document
---

<!-- The file analysis/chunks.R contains chunks that define default settings
shared across the workflowr files. -->
```{r read-chunk, include=FALSE, cache=FALSE}
knitr::read_chunk("chunks.R")
```

<!-- Update knitr chunk options -->
```{r knitr-opts-chunk, include=FALSE}
```

<!-- Insert the date the file was last updated -->
```{r last-updated, echo=FALSE, results='asis'}
```

<!-- Insert the code version (Git commit SHA1) if Git repository exists and R
 package git2r is installed -->
```{r code-version, echo=FALSE, results='asis'}
```

<!-- Add your analysis here -->

## Preface 

From vignette of `rsample` and `recipes`

https://cran.r-project.org/web/packages/rsample/vignettes/Recipes_and_rsample.html


## Modeling 

### Setup

```{r}
library(recipes)
```

#### Get data
```{r}
library(AmesHousing)
ames <- make_ames()
```

### Create analysis pipeline for each (re)sample called a recipe

```{r}
rec <- recipe(Sale_Price ~ Neighborhood + House_Style + Year_Sold + Lot_Area,
              data = ames) %>%
  # transform outcome
  step_log(Sale_Price, base = 10) %>%
  # map low-count categories of factor variables to 'other' based on a threshold
  step_other(Neighborhood, House_Style, threshold = 0.05) %>%
  # make dummy variables of 'nominal' variables
  step_dummy(all_nominal()) %>%
  # unskew a predictor
  step_BoxCox(Lot_Area) %>%
  # normalize all
  step_center(all_predictors()) %>%
  step_scale(all_predictors())

rec
  
  
```

#### Prepare on training data

```{r}
rec_training_set <- prep(rec, training = ames, retain = T, verbose = T)
```

#### Get values on training data

```{r}
bake(rec_training_set, newdata = head(ames))
```

#### Get values of training data

```{r}
juice(rec_training_set) %>% head()
```


### Get resamples

```{r}
library(rsample)
set.seed(2)

bt_samples <- bootstraps(ames)
bt_samples
```

```{r}
dim(ames)
bt_samples$splits[[1]]
```

#### Add recipy to bootstrap samples

The function `prepper` from `rsample` which calls `prep` on the splits
```{r}
library(purrr)

bt_samples$recipes <- map(bt_samples$splits, prepper, 
                          recipe = rec, retain = TRUE, verbose = FALSE)

```

### Model on re-samples

Make generic `lm` function, add `lm` models to the bootstrep samples

```{r}
fit_lm <- function(rec_obj, ...) {
  lm(..., data = juice(rec_obj, everything()))
}

bt_samples$lm_mod <- 
  map(
    bt_samples$recipes, 
    fit_lm, 
    Sale_Price ~ .
  )

bt_samples
bt_samples$lm_mod[[1]]
```

#### Get predictions

```{r}
pred_lm <- function(split_obj, rec_obj, model_obj, ...) {
  mod_data <- bake(
    rec_obj, 
    newdata = assessment(split_obj), # assessments is like as.data.frame for `rsplit`
    all_predictors(),
    all_outcomes()
  ) 
  
  out <- mod_data %>% select(Sale_Price)
  out$predicted <- predict(model_obj, newdata = mod_data %>% select(-Sale_Price))
  out
}

bt_samples$pred <- 
  pmap(
    lst(
      split_obj = bt_samples$splits, 
      rec_obj = bt_samples$recipes, 
      model_obj = bt_samples$lm_mod
    ),
    pred_lm 
  )

bt_samples$pred[[1]]
```


#### Get RMSE

```{r}
rmse <- function(dat) 
  sqrt(mean((dat$Sale_Price - dat$predicted)^2))
bt_samples$RMSE <- map_dbl(bt_samples$pred, rmse)
summary(bt_samples$RMSE)
```



## Session information

<!-- Insert the session information into the document -->
```{r session-info}
```
