---
title: "Intro to intro to statistics, classical methods in statistics"
author: "Wouter van Amsterdam"
teacher: "Kruitwagen"
date: 2017-10-23
output: html_document
---

<!-- The file analysis/chunks.R contains chunks that define default settings
shared across the workflowr files. -->
```{r read-chunk, include=FALSE, cache=FALSE}
knitr::read_chunk("chunks.R")
```

<!-- Update knitr chunk options -->
```{r knitr-opts-chunk, include=FALSE}
```

<!-- Insert the date the file was last updated -->
```{r last-updated, echo=FALSE, results='asis'}
```

<!-- Insert the code version (Git commit SHA1) if Git repository exists and R
 package git2r is installed -->
```{r code-version, echo=FALSE, results='asis'}
```

<!-- Add your analysis here -->
## Setup
Teacher: Cas Kruitwagen, <c.l.j.j.kruitwagen@umcutrecht.nl>

### Load some R-packages
```{r, message = F}
library(data.table)
library(amstR)
library(magrittr)
library(ggplot2)
```


## Introduction to case studies
**Basic steps**

1. research objective
2. which variables in your datacan be analyzed to answer the objective
3. explore data
4. decide on statistical analysis
5. interpret the output
6. draw conclusions
7. relate results back to research question

### Definitions

* 'skew' direction: based on the **tail** of the data.

### Case study: Balance
Authors: Teasdal et al.

**Research question**

* Maintain balance when you are concentrating
* Is the difficulty more pronounced in the elderly
* Is there a difference between swaying in the forward/backward and side-to-side direction

Old: mean age 71.1
Young: mean age 24.8

**Variables**
FB: forwards / backwards motion (mm)
SS: Side-to-side motion (mm)
Age-group

```{r}
load(amstR::fromParentDir("data/balance.RData"))
summary(balance)
head(balance)
```

#### Plots
```{r}
boxplot(FB~age, data = balance)
boxplot(SS~age, data = balance)

# pairs(balance)
```

Distributions in scatterplot

```{r}
with(balance,
     amstR::marginal_plot(x = FB,
                          y = SS,
                          group = age))
```



#### Stats
```{r}
t.test(balance$FB)
t.test(FB~age, data = balance, paired = F)
t.test(SS~age, data = balance, paired = F)
t.test(balance$FB, balance$SS, paired = T)
t.test(balance$FB - balance$SS)
```



#### Conclusions

* Research question 1 is not answerable, there is no comparison between concentrating
* Whith a paired t-test, the amount of sway is different between forward/backward and sideward

### Case study Eggs
Quin(1988)
Effect of season (winter/spring and summer/autumn) and adult density (8, 15, 30, 45 per 225cm^2 enclosure), on prudiction of egg masses by intertidal pulmona limpets (Siphonaria diemenensis)

3 replicates per exposure situation

#### import data
```{r}
load(fromParentDir("data/eggs.Rdata"))
summary(limpet.eggs)
```

#### Plots
```{r}
boxplot(eggs~season, data = limpet.eggs)
boxplot(eggs~as.factor(density), data = limpet.eggs)
boxplot(eggs~season*density, data = limpet.eggs,
        col = c("blue","red"))

limpet.eggs %>%
  ggplot(aes(x = as.factor(density), y = eggs, col = season)) + 
  geom_boxplot() + 
  geom_point()
```

#### Stats

```{r}
fit0 <- lm(eggs~season + density, data = limpet.eggs)
fit1 <- lm(eggs~season + density + season*density, data = limpet.eggs)
summary(fit0)
summary(fit1)
```

```{r}
plot(fit0)
```

### Case study: platelet calcium and blood pressure
Reation between calcium in cells (platetelet) and blood pressure
* 38 subjects with normal blood pressure
* Blood pressure (average of systolic and diastolic blood pressure)
* Free calcium concentration in blood platelets

#### Import data
```{r}
load(fromParentDir("data/bloodpressurecalcium.RData"))
```

#### Plots
```{r}
plot(calcium~bp, data = bloodpressurecalcium)
```


#### Stats
```{r}
fit0 <- lm(calcium~bp, data = bloodpressurecalcium)
summary(fit0)
plot(fit0)
cor.test(~bp+calcium, data = bloodpressurecalcium)
```

### SO2 (with latex $SO_{2}$)
Air pollution with sulfur dioxide, $SO_{2}$

-	Temp	mean year temperature (in oF)
-	Factories	number of factories with more than 20 employees
-	Population	size of the population (in thousands)
-	Wind	mean wind speed over a year (miles per hour)
-	Rainfall	mean yearly rainfall (in inches)
-	Daysrain	mean number of days of rainfall per year
-	SO2	The amount of SO2 in the air (in μg/m3)

#### import data
```{r}
load(fromParentDir("data/so2.RData"))
summary(so2)
head(so2)
```

#### Plots
```{r}
plot(SO2~temp, data = so2)
boxplot(SO2~region, data = so2)
pairs(so2[, setdiff(colnames(so2), c("city"))])
```

#### Stats
```{r}
fit_all <- lm(
  reformulate(termlabels = setdiff(colnames(so2), c("SO2", "city", "factories")),
              response = "SO2"), 
  data = so2)
plot(fit_all)
summary(fit_all)
```



## Session information

<!-- Insert the session information into the document -->
```{r session-info}
```
