---
title: "Correlation and Regression"
author: "Wouter van Amsterdam"
date: 2017-11-15
output: html_document
---

<!-- The file analysis/chunks.R contains chunks that define default settings
shared across the workflowr files. -->
```{r read-chunk, include=FALSE, cache=FALSE}
knitr::read_chunk("chunks.R")
```

<!-- Update knitr chunk options -->
```{r knitr-opts-chunk, include=FALSE}
```

<!-- Insert the date the file was last updated -->
```{r last-updated, echo=FALSE, results='asis'}
```

<!-- Insert the code version (Git commit SHA1) if Git repository exists and R
 package git2r is installed -->
```{r code-version, echo=FALSE, results='asis'}
```

<!-- Add your analysis here -->

### Pearson vs Spearman
Spearman = Pearson on ranks

```{r}
set.seed(3)
n = 20
noise = .4
x = runif(n)
y = x + rnorm(n, sd = noise)
plot(x,y)
cor.test(~x+y, method = "p")
cor.test(~x+y, method = "s")
cor.test(~rank(x)+rank(y), method = "p")
```

Note that the estimate is equal for spearman vs pearson of rank, however, 
the p-value for pearson on the ranks is lower. 

Is this always the case?

```{r}
set.seed(2)
nsim = 100
n = 20
noise = .4

sim <- data.frame(
  rp  = numeric(nsim),
  rs  = numeric(nsim),
  rpr = numeric(nsim),
  blm = numeric(nsim),
  sqr_rsq = numeric(nsim),
  pp  = numeric(nsim),
  ps  = numeric(nsim),
  ppr = numeric(nsim),
  plm = numeric(nsim)
)

for (i in 1:nsim) {
  x = runif(min = -1, max = 1, n = n)
  y = x + rnorm(n, sd = noise)
  cor_p  <- cor.test(~x+y, method = "p")
  cor_s  <- cor.test(~x+y, method = "s")
  cor_pr <- cor.test(~rank(x)+rank(y), method = "p")
  lm_fit <- lm(y~x)
  sum_fit<- summary(lm_fit)
  sim[i, ] <- c(cor_p$estimate, cor_s$estimate, cor_pr$estimate, 
                lm_fit$coefficients[2],
                sqrt(sum_fit$r.squared),
                cor_p$p.value, cor_s$p.value, cor_pr$p.value,
                sum_fit$coefficients[2,4])
}

plot(rs~rpr, data = sim)
plot(ps~ppr, data = sim)
plot(rp~sqr_rsq, data = sim)
plot(sim$ps-sim$ppr)
plot(sim$pp-sim$plm)
```

Conclusions

* Pearson correlation = $R^2$ from simple linear regression with intercept, 
and p-values are equal
* Spearman correlation = Pearson on the ranks of the data, 
p-values differ in that the spearman p-values are sometimes higher.

### LM predictions
```{r}
set.seed(3)
n = 10
noise = .4
x = runif(min = -2, max = 2, n)
x = rnorm(n)
y = x + rnorm(n, sd = noise)

fit <- lm(y~x) 
summary(fit)

plot(x,y)
abline(fit, lty = 2)

x_preds  <- seq(from = floor(min(x)), to = ceiling(max(x)), length.out = 50)
yhat_pred <- predict(fit, newdata = data.frame(x = x_preds),
                     what = "terms", interval = "confidence")
y_pred    <- predict(fit, newdata = data.frame(x = x_preds),
                     what = "response", interval = "prediction")

points(x_preds, yhat_pred[,2], pch = 3, col = "blue")
points(x_preds, yhat_pred[,3], pch = 3, col = "blue")
points(x_preds, y_pred[,2], pch = 3, col = "red")
points(x_preds, y_pred[,3], pch = 3, col = "red")
```


## Session information

<!-- Insert the session information into the document -->
```{r session-info}
```
