---
title: "Case Studies - not finished"
author: "Wouter van Amsterdam"
date: 2017-11-10
output: html_document
---

<!-- The file analysis/chunks.R contains chunks that define default settings
shared across the workflowr files. -->
```{r read-chunk, include=FALSE, cache=FALSE}
knitr::read_chunk("chunks.R")
```

<!-- Update knitr chunk options -->
```{r knitr-opts-chunk, include=FALSE}
```

<!-- Insert the date the file was last updated -->
```{r last-updated, echo=FALSE, results='asis'}
```

<!-- Insert the code version (Git commit SHA1) if Git repository exists and R
 package git2r is installed -->
```{r code-version, echo=FALSE, results='asis'}
```

<!-- Add your analysis here -->
## NB Not finished
Please note that this is hasty work, most of the assignments are not thoroughly 
answered.

## Instructions

Use the morning to examine the datasets for the case studies. For each case:
1.	get the appropriate descriptive statistics;
2.	form an analysis plan;
3.	try to develop a sense (“gut feeling”) of what the results of the analysis will be; and
4.	if you have time, perform a full analysis on at least one case.

Remember the “Steps of Data Analysis” presented on the first day of the course!

You may work on your own or in a group, at your own computer or at the scheduled location, or in one of the computer rooms (if available) or in ‘Studielandschap’. Note that there is no assistance available during the morning.

During the lecture (13.30 to 17.00 at the very latest) we will discuss the analysis of the cases. There will also be plenty of time for you to ask questions.

## The Case Studies
### Case Study 1: fruitflies
For female fruitflies it is known that increased reproduction leads to a shortened life span. For male fruitflies this has never been ascertained. Male fruitflies that mate on average once a day are known to live for 60 days. The question is whether manipulation of sexual activity influences life span. To answer this question the following experiment was done: For one specific species of fruitflies, sexual activity was manipulated by providing individual males with either one or eight receptive, virginal females per day. Further, individual males were provided with one or eight recently inseminated females (these females do not wish to be inseminated again). A number of males were kept solitarily. Relevant variables were coded as follows: The data are given in the file fruitfly.sav or fruitfly.RData

variable | explanation
---|---
number | number of females provided
type | type of company (0: recently inseminated, 1: virginal, 9: none)
day | day=0 if life span is shorter than 60 days, day=1 if life span is longer than 60 days
thorax | thorax length of male
sleep | average percentage of time spent sleeping per day

#### Research objective
Answer the question: do male fruitflies have a shorte life span when they 
reproduce more?

#### Variables of interest
* determinant: number AND type
* outcome: day
* possible confounders: thorax, sleep

Number and type will be included, as they will together determine how many 
times a male fruitfly may reproduce. Thorax and sleep are included, because 
both could be related to the outcome, and possibly the determinant.

#### Data exploration
```{r}
load(epistats::fromParentDir("data/fruitfly.RData"))
str(fruitfly)
```

##### Data curation
Number is coded as numeric. However, it only has two distinct values: 0 and 8.
It may be better to recode this as a categorical variable.

```{r}
table(fruitfly$NUMBER)
# fruitfly$NUMBER <- factor(fruitfly$NUMBER)
```

##### Pairs plot
```{r, warning = F, message = F}
require(GGally)
GGally::ggpairs(fruitfly)
```

From this exploratory analysis we can observe
Day seems to be associated with thorax, and maybe with number and type too.

The distribution of thorax is a little left-skewed, while sleep is somewhat 
right-skewed.


#### Analysis plan
Day is a binary outcome, we are interested in the rate of Day > 60. 
For univariate analysis, we can create a contingency table and perform a
chi-squared test,
 for multivariate analysis we can use logistic regression. As this is an 
 etiologic question, we will include all possible confounders in the 
 multiple regression model. We expect that type may modify the effect of 
 number (as this determines the number of available mates), so we will include 
 an interaction term with type and number.

Other possibility: recode number and type into 5 groups, do chi-squared

#### Interpretation of results
##### Crude analysis
```{r}
tab <- xtabs(~NUMBER+DAY, data = fruitfly)
tab
prop.table(tab, margin = 1)
chisq.test(tab)
```
In crude analysis, there seems to be no statistically significant association 
between day and number.

##### Multiple regression
```{r}
fit <- glm(DAY~NUMBER*TYPE+THORAX+SLEEP, data = fruitfly, family = binomial)
fit_scaled <- glm(DAY~NUMBER*TYPE+scale(THORAX, center = T, scale = T)+SLEEP, data = fruitfly, family = binomial)
fit_parsimonious <- glm(DAY~NUMBER*TYPE, data = fruitfly, family = binomial)
summary(fit)
summary(fit_parsimonious)
lmtest::lrtest(fit)
drop1(fit)
```

Longevity can partly be explained by the provided model. 
The length of the male (`THORAX`) seems to by the most important factor, 
fruitflies that are longer seem to live longer.  
Number on itself is not significantly associated with the outcome. 
However from the interaction term, we see that the number of virginal female 
fruitflies is associated with shorter survival. Sleep does not seem to 
contribute significantly to the model.

#### Conclusions
Male fruitlies that reproduce more often seem to have a shorter life-span. 
However, the size of the thorax of the fruitfly is a more important predictor, 
with larger fruitflies living longer.

### Case Study 2, Heroin rehab clinic
Caplehorn (1991) was interested in factors that have an effect on the success probability of retention of heroin addicts in a clinic. In a multi-center study he collected the following data from 238 heroin addicts. The data are given in the file heroin.sav or heroin.RData. 

variable | explanation
---|---
ID | Identification code
CLINIC | Number of clinic (1 or 2)
STATUS | Status (0=not (yet) departed; 1= departed from clinic)
TIMES | Time in days that heroin addicts spent in the clinic
PRISON | Prison record (1=yes; 0=no)
DOSE | Methadone dosage in mg/day	

(Source: Caplehorn J., 1991, Methadone dosage and retention of patients in maintenance treatment, Medical Journal of Australia, 154, 195 99)

The researchers are only interested in the effect of each explanatory variable separately. Would you advise further statistical analysis?


#### Research objective
##### Research question
What factors are associated with retention of heroin addicts in a clinic.

#### Variables of interest
* determinants: CLINIC, TIMES, PRISON, DOSE
* outcome: STATUS
* possible confounders:

#### Data exploration
```{r}
load(epistats::fromParentDir("data/heroin.RData"))
str(heroin)
```

##### Data curation
Let's throw out ID, since it is uninformative and throwing it out means that 
we don't have to do that each time in the analysis

Clinic, status and prison should be coded as factor variables

```{r, message = F, warning = F}
heroin$ID <- NULL

require(dplyr)
heroin <- heroin %>%
  mutate_at(.vars = c("CLINIC", "STATUS", "PRISON"), factor)

str(heroin)
```


##### Pairs plot
```{r, message = F, warning=F}
require(ggplot2)
GGally::ggpairs(heroin)
```

Satus seems to be related with times, clinic, and maybe prison. However, 
clinic and times also seem to be related, so it could be that one of them 
explains the effect of the other. Also, TIMES and DOSE seem to be associated 
with each other.

#### Analysis plan
We will use multiple logistic-regression analysis, as the effect of any single 
factor (or lack thereof) may be confounded by other factors.

Better plan: treat as survival data, predict how long they stay

```{r}
require(survival)
surv_obj <- Surv(time = heroin$TIMES, event = heroin$STATUS)
km_fit <- survfit(Surv(time = TIMES, event = STATUS==1)~1, data = heroin)
plot(km_fit)

lines(survfit(Surv(time = TIMES, event = STATUS==1)~CLINIC, data = heroin),
     main = "retention by clinic", col = c("red", "blue"))


plot(survfit(Surv(time = TIMES, event = STATUS==1)~CLINIC, data = heroin),
     main = "retention by prison", col = c("red", "blue"))

plot(survfit(Surv(time = TIMES, event = STATUS==1)~epistats::quant(DOSE), data = heroin),
     main = "retention by quantiles of dose", lty = 1:4)



```

```{r}
cox_fit <- coxph(Surv(time = TIMES, event = STATUS==1)~CLINIC+PRISON+DOSE, data = heroin)
summary(cox_fit)
```

We can use logrank test for 2 groups

##### Crude analysis

##### Multiple regression

#### Interpretation of results


#### Conclusions


### Case Study 3, Maternal mortality rates
Researchers examined maternal mortality rates (MMRs) in relation to the global distribution of physicians and nurses.  A UN database on 155 countries was constructed in order to determine whether MMRs were related to the proportion of births attended by medical and nonmedical staff (defined as “attendance at birth by trained personnel” (physicians, nurses, midwives or primary health care workers trained in midwifery skills), the ratio of physicians and nurses to the population, female literacy and gross national product (GNP) per capita. The data are given in the file maternalmort.sav or maternalmort.RData.

variable | explanation
---|---
Country | name of country
Physician | physicians per 1000 population
Nurses | nurses per 1000 population
GNP | gross national product per capita in dollars
MMR | maternal mortality rate (per 100,000 births)
Attended | births attended by trained personnel (%)
Femlit | female literacy (%)

#### Research objective
##### Research question
What is the relationship between the proportion of births attended by 
trained medical staff and maternal mortality rates, 
adjusted for gross national product per capita and female literacy

#### Variables of interest
* determinant: Attended
* outcome: MMR
* possible confounders: Physician, Nurses, GNP, Femlit

#### Data exploration
```{r}
load(epistats::fromParentDir("data/maternalmort.RData"))
str(maternalmort)
```

##### Data curation

##### Pairs plot

```{r}
GGally::ggpairs(maternalmort[, -1])
```

```{r, message = F, warning = F}
require(dplyr)
maternalmort %>%
  mutate_at(.vars = c("physician", "nurses", "GNP"), function(x) log(x+.01)) %>%
  select(-Country) %>%
  GGally::ggpairs()
```

Looks like strong collinearity for attended and femlit

#### Analysis plan
We are trying to model a continous variable: MMR. Strictly speaking it is a 
rate, which should optimally be modeled with Poisson regression, 
however we do not have the original counts. 
However for large enough rates, the poisson distribution opproximates the normal 
distribution, so we can also use (multiple) linear regression.


##### Crude analysis

##### Multiple regression

```{r}
fit1 <- maternalmort %>%
  mutate_at(.vars = c("physician", "nurses", "GNP"), 
            function(x) log(x + 0.01)) %>%
  select(-Country) %>%
  lm(MMR~., data = .)

plot(fit1, which = c(1,2))

fit2 <- maternalmort %>%
  mutate_at(.vars = c("physician", "nurses", "GNP", "MMR"), 
            function(x) log(x + 0.01)) %>%
  select(-Country) %>%
  lm(MMR~., data = .)

plot(fit2, which = c(1,2))

fit3 <- maternalmort %>%
  mutate_at(.vars = c("MMR"), 
            function(x) log(x + 0.01)) %>%
  select(-Country) %>%
  lm(MMR~., data = .)

plot(fit3, which = c(1,2))

fit4 <- maternalmort %>%
  mutate_at(.vars = c("MMR"), 
            function(x) log(x + 0.01)) %>%
  select(-c(Country, femlit)) %>%
  lm(MMR~., data = .)

plot(fit4, which = c(1,2))
drop1(fit4, test = "F")


fit5 <- maternalmort %>%
  mutate_at(.vars = c("MMR"), 
            function(x) log(x + 0.01)) %>%
  select(-c(Country, femlit, nurses)) %>%
  lm(MMR~., data = .)

plot(fit5, which = c(1,2))
drop1(fit5, test = "F")
1/car::vif(fit5)
```

#### Interpretation of results


Look at Countries with highest residual, which means high rates, that is 
unexplained by the model.

```{r}
resids <- fit5$residuals
ordered_resids <- sort(resids, decreasing = T)
included_observations <- names(ordered_resids)


data.frame(residual = ordered_resids,
           maternalmort[names(ordered_resids),])

```



#### Conclusions



### Case Study 4, Spider web
Some biologists were interested in the effect of temperature on the size of the cobweb from the spider Achaearanea tepidariorum. They collected the following data (see the file cobweb.sav or cobweb.RData):



Temperature
(in oC)	log10(cobweb weight/spider weight)
10	-2.81508     -2.74254     -2.69473     -2.58087     -2.30309
-2.82513     -2.63073     -2.10109
15	-2.23378     -2.85680     -2.31367     -2.93023     -2.69093
-2.56165     -2.61604     -2.97926
20	-1.97753     -1.98788     -1.92117     -2.42344     -2.53879
-2.29301     -2.19948     -2.27348     -2.13957     -2.43604
25	-2.02078     -2.54437     -2.41178     -1.95839     -3.02343
-2.63536     -2.63536     -2.75174     -1.99863

Analyze these data and give possible explanations for the transformation of the dependent variable.


#### Research objective
##### Research question
Is the size of the cobweb of the Achaearanea tepiariorum dependant on the 
temperature?

#### Variables of interest
* determinant: Temperature
* outcome: Size of cobweb (= cobweb weight/spider weight)
* possible confounders:


#### Data exploration
```{r}
load(epistats::fromParentDir("data/cobweb.RData"))
str(cobweb)
```



##### Data curation

##### Pairs plot
```{r}
boxplot(LGWEIGHT~TEMP, data = cobweb)
```

From the boxplots, it seems that LGWEIGHT increases with temperature. 
Also, the variance of LGWEIGHT seems to increase a little with temperature.

#### Analysis plan
Since we have a continous outcome and one categorical variable, we can analyse 
this with ANOVA (or equivalently: simple linear regression). We can treat 
temperature as a continous variable, or as a categorical one. Since we 
expect a monotonous relationship on the range of tested temperatures, we will 
use it as a continous variable, 
as this keeps in the information that $temp = 10 < 15 < 20 < 25$

##### Crude analysis
```{r}
fit <- lm(LGWEIGHT~TEMP, data = cobweb)
summary(fit)
plot(fit, which = c(1,2))
cor.test(~LGWEIGHT+TEMP, data = cobweb, method = "spearman")
```
From the analysis, it looks like there is 

```{r}
fit_anova <- aov(LGWEIGHT~factor(TEMP), data = cobweb)
summary(fit_anova)
TukeyHSD(fit_anova)
```



##### Multiple regression

#### Interpretation of results


#### Conclusions


### Case Study 5, immediate vs sustained release
Immediate release medications quickly liberate their drug content into the body, with the maximum concentration reached in a short time followed by a rapid decline in concentration. Sustained release medications, on the other hand, take longer to reach maximum concentration in the body and stay active for longer periods of time.

A study of two such pain relief medications compared immediate release codeine (IRC) with sustained release codeine (SRC). Thirteen healthy patients were randomly assigned to one of the two types of codeine and treated for 2.5 days. After a 7-day wash-out period, the same patients were given the other type of codeine. Measurements include the maximum concentration in ng/mL, the time to maximum concentration in hours, the total amount of drug available over the life of the treatment in (ng*mL)/hr and the age of the patient. These data are shown below (see also the file codeine.sav or codeine.RData.

	Concentration	Time	Total amount
Age	IRC	SRC	IRC	SRC	IRC	SRC
33	181.8	195.7	0.5	2.0	1091.3	1308.5
40	466.9	167.0	1.0	3.0	1064.5	1494.2
41	136.0	217.3	0.5	3.0	1281.1	1382.2
43	221.3	375.7	1.5	4.5	1921.4	1978.3
25	195.1	285.7	0.5	2.0	1649.9	2004.6
30	112.7	177.2	1.0	2.0	1423.6	*
24	84.2	220.3	2.0	1.5	1308.4	1211.1
44	78.5	243.5	1.0	3.0	1192.1	1002.4
42	85.9	141.6	1.5	1.5	766.2	866.6
33	85.3	127.2	2.0	4.5	978.6	1345.8
38	217.2	345.2	0.5	1.5	1618.9	979.2
39	49.7	112.1	1.5	1.0	582.9	576.3
43	190.0	223.4	0.5	1.0	972.1	999.1

The research questions are: are there any differences on these three parameters between immediate release codeine and sustained release codeine? Are these differences dependent on age?


#### Research objective
Assume random order.

##### Research question

#### Variables of interest
* determinant:
* outcome:
* possible confounders:


#### Data exploration
```{r}
load(epistats::fromParentDir("data/codeine.RData"))
str(codeine)
codeine <- as.data.frame(codeine)
```



##### Data curation

Melt into easier format
```{r}
codein_melted <- data.table::melt(codeine, id.vars = "AGE")
```



##### Pairs plot

#### Analysis plan
We will use the parametric tests. Too little data to check assumptions, 
too little data for power for non-parametric tests.

##### Crude analysis
```{r}
t.test(codeine)
```



##### Multiple regression
Dependent on age:

Difference ~ age (dependent -> regression, associated -> correlation)


#### Interpretation of results


#### Conclusions

## General structure
(for later take heading levels up 1 step)
#### Case description

#### Research objective
##### Research question

#### Variables of interest
* determinant:
* outcome:
* possible confounders:


#### Data exploration

##### Data curation

##### Pairs plot

#### Analysis plan
##### Crude analysis

##### Multiple regression

#### Interpretation of results


#### Conclusions



## Session information

<!-- Insert the session information into the document -->
```{r session-info}
```
