---
title: "Positioning as traingles with equal sides"
author: "Wouter van Amsterdam"
date: 2018-02-03
output: html_document
---

<!-- The file analysis/chunks.R contains chunks that define default settings
shared across the workflowr files. -->
```{r read-chunk, include=FALSE, cache=FALSE}
knitr::read_chunk("chunks.R")
```

<!-- Update knitr chunk options -->
```{r knitr-opts-chunk, include=FALSE}
```

<!-- Insert the date the file was last updated -->
```{r last-updated, echo=FALSE, results='asis'}
```

<!-- Insert the code version (Git commit SHA1) if Git repository exists and R
 package git2r is installed -->
```{r code-version, echo=FALSE, results='asis'}
```

<!-- Add your analysis here -->

# Setup

## Game definition

Consider the following challenge: a group of 50 people stands in a circle,
each one is told to secretly pick two other persons ('marks'). The objective for 
everyone is to move to such a position that they form a triangle with 
equal sides with respect to their two marks.

To add some details, say they are standing in a room with dimensions 
20 by 20 meters. 
For starters, we will assume that people's positions may coincide. 
If this is not the case, it is immediately clear that there does not always 
exist a solution: if 3 persons happen to choose the same 2 marks, since 
when 2 marks are chosen, exactly 2 points exist that form a traingle of 
equal sides with them (given that they the 2 marks don't happen to stand 
on the same position). However, the chance of this happening is very small,
the probability for 2 out of $n$ people to have chosen the same two marks is
$(n-1) * \frac{1}{n\choose{2}}$, the chance for 3 persons to have chosen the 
same 2 marks is even lower.
Assume that the stride length is the same for everyone: 0.5 meters per step.
People pick their marks randomly (you could pick other strategies,
 but this adds some nice randomness to the process)

## R equivalent

Let's put in in a picture

Initialize parameters, and load needed R-packages
```{r, message = F}
library(data.table)
library(ggplot2)
library(magrittr)

npeople = 8
room_dims = c(20,20)
stride = 0.5
pspace = 0.3
```

Initialize circle

```{r}
r0 = 3 # radius of initial circle

df <- data.table(
  id = 1:npeople
)

df[, c("x0", "y0"):=
     list(r0 * cos(2*pi*(id-1)/npeople),
          r0 * sin(2*pi*(id-1)/npeople))]

df[, id:=as.factor(id)]

circle0 <- df %>%
  ggplot(aes(x = x0, y = y0, col = id)) + 
  geom_point() + 
  lims(x = c(-1,1)*room_dims[1]/2, y = c(-1,1)*room_dims[2]/2) +
  coord_fixed() + 
  theme_minimal() +
  theme(legend.position = "none")

circle0_labeled <- 
  circle0 + 
  geom_label(aes(label = id))

circle0_labeled
```

Let's pick random marks for each individual (excluding themselves)

```{r}
set.seed(1234)

df[, c("mark_a", "mark_b"):=data.frame(t(apply(.SD, 1, function(row) 
  sample(setdiff(factor(1:npeople), row[["id"]]), size = 2, replace = F))))]

setkey(df, id)

df[1:10, list(id, mark_a, mark_b)]
```



Visualize a few subject and mark pairs

```{r}
df[, c("ma_x0", "ma_y0"):=.SD[as.character(mark_a), list(x0, y0)]]
df[, c("mb_x0", "mb_y0"):=.SD[as.character(mark_b), list(x0, y0)]]
```


```{r}
circle0 +
  geom_segment(data = df[c(1)], 
               aes(x = x0, xend = ma_x0, y = y0, yend = ma_y0), 
               linetype = 2) + 
  geom_segment(data = df[c(1)], 
               aes(x = x0, xend = mb_x0, y = y0, yend = mb_y0),
               linetype = 2)
```




# Model behaviour

Now let's see if we can converge to a solution where everyone get's to a position
 in which they are in a equal side triangle with their two targets.

## Solution 1: single step towards goal

Say everyone has a keen eye for where they should stand, given that 
their targets stand at points $x_{target_1},y_{target_1}$ and 
$x_{target_2}, y_{target_2}$.
With some math, we can see that the subject should move to one of 2 positions.

The target point for each subject lies on the circle with as center one of
the marks (let's take mark 'a' for simplicity), with a radius of the distance 
between mark a and mark b, and at a position of 60 degrees more or less than 
the angle between mark a and b.

Some restritions apply to this target point:

* it must be within the room
* when there are two options, choose the one that's nearest by

### Set target position

Let's create a function to determine the target point for a subject

```{r}
get_target.single <- function(x0, y0, ma_x0, ma_y0, mb_x0, mb_y0,
                              if_none = "original") {
  
  p0  = c(x0, y0)
  ma = c(ma_x0, ma_y0)
  mb = c(mb_x0, mb_y0)
  m_center = (ma + mb) / 2
  
  # calculate distance between mark a and b, and the angles between them
  theta_a  = atan2(ma_y0, ma_x0)
  theta_b  = atan2(mb_y0, mb_x0)
  theta_ab = theta_b - theta_a

  d_ab     = sqrt(sum((mb - ma) ^ 2))
  
  # target points are determined by taking a step from the middle between 
  # the marks
  stride   = (.5*d_ab) / tan(.5 * (1/6) * 2 * pi)

  # calculate candidate target points
  tx1 = m_center[1] + stride * cos(theta_a + theta_ab / 2)
  ty1 = m_center[2] + stride * sin(theta_a + theta_ab / 2)
  tx2 = m_center[1] - stride * cos(theta_a + theta_ab / 2)
  ty2 = m_center[2] - stride * sin(theta_a + theta_ab / 2)
  # ty1 = ma_y0 + radius * sin(angle_ab + 2*pi*60/360)
  # tx2 = ma_x0 + radius * cos(angle_ab - 2*pi*60/360)
  # ty2 = ma_y0 + radius * sin(angle_ab - 2*pi*60/360)

  # check if both target points are within the room
  t1_ok = abs(tx1) < room_dims[1] / 2 & abs(ty1) < room_dims[2] / 2
  t2_ok = abs(tx2) < room_dims[1] / 2 & abs(ty2) < room_dims[2] / 2
  
  # if 1 of both is ok, take that point
  if (sum(c(t1_ok, t2_ok)) == 1) {
    if (t1_ok) return(c(tx1, ty1))
    return(c(tx2, tx2))
  } 
  
  # if none are ok, return NA's or the original position
  if (sum(c(t1_ok, t2_ok)) == 0) {
    if (if_none == "original") return(c(x0,y0))
    return(c(NA,NA))
  }
  
  # if both are ok, take the one that's closest by, or the first one
  d1 = sqrt(sum((c(tx1, ty1) - p0) ^ 2))
  d2 = sqrt(sum((c(tx2, ty2) - p0) ^ 2))
  
  if (d1 <= d2) {
    return(c(tx1, ty1))
  }
  return(c(tx2, ty2))
}

get_target <- function(x0, y0, ma_x0, ma_y0, mb_x0, mb_y0) {
  df <- cbind(x0, y0, ma_x0, ma_y0, mb_x0, mb_y0)
  targets <- apply(df, 1, function(x) {
    get_target.single(x[[1]], x[[2]], x[[3]], x[[4]], x[[5]], x[[6]])
  })
  data.frame(t(targets))
}
  
df[, c("tx", "ty"):=get_target(x0, y0, ma_x0, ma_y0, mb_x0, mb_y0)]

```


Check correctness of solutions

```{r}
df[, t_correct:={
  da  = norm(matrix(c(tx - ma_x0, ty - ma_y0)), type = "F")
  db  = norm(matrix(c(tx - mb_x0, ty - mb_y0)), type = "F")
  dab = norm(matrix(c(mb_x0 - ma_x0, mb_y0 - ma_y0)), type = "F")
  max(dist(c(da, db, dab)) < 0.001)
}, by = "id"]
```

Visualize target positions for a few subjects

```{r}
samples = sample(1:npeople, size = 2)
circle0 + 
  geom_segment(data = df[samples], 
               aes(x = tx, xend = ma_x0, y = ty, yend = ma_y0)) +
  geom_segment(data = df[samples], 
               aes(x = tx, xend = mb_x0, y = ty, yend = mb_y0)) +
  geom_segment(data = df[samples], 
               aes(x = ma_x0, xend = mb_x0, y = ma_y0, yend = mb_y0)) +
  geom_segment(data = df[samples], 
               aes(x = x0, xend = tx, y = y0, yend = ty),
               arrow = arrow(), linetype = 2, alpha = .5)
```

### Take step to target position

Create a function to define a step to the target position

```{r}
take_step.single <- function(x0, y0, tx, ty, stride = .5) {
  direction <- c(tx, ty) - c(x0, y0)
  direction_normalized <- direction / sqrt(sum(direction^2))
}
```


To track the positions for each participants, let's create a new 
data structure.

```{r, message=F, eval = F}
# library(tibbletime)
# nsteps = 100
# 
# positions <- create_series('00:00:00' ~ '00:05:00', 'second', class = "hms")

```


```{r}
nsteps = 2

positions <- array(dim = c(nsteps, npeople, 2), 
                   dimnames = list(paste0("step-",1:nsteps),
                                   paste0("id-", 1:npeople), 
                                   c("x", "y")))

# for (i in 1:nsteps) {
  # positions[i,,] = 
# }
```


## Session information

<!-- Insert the session information into the document -->
```{r session-info}
```
