---
title: "Positioning as traingles with equal sides"
author: "Wouter van Amsterdam"
date: 2018-02-03
output: html_document
---

<!-- The file analysis/chunks.R contains chunks that define default settings
shared across the workflowr files. -->
```{r read-chunk, include=FALSE, cache=FALSE}
knitr::read_chunk("chunks.R")
```

<!-- Update knitr chunk options -->
```{r knitr-opts-chunk, include=FALSE}
```

<!-- Insert the date the file was last updated -->
```{r last-updated, echo=FALSE, results='asis'}
```

<!-- Insert the code version (Git commit SHA1) if Git repository exists and R
 package git2r is installed -->
```{r code-version, echo=FALSE, results='asis'}
```

<!-- Add your analysis here -->

# Setup

## Game definition

Consider the following challenge: a group of 50 people stands in a circle,
each one is told to secretly pick two other persons ('marks'). The objective for 
everyone is to move to such a position that they form a triangle with 
equal sides with respect to their two marks.

To add some details, say they are standing in a room with dimensions 
20 by 20 meters. 
For starters, we will assume that people's positions may coincide. 
If this is not the case, it is immediately clear that there does not always 
exist a solution: if 3 persons happen to choose the same 2 marks, since 
when 2 marks are chosen, exactly 2 points exist that form a traingle of 
equal sides with them (given that they the 2 marks don't happen to stand 
on the same position). However, the chance of this happening is very small,
the probability for 2 out of $n$ people to have chosen the same two marks is
$(n-1) * \frac{1}{n\choose{2}}$, the chance for 3 persons to have chosen the 
same 2 marks is even lower.
Assume that the stride length is the same for everyone: 0.5 meters per step.
People pick their marks randomly (you could pick other strategies,
 but this adds some nice randomness to the process)

## R equivalent

Let's put in in a picture

Initialize parameters, and load needed R-packages
```{r, message = F}
library(data.table)
library(ggplot2)
library(magrittr)

npeople = 8
room_dims = c(20,20)
stride = 0.5
pspace = 0.3
```

Initialize circle

```{r}
r0 = 3 # radius of initial circle

df <- data.table(
  id = 1:npeople
)

df[, c("x0", "y0"):=
     list(r0 * cos(2*pi*(id-1)/npeople),
          r0 * sin(2*pi*(id-1)/npeople))]

df[, id:=as.factor(id)]

circle0 <- df %>%
  ggplot(aes(x = x0, y = y0, col = id)) + 
  geom_point() + 
  lims(x = c(-1,1)*room_dims[1]/2, y = c(-1,1)*room_dims[2]/2) +
  coord_fixed() + 
  theme_minimal() +
  theme(legend.position = "none")

circle0_labeled <- 
  circle0 + 
  geom_label(aes(label = id))

circle0_labeled
```

Let's pick random marks for each individual (excluding themselves)

**NB improve code with 'by = "id"'

```{r}
set.seed(1234)

df[, c("mark_a", "mark_b"):=data.frame(t(apply(.SD, 1, function(row) 
  sample(setdiff(factor(1:npeople), row[["id"]]), size = 2, replace = F))))]

setkey(df, id)

df[1:10, list(id, mark_a, mark_b)]
```



Visualize a few subject and mark pairs

```{r}
df[, c("ma_x0", "ma_y0"):=.SD[as.character(mark_a), list(x0, y0)]]
df[, c("mb_x0", "mb_y0"):=.SD[as.character(mark_b), list(x0, y0)]]
```


```{r}
circle0 +
  geom_segment(data = df[c(1)], 
               aes(x = x0, xend = ma_x0, y = y0, yend = ma_y0), 
               linetype = 2) + 
  geom_segment(data = df[c(1)], 
               aes(x = x0, xend = mb_x0, y = y0, yend = mb_y0),
               linetype = 2)
```




# Model behaviour

Now let's see if we can converge to a solution where everyone get's to a position
 in which they are in a equal side triangle with their two targets.

## Solution 1: single step towards goal

Say everyone has a keen eye for where they should stand, given that 
their targets stand at points $x_{target_1},y_{target_1}$ and 
$x_{target_2}, y_{target_2}$.
With some math, we can see that the subject should move to one of 2 positions.

The target point for each subject lies on the circle with as center one of
the marks (let's take mark 'a' for simplicity), with a radius of the distance 
between mark a and mark b, and at a position of 60 degrees more or less than 
the angle between mark a and b.

Some restritions apply to this target point:

* it must be within the room
* when there are two options, choose the one that's nearest by

### Set target position

Let's create a function to determine the target point for a subject
(this took a piece of paper and some fiddling with the trigonometric functions)

```{r}
get_target.single <- function(x0, y0, ma_x0, ma_y0, mb_x0, mb_y0,
                              if_none = "original") {
  
  p0  = c(x0, y0)
  ma = c(ma_x0, ma_y0)
  mb = c(mb_x0, mb_y0)
  m_center = (ma + mb) / 2
  
  # calculate distance between mark a and b, and the angles between them
  theta_a  = atan2(ma_y0, ma_x0)
  theta_b  = atan2(mb_y0, mb_x0)
  theta_ab = theta_b - theta_a

  d_ab     = sqrt(sum((mb - ma) ^ 2))
  
  # target points are determined by taking a step from the middle between 
  # the marks
  stride   = (.5*d_ab) / tan(.5 * (1/6) * 2 * pi)

  # calculate candidate target points
  tx1 = m_center[1] + stride * cos(theta_a + theta_ab / 2)
  ty1 = m_center[2] + stride * sin(theta_a + theta_ab / 2)
  tx2 = m_center[1] - stride * cos(theta_a + theta_ab / 2)
  ty2 = m_center[2] - stride * sin(theta_a + theta_ab / 2)
  # ty1 = ma_y0 + radius * sin(angle_ab + 2*pi*60/360)
  # tx2 = ma_x0 + radius * cos(angle_ab - 2*pi*60/360)
  # ty2 = ma_y0 + radius * sin(angle_ab - 2*pi*60/360)

  # check if both target points are within the room
  t1_ok = abs(tx1) < room_dims[1] / 2 & abs(ty1) < room_dims[2] / 2
  t2_ok = abs(tx2) < room_dims[1] / 2 & abs(ty2) < room_dims[2] / 2
  
  # if 1 of both is ok, take that point
  if (sum(c(t1_ok, t2_ok)) == 1) {
    if (t1_ok) return(c(tx1, ty1))
    return(c(tx2, tx2))
  } 
  
  # if none are ok, return NA's or the original position
  if (sum(c(t1_ok, t2_ok)) == 0) {
    if (if_none == "original") return(c(x0,y0))
    return(c(NA,NA))
  }
  
  # if both are ok, take the one that's closest by, or the first one
  d1 = sqrt(sum((c(tx1, ty1) - p0) ^ 2))
  d2 = sqrt(sum((c(tx2, ty2) - p0) ^ 2))
  
  if (d1 <= d2) {
    return(c(tx1, ty1))
  }
  return(c(tx2, ty2))
}

get_target <- function(x0, y0, ma_x0, ma_y0, mb_x0, mb_y0) {
  df <- cbind(x0, y0, ma_x0, ma_y0, mb_x0, mb_y0)
  targets <- apply(df, 1, function(x) {
    get_target.single(x[[1]], x[[2]], x[[3]], x[[4]], x[[5]], x[[6]])
  })
  data.frame(t(targets))
}
  
df[, c("tx0", "ty0"):=get_target(x0, y0, ma_x0, ma_y0, mb_x0, mb_y0)]

```


Check correctness of solutions

```{r}
df[, t_correct:={
  da  = norm(matrix(c(tx0 - ma_x0, ty0 - ma_y0)), type = "F")
  db  = norm(matrix(c(tx0 - mb_x0, ty0 - mb_y0)), type = "F")
  dab = norm(matrix(c(mb_x0 - ma_x0, mb_y0 - ma_y0)), type = "F")
  max(dist(c(da, db, dab)) < 0.001)
}, by = "id"]
```

Visualize target positions for a few subjects

```{r}
samples = sample(1:npeople, size = 2)
circle0 + 
  geom_segment(data = df[samples], 
               aes(x = tx0, xend = ma_x0, y = ty0, yend = ma_y0)) +
  geom_segment(data = df[samples], 
               aes(x = tx0, xend = mb_x0, y = ty0, yend = mb_y0)) +
  geom_segment(data = df[samples], 
               aes(x = ma_x0, xend = mb_x0, y = ma_y0, yend = mb_y0)) +
  geom_segment(data = df[samples], 
               aes(x = x0, xend = tx0, y = y0, yend = ty0),
               arrow = arrow(), linetype = 2, alpha = .5)
```

### Take step to target position

Create a function to define a step to the target position

```{r}
take_step.single <- function(x, y, tx, ty, stride = .5) {
  direction <- c(tx, ty) - c(x, y)
  direction_normalized <- direction / sqrt(sum(direction^2))
  next_step = stride * direction_normalized
  new_position = c(x, y) + next_step
}
take_step <- function(x, y, tx, ty, stride = .5) {
  data = cbind(x, y, tx, ty)
  as.data.frame(t(apply(data, 1, function(row) 
    take_step.single(row[1], row[2], row[3], row[4], stride = stride))))
}
```


To track the positions for each participants, let's create a new 
data structure.

```{r, message=F, eval = F}
# library(tibbletime)
# nsteps = 100
# 
# positions <- create_series('00:00:00' ~ '00:05:00', 'second', class = "hms")

```


```{r}
nsteps = 100
stride = 0.5
positions <- array(dim = c(nsteps+1, npeople, 2), 
                   dimnames = list(paste0("step-",0:nsteps),
                                   paste0("id-", 1:npeople), 
                                   c("x", "y")))

positions[1,,1] <- df$x0
positions[1,,2] <- df$y0

## Set initial positions
df[, c("x", "y"):=list(x0, y0)]

for (i in 1:nsteps) {
  df[, c("ma_x", "ma_y"):= .SD[as.character(mark_a), list(x, y)]]
  df[, c("mb_x", "mb_y"):= .SD[as.character(mark_b), list(x, y)]]
  df[, c("tx", "ty")    := get_target(x, y, ma_x, ma_y, mb_x, mb_y)]
  df[, c("x", "y")      := take_step(x, y, tx, ty, stride = stride)]
  positions[i+1,,1] = df$x
  positions[i+1,,2] = df$y
}

```

Plot x-coordinate of a subject, to see if this process converges

```{r}
plot(positions[,1,1])
```


Let's try decreasing the step size after some time

```{r}
nsteps = 100
stride = 0.5
strides= ((nsteps - 1:nsteps) / nsteps) * stride
positions <- array(dim = c(nsteps+1, npeople, 2), 
                   dimnames = list(paste0("step-",0:nsteps),
                                   paste0("id-", 1:npeople), 
                                   c("x", "y")))

positions[1,,1] <- df$x0
positions[1,,2] <- df$y0

## Set initial positions
df[, c("x", "y"):=list(x0, y0)]

for (i in 1:nsteps) {
  stride = strides[i]
  df[, c("ma_x", "ma_y"):= .SD[as.character(mark_a), list(x, y)]]
  df[, c("mb_x", "mb_y"):= .SD[as.character(mark_b), list(x, y)]]
  df[, c("tx", "ty")    := get_target(x, y, ma_x, ma_y, mb_x, mb_y)]
  df[, c("x", "y")      := take_step(x, y, tx, ty, stride = stride)]
  positions[i+1,,1] = df$x
  positions[i+1,,2] = df$y
}

```

Plot x-coordinate of a subject, to see if this process converges, 
but now weighted for the stride size

```{r}
plot(positions[,1,1] / c(1,strides))
```


## Define how good a solution is

We can say that for each person, the cost is the sum of the differences 
between the distance between their marks and the distance of the subject 
to each mark.

$$cost = |d_{subject, mark_a} - d_{mark_a, mark_b}| + |d_{subject, mark_b} - d_{mark_a, mark_b}|$$

```{r}
cost_function <- function(x, y, ma_x, ma_y, mb_x, mb_y) {
  d_ab <- sqrt(sum((ma_x - mb_x)^2, (ma_y - mb_y)^2))
  d_a <- sqrt(sum((x - ma_x)^2, (y - ma_y)^2))
  d_b <- sqrt(sum((x - mb_x)^2, (y - mb_y)^2))
  abs(d_a - d_ab) + abs(d_b - d_ab)
}
```


Re run the simulation, keeping track of the cost (for each subject)

```{r}
nsteps = 100
stride = 0.5
strides= ((nsteps - 1:nsteps) / nsteps) * stride
positions <- array(dim = c(nsteps+1, npeople, 2), 
                   dimnames = list(paste0("step-",0:nsteps),
                                   paste0("id-", 1:npeople), 
                                   c("x", "y")))
costs <- array(dim = c(nsteps+1, npeople, 1))

positions[1,,1] <- df$x0
positions[1,,2] <- df$y0
costs[1,,1] <- df[, list(cost = cost_function(x0, y0, ma_x0, ma_y0, mb_x0, ma_y0)), by = "id"]$cost

## Set initial positions
df[, c("x", "y"):=list(x0, y0)]

for (i in 1:nsteps) {
  stride = strides[i]
  df[, c("ma_x", "ma_y"):= .SD[as.character(mark_a), list(x, y)]]
  df[, c("mb_x", "mb_y"):= .SD[as.character(mark_b), list(x, y)]]
  df[, c("tx", "ty")    := get_target(x, y, ma_x, ma_y, mb_x, mb_y)]
  df[, c("x", "y")      := take_step(x, y, tx, ty, stride = stride)]
  positions[i+1,,1] = df$x
  positions[i+1,,2] = df$y
  costs[i+1,,1] <- df[, list(cost = cost_function(x, y, ma_x, ma_y, mb_x, ma_y)), by = "id"]$cost
}

```

Gather costs in a convenient data.table format, 
and plot the costs per subject over time

```{r}
cost_df <- as.data.table(costs)
cost_df[, id:=factor(rep(1:npeople, each = nsteps + 1))]
cost_df[, step:=rep(0:nsteps, times = npeople)]

cost_df %>%
  ggplot(aes(x = step, y = costs, col = id)) + 
  geom_line()
```

As we see, the cost decreases, but it never saturates.

Let's now gather all the positions in a table, see if  we can visualize this nicely

```{r}
position_df <- as.data.table(positions)
position_df[, step:=rep(0:nsteps, times = npeople * 2)]
position_df[, coordinate:=rep(c("x", "y"), each = (nsteps + 1) * npeople)]
position_df[, id:=factor(rep(1:npeople, each = nsteps + 1, times = 2))]
position_df <- dcast.data.table(position_df, step + id ~ coordinate, value.var = "positions")
```

Animate using `gganimate` (available on github)

```{r}
library(gganimate)
p <- position_df %>%
  ggplot(aes(x = x, y = y, col = id, frame = step)) + 
  geom_point() + 
  lims(x = c(-1,1)*room_dims[1]/2, y = c(-1,1)*room_dims[2]/2) +
  coord_fixed() + 
  theme_minimal() +
  theme(legend.position = "none")

gganimate(p)
```


## Session information

<!-- Insert the session information into the document -->
```{r session-info}
```
