---
title: "Notes to Mathematical Statistics with Applications by J.E. Freund"
author: "Wouter van Amsterdam"
date: 2017-11-13
output: html_document
---

<!-- The file analysis/chunks.R contains chunks that define default settings
shared across the workflowr files. -->
```{r read-chunk, include=FALSE, cache=FALSE}
knitr::read_chunk("chunks.R")
```

<!-- Update knitr chunk options -->
```{r knitr-opts-chunk, include=FALSE}
```

<!-- Insert the date the file was last updated -->
```{r last-updated, echo=FALSE, results='asis'}
```

<!-- Insert the code version (Git commit SHA1) if Git repository exists and R
 package git2r is installed -->
```{r code-version, echo=FALSE, results='asis'}
```

<!-- Add your analysis here -->
## 2. Probability
### 2.2 Sample spaces
Sample space is collection of possible outcomes of an experiment.
For coinflip: $S = \left\{H, T\right\}$.

In general, take sample space that cannot be divided. 
$S = \left\{1,2,3,4,5,6\right\}$, better than $S = \left\{odd, even\right\}$.

### 2.3 Events
#### De Morgan Laws

* $(A\cap B)' = A'\cup B'$
* $(A\cup B)' = A'\cap B'$

### 2.4 Probability of an event

* Postulate 1: The probability of an event is a nonnegative real number;
that is, $P(A) \geq 0$ for any subset $A$ of $S$.
* Postulate 2: $P(S) = 1$
* Postulate 3: If $A_1,A_2,A_3,...$, is a finite or infinite sequence of 
mutually exclusive events of $S$, then 
$P(A_1 \cup A_2 \cup A_3 \cup ...)=P(A_1)+P(A_2)+P(A_3)+...$

### 2.5 Some rules of probability
#### General addition rule
If $A$ and $B$ are any two events in sample space $S$, then
$$P(A \cup B) = P(A) + P(B) - P(A \cap B)$$

If $A$, $B$ and $C$ are any three events in sample space $S$, then
$$P(A \cup B \cup C) = P(A) + P(B) + P(C) - P(A \cap B) - P(B \cap C) 
- P(B \cap C) + P(A \cap B \cap C)$$

### 2.6 Conditional probability
#### Conditional probability definition
If $A$ and $B$ are two events in sample space $S$ and $P(A) \neq 0$, then 
the **conditional probability** of $B$ given $A$ is:
$$P(B|A) = \frac{P(A \cap B)}{P(A)}$$

#### Definition of independence
Two events $A$ and $B$ are **independent** if and only if
$$P(A \cap B) = P(A)*P(B)$$

Which means that $P(B|A) = P(B)$ and $P(A|B) = P(A)$. 
These two conditions are equivalent when both $P(A) \neq 0$ and $P(B) \neq 0$.


* independent set: any intersection of k-members equals product of probabilities of members
* set can be pairwise independent, but not independent

### 2.8 Bayes' theorem
#### Rule of total probability / rule of elmination

If the events $B_1, B_2, ..., B_k$ constitute apartition of the sample space $S$,
with $P(B_i) \neq 0$ for any $i$, then for any event $A$ in $S$:
$$P(A) = \sum_{i=1}^k{P(B_i)*P(A|B_i)}$$

$B_i$ is a partition for $S$ when they are mutually exclusive, and their union 
equals $S$.

#### Bayes' theorem
Given a partition $B_i \neq 0$, $B_1, B_2, ..., B_k$ of $S$, 
and an event $A$ with $P(A) \neq 0$ in $S$, 
that was reached via one of the $k$ branches, 
then the probability that $A$ was reached through branch $r$ is:
$$P(B_r|A) = \frac{P(B_r)*P(A|B_r)}{\sum_{i=1}^k{P(B_i)*P(A|B_i)}}$$

$B_i$ are **prior probabilities**.

## Session information

<!-- Insert the session information into the document -->
```{r session-info}
```
