---
title: "Intro to generalized linear models"
author: "Wouter van Amsterdam"
date: 2018-03-05
output: html_document
---

<!-- The file analysis/chunks.R contains chunks that define default settings
shared across the workflowr files. -->
```{r read-chunk, include=FALSE, cache=FALSE}
knitr::read_chunk("chunks.R")
```

<!-- Update knitr chunk options -->
```{r knitr-opts-chunk, include=FALSE}
```

<!-- Insert the date the file was last updated -->
```{r last-updated, echo=FALSE, results='asis'}
```

<!-- Insert the code version (Git commit SHA1) if Git repository exists and R
 package git2r is installed -->
```{r code-version, echo=FALSE, results='asis'}
```

<!-- Add your analysis here -->

# Setup

Tutor: Rebecca Stellato

# GLM

1. response follows probability distribution from exponential family
2. linear predictor $\eta_i$
3. link function describing mean outcome as function of linear predictor
$g(\mu_i)=\eta_i$

## Parameter estimation

In general: maximum likelihood estimation

Linear model: ordinary least squares

## Assessing fit

Calculate likelihood of observed data; using $p_i = \frac{y_i}{n_i}$

Residual deviance $-2 * (l_C - l_S)$

Null deviance: $-2 * (l_0 - l_C)$; with no predictors, just intercept

Deviance is a likelihood ratio-test, which is approximately chi-squared 
distributed with df = difference in df of models

When residual deviance < degrees of freedom, no evidence of lack of fit.

Not the same as: this is the correct/best model

### Residuals

Problem: variance not independent of $y$

Pearson residuals: $r_p = \frac{y-\bar{\mu}}{\sqrt{var(\bar{\mu})}}$

Square and add gives Pearson's Chi-squared

Deviance residuals: residuals defined such that the squared residuals sum up to deviance

also: response residuals, working residuals, jackknife residuals

You can plot residuals against a non-included variable, if they show structure,
 it should probably be included in the model.
 
- jackknife residuals: leave one out, refit the model, evaluate deviance
- leverage: influence in X diraction
- cook's distance: combine delta-beta (
leave on out, how much do the parameters change (combined))

Faraway: 

- half-normal plots (take absolute values of the different residuals),
plot against half-normal distribution; straight line not necessary, but 
should be on a curve

## Session information

<!-- Insert the session information into the document -->
```{r session-info}
```
