---
title: "Presentation Jointbleed"
author: "Aernoud Fiolet, Koos van Kesteren, Wouter van Amsterdam"
date: 2018-03-08
output: html_document
---

<!-- The file analysis/chunks.R contains chunks that define default settings
shared across the workflowr files. -->
```{r read-chunk, include=FALSE, cache=FALSE}
knitr::read_chunk("chunks.R")
```

<!-- Update knitr chunk options -->
```{r knitr-opts-chunk, include=FALSE}
```

<!-- Insert the date the file was last updated -->
```{r last-updated, echo=FALSE, results='asis'}
```

<!-- Insert the code version (Git commit SHA1) if Git repository exists and R
 package git2r is installed -->
```{r code-version, echo=FALSE, results='asis'}
```

<!-- Add your analysis here -->

# Setup R


```{r}
library(dplyr)
library(data.table)
library(magrittr)
library(purrr)
library(here) # for tracking working directory
library(ggplot2); theme_set(theme_minimal())
library(epistats)
library(broom)
```


Get data

```{r}
joint <- read.table(here("data", "jointbleed.txt"), sep = ",", header = T)
str(joint)
```

Table

```{r}
table(joint$jb)
```


Marginal outcome distribution

```{r}
joint %>%
  ggplot(aes(x = jb)) + 
  geom_histogram() + 
  theme_minimal()
```

Very high count of zeros, zero inflated poisson

## Curation

Check for outliers

```{r}
require(tidyr)

joint %>%
  gather(key = "variable", value = "value") %>%
  ggplot(aes(x = value)) + 
  geom_histogram() +
  facet_wrap(~variable)
```

Age has an outlier

```{r}
joint %>%
  arrange(-age) %>% filter(row_number() < 10)
```

Remove this case, this must be an error. Highly probable that this is random

```{r}
joint %<>% filter(age < 120)
```

### Look at pairwise marginal distributions

```{r}
joint %>%
  ggplot(aes(x = age, y = jb)) + 
  geom_point()
```

```{r}
joint %>%
  ggplot(aes(x = age, y = FVIII)) + 
  geom_point()
```


## Modeling

Based on theoretical considerations, these models will be evaluated:

- linear regression (with log-transform of response)
- poisson (due to nature of outcome)
- quasipoisson (to account for possible overdispersion)
- zero-inflated poisson (based on the outcome distribution)
- negative binomial

Liklihood methods:

- linear regression: ordinary least squares
- poisson: maximum likelihood
- quasipoisson: quasi-likelihood
- zero-inflated poisson: maximum likelihood
- negative binomial: maximum likelihood

This means we can directly compare poisson, ZIP and negative binomial,
but not quasi-poisson. 


Strategy:

For each link function and error distribution

- fit model with age and FVIII and interaction
- reduce model using likelihood ratio test
- save final model

### Variable transformations

Add a log + 0.5 transformed jb to account for zeros
Also log-transform FVIII as it is a ratio

```{r}
joint %<>% mutate(
  jb_log = log(jb + 0.5),
  FVIII_log = log(FVIII)
  )
```


```{r}
joint %>%
  select(-age, -severity) %>%
  gather(FVIII, FVIII_log, key = "variable", value = "xvalue") %>%
  gather(jb, jb_log, key = "outcome", value = "yvalue") %>%
  ggplot(aes(x = xvalue, y = yvalue)) +
  geom_point() + geom_smooth() +
  facet_grid(outcome~variable, scales = "free") + 
  theme_minimal()

```


Remove zeros


```{r}
joint %>%
  filter(jb > 0) %>%
  ggplot(aes(x = FVIII, y = jb)) +
  geom_point()
```


### poisson

```{r}
fit_pois <- glm(jb ~ age * FVIII, data = joint, family = poisson)
drop1(fit_pois, test = "Chisq")
fit_pois <- glm(jb ~ age + FVIII, data = joint, family = poisson)
drop1(fit_pois, test = "Chisq")
fit_pois <- glm(jb ~ FVIII, data = joint, family = poisson)
drop1(fit_pois, test = "Chisq")
dispersion(fit_pois)
```

### lm

Create a likelihood function for the linear model, this makes the models 
directly comparable




```{r}
fit_lm <- glm(jb_log ~ age * FVIII, data = joint)
drop1(fit_lm, test = "Chisq")
fit_lm <- glm(jb_log ~ age + FVIII, data = joint)
drop1(fit_lm, test = "Chisq")
fit_lm <- glm(jb_log ~ FVIII, data = joint)
drop1(fit_lm, test = "Chisq")
summary(glm(jb_log ~ FVIII, data = joint))
```

## quisipoisson

```{r}
fit_qpois <- glm(jb ~ age * FVIII, data = joint, family = quasipoisson)
drop1(fit_qpois, test = "Chisq")
fit_qpois <- glm(jb ~ age + FVIII, data = joint, family = quasipoisson)
drop1(fit_qpois, test = "Chisq")
fit_qpois <- glm(jb ~ FVIII, data = joint, family = quasipoisson)
drop1(fit_qpois, test = "Chisq")
```

## zero-inflated poisson

```{r}
require(pscl)
fit_zip <- zeroinfl(jb ~ age * FVIII, data = joint)
AIC(fit_zip)
fit_zip <- zeroinfl(jb ~ age * FVIII, data = joint, dist = "negbin")
AIC(fit_zip)
# drop1(fit_zip, test = "Chisq")
fit_zip <- zeroinfl(jb ~ age * FVIII | FVIII, data = joint)
AIC(fit_zip)
# drop1(fit_zip, test = "Chisq")
```

No further reduction possible

## negative binomial 

```{r}
require(MASS); select <- dplyr::select
fit_nb <- glm.nb(jb ~ age * FVIII, data = joint)
drop1(fit_nb, test = "Chisq")
fit_nb <- glm.nb(jb ~ age + FVIII, data = joint)
drop1(fit_nb, test = "Chisq")
fit_nb <- glm.nb(jb ~ FVIII, data = joint)
drop1(fit_nb, test = "Chisq")
```

Check non-linear terms

Neither splines or polynomial terms improved AIC

```{r}
spline_dfs <- as.list(1:5)
spline_dfs %>%
  map_dbl(function(df) glm.nb(jb ~ splines::ns(FVIII, df = df), data = joint) %>%
        AIC)
spline_dfs %>%
  map_dbl(function(df) glm.nb(jb ~ poly(FVIII, degree = df), data = joint) %>%
        AIC)
```

```{r}
glm.nb(jb ~ splines::ns(FVIII, df = 2), data = joint)
glm.nb(jb ~ FVIII, data = joint)
```

### Zero-inflated negative binomial

```{r}
fit_nbzip <- zeroinfl(jb ~ age * FVIII, data = joint, dist = "negbin")
drop1(fit_nbzip, test = "Chisq")
fit_nbzip <- zeroinfl(jb ~ age + FVIII, data = joint, dist = "negbin")
drop1(fit_nbzip, test = "Chisq")
```

First check covariates for zero distribution

```{r}
fit_nbzip <- zeroinfl(jb ~ age * FVIII | age * FVIII, data = joint, dist = "negbin")
AIC(fit_nbzip)
fit_nbzip <- zeroinfl(jb ~ age * FVIII | age + FVIII, data = joint, dist = "negbin")
AIC(fit_nbzip)
fit_nbzip <- zeroinfl(jb ~ age * FVIII | FVIII, data = joint, dist = "negbin")
AIC(fit_nbzip)
fit_nbzip <- zeroinfl(jb ~ age * FVIII | age, data = joint, dist = "negbin")
AIC(fit_nbzip)
fit_nbzip <- zeroinfl(jb ~ age * FVIII | 1, data = joint, dist = "negbin")
AIC(fit_nbzip)

```

Check part for count distribution

```{r}
fit_nbzip <- zeroinfl(jb ~ age * FVIII | age + FVIII, data = joint, dist = "negbin")
AIC(fit_nbzip)
fit_nbzip <- zeroinfl(jb ~ age + FVIII | age + FVIII, data = joint, dist = "negbin")
AIC(fit_nbzip)
fit_nbzip <- zeroinfl(jb ~ age | age + FVIII, data = joint, dist = "negbin")
AIC(fit_nbzip)
fit_nbzip <- zeroinfl(jb ~ 1 | age + FVIII, data = joint, dist = "negbin")
AIC(fit_nbzip)
fit_nbzip <- zeroinfl(jb ~ FVIII | age + FVIII, data = joint, dist = "negbin")
AIC(fit_nbzip)
```


```{r}
summary(fit_nbzip)
```

```{r}
cov_grid <- expand.grid(
  FVIII = seq(0, 20, length.out = 100),
  age = seq(0, 100, length.out = 100)
)
grid_predictions <- predict(fit_nbzip, newdata = cov_grid)
preds <- data.frame(cov_grid, pred_count = grid_predictions)
```


```{r}
preds %>%
  ggplot(aes(x = age, y= FVIII, fill = pred_count)) + 
  geom_tile()
```


### Compare models

```{r}
fits <- list(
  poisson = fit_pois,
  # lm = fit_lm,
  # quasipoisson = fit_qpois,
  zip = fit_zip,
  negative_binomial = fit_nb,
  zip_negative_binomial = fit_nbzip
)
```

Compare with AIC


```{r}
fits %>%
  map_df(function(fit) data.frame(
    deviance = ifelse(is.null(deviance(fit)), NA, deviance(fit)),
    loglik = logLik(fit),
    AIC = AIC(fit)
    ), .id = "model")
```



```{r}
fits %>%
  map_df(function(fit) data.frame(
    deviance = ifelse(is.null(deviance(fit)), NA, deviance(fit)),
    loglik = logLik(fit),
    AIC = AIC(fit)
    ), .id = "model")
```


## Model diagnostics

```{r}
plot(fit_nb)
```

All observations fall within a Cook's distance of < 0.5


## Discussion

Tried log-transform of FVIII

## Session information

<!-- Insert the session information into the document -->
```{r session-info}
```
