---
title: "Missing data notes"
author: "Wouter van Amsterdam"
date: 2018-01-19
output: html_document
---

<!-- The file analysis/chunks.R contains chunks that define default settings
shared across the workflowr files. -->
```{r read-chunk, include=FALSE, cache=FALSE}
knitr::read_chunk("chunks.R")
```

<!-- Update knitr chunk options -->
```{r knitr-opts-chunk, include=FALSE}
```

<!-- Insert the date the file was last updated -->
```{r last-updated, echo=FALSE, results='asis'}
```

<!-- Insert the code version (Git commit SHA1) if Git repository exists and R
 package git2r is installed -->
```{r code-version, echo=FALSE, results='asis'}
```

<!-- Add your analysis here -->

## Intro

Tutor: Thomas Debray

## Multiple missing variables

Initialize imputed dataset with some arbitrary mechanism (zero, mean, random, model based on complete cases...)

Updated with Gibbs sampler

Or: do jomo (joint imputation), assume multivariate normal distribution

## Congeniality

assumptions in the analysis model should also be present in the imputation model

potential sources of uncongeniality

- adjustment for confounders
- nonlinearity terms
- covariate interactions
- between-study heterogeneity (e.g. in IPD-MA)
---> ignoring clustering with imputation will reduce heterogeneity between studies
- transformed versions of a variable (e.g. BMI)

### strategies to remove congeniality

- generate new variables for interactions, as 'just another variable' in the imputation model (not perfect but works ok; could be that imputed interaction is not X1*X2, works but doesn't make a lot of sense)
- passive imputation: impute original data, and calculate the interaction
- flexible model: imputing variables with non-linear trends (e.g. broken-stick)

## Imputation with time-to-event

Don't just include event indicator and time (imputation model assumes linear association with time)

better (for Cox):

- include D, and the cumulative baseline hazard
- include D and the Nelson-Aalen estimator

for accelerated failure time models: think about interactions between covariates and time

## Individual patient missing data

- Partially missing data missing data missing datamissing data
- Systematically missing data missing data missing datamissing data
- Entire studie(s) missing studie(s) missing studie(s) missing studie(s) missing
- Combination of above aboveissues issues

-> imputation models should account for clustering

### strategies

imputation strategy
- within-study imputation
- stratified imputation (stack all data, include study ID as dummy variable)
- hierarchical imputation (e.g. mixed effects models; for example allow for a different intercept of the imputation models in each study)

imputation paradigm
- fully conditional specification (FCS)
- joint modeling (JM)

fully condiational specification
- FCS-2stage (resche-rigon & white)
- FCS-GLM (Debray, binary, continous and count); works when clusters are small; computationally intensive

joint modeling
- JM-JOMO (quartagno); (bias when few people and/or few studies; conservative w.r.t. heterogeneity)

for IPD-MA with clustered data, use each measurement*timepoint as different variable (only works when time-points are fixed between patients)

## Predictive mean matching

For semi-continous data, 
- e.g. mixture (packyears, or count of bacteries)
- seasonal data

predictive mean matching ('nearest neighbor' matching)

- never generates values that are impossible;
- take best matches
- add noise

stable performance
- free of distributional assumptoins
- accurate inferences in extreme conditions

less powerfull wehn
- missing outlying values in very skewed data
- small datasets
- multivariate missingness


mahalanobis;





## Session information

<!-- Insert the session information into the document -->
```{r session-info}
```
