---
title: "Intro to multi-level models"
author: "Wouter van Amsterdam"
date: 2018-04-16
output: html_document
---

<!-- The file analysis/chunks.R contains chunks that define default settings
shared across the workflowr files. -->
```{r read-chunk, include=FALSE, cache=FALSE}
knitr::read_chunk("chunks.R")
```

<!-- Update knitr chunk options -->
```{r knitr-opts-chunk, include=FALSE}
```

<!-- Insert the date the file was last updated -->
```{r last-updated, echo=FALSE, results='asis'}
```

<!-- Insert the code version (Git commit SHA1) if Git repository exists and R
 package git2r is installed -->
```{r code-version, echo=FALSE, results='asis'}
```

<!-- Add your analysis here -->

## Setup R-session


```{r}
library(dplyr)
library(data.table)
library(magrittr)
library(purrr)
library(here) # for tracking working directory
library(ggplot2)
library(epistats)
library(broom)
```

```{r}
scl <- read.table(here("data", "school.dat"), header = T)
str(scl)
```


Possible approach

- aggregate by school, regress on mean
- do linear regression on all participants, ignoring clustering

All data

```{r}
scl %>%
  ggplot(aes(y = normexam, x = standlrt)) + 
  geom_point()
```


Aggregate by school, take means

```{r}
scl %>%
  group_by(school) %>%
  summarize_at(vars(normexam, standlrt), funs(mean)) %>% 
  ggplot(aes(y = normexam, x = standlrt)) + 
  geom_point()
```


Aggregating creates possibility of "ecological fallacy"

applying group level observations to individual level


### Do regression by school

```{r}
require(tidyr)
scl_nested <- scl %>% group_by(school) %>% nest()
```



### Problems with groups as interactions

- many degrees of freedom taken
- cannot generalize to other schools
- sample size inflation
- no single answer to coefficients

## Mixed model

Fixed effects: usually what you are interested in
Random: only include a few, makes model much more complicated; the included groups are random samples from all groups

Random effects for intercept and slope take up only 3 degrees of freedom
Taken from normal distribution with mean 0 

- variance of intercept, slope and covariance of these 2


## Session information

<!-- Insert the session information into the document -->
```{r session-info}
```
