---
title: "Mixed model testing"
author: "Wouter van Amsterdam"
date: 2018-04-18
output: html_document
---

<!-- The file analysis/chunks.R contains chunks that define default settings
shared across the workflowr files. -->
```{r read-chunk, include=FALSE, cache=FALSE}
knitr::read_chunk("chunks.R")
```

<!-- Update knitr chunk options -->
```{r knitr-opts-chunk, include=FALSE}
```

<!-- Insert the date the file was last updated -->
```{r last-updated, echo=FALSE, results='asis'}
```

<!-- Insert the code version (Git commit SHA1) if Git repository exists and R
 package git2r is installed -->
```{r code-version, echo=FALSE, results='asis'}
```

<!-- Add your analysis here -->

# Intro

Tutor: Rebecca Stellato

## ML vs REML

ML underestimates variances of parameters.

REML is better. 

Testing models that differ in correlation part: both ML or REML
Testing models that differ in fixed part: only ML is OK

for reporting, do REML

## Procedure

- start with all possible covariates in fixed model
- experiment with random part to fix correlation structure
- then reduce fixed part

### Testing random effects

variance parameters are strictly >= 0; divide p-value by 2 for LRT chi-square
AIC is ok, but don't use +2 as criterion (is not often used in literature)

### Testing fixed effects

use LRT (under ML)


Mixed models work pretty well with missing data on outcome
Uses some implicit imputation for missing values.

For missing predictors, imputation will help

## Model assumptions

- linearity of effects (for those you include)
- normality of residuals
- normality of intercepts and slopes (however, these are more or less 
coerced to normal distribution by the random modeling)

## Centering covariates

Center 'time' axis will change fixed intercept (when there is a fixed effect of time),
and SD of random intercept effect (if variation increases with time)
and correlation of intercept and slope of random parts

## Multilevel models

e.g. longitudinal data from patients of different hospitals

Usually, you want a random intercept on hospital level, due to design,
even if it's not needed by LRT


-> great range of possibilities, think about what makes sense!


Time-varying covariates no problem for mixed models. 
Gives a single fixed effect.

--> think about case-mix and what to include in hospital level covariates


## Session information

<!-- Insert the session information into the document -->
```{r session-info}
```
