---
title: "Presentation lung cancer survival prediction"
author: "Wouter van Amsterdam"
date: 2018-04-20
output: word_document
---

<!-- The file analysis/chunks.R contains chunks that define default settings
shared across the workflowr files. -->
```{r read-chunk, include=FALSE, cache=FALSE}
knitr::read_chunk("chunks.R")
```

<!-- Update knitr chunk options -->
```{r knitr-opts-chunk, include=FALSE}
```

<!-- Insert the date the file was last updated -->
```{r last-updated, echo=FALSE, results='asis'}
```

<!-- Insert the code version (Git commit SHA1) if Git repository exists and R
 package git2r is installed -->
```{r code-version, echo=FALSE, results='asis'}
```

<!-- Add your analysis here -->


## Setup R environment


```{r}
library(dplyr)
library(data.table)
library(magrittr)
library(purrr)
library(here) # for tracking working directory
library(ggplot2)
library(epistats)
library(broom)
library(survival)
```

## Get Data

```{r}
lc <- read.csv(here("data", "lungca.csv"))

cont_vars <- c("age", "Karn.phys", "Karn.pt", "wtloss")

lc %<>% 
  mutate(status = status - 1,
         sex = factor(sex, levels = c("1", "2"), labels = c("male", "female")),
         wtloss = 0.453592 * wtloss,
         ECOG.phys = factor(ifelse(ECOG.phys == "3", "2", ECOG.phys)),
         ECOG.phys = factor(ECOG.phys, levels = c("0", "1", "2"), 
                            labels = c("0", "1", "2-3")))
         # age = age - mean(age))


str(lc)
```

```{r}
summary(lc)
nna(lc)
```

```{r}
table(lc$ECOG.phys, lc$sex)
```


```{r}
lc %>%
  data.table::melt(measure.vars = cont_vars) %>% 
  ggplot(aes(x = value)) + 
  geom_histogram() + 
  facet_grid(sex~variable, scales = "free_x") + 
  theme_minimal()
```

Recode Karnophsky as factors

```{r}
factor_vars <- c("center", "sex", "ECOG.phys", "Karn.phys", "Karn.pt")
lc %<>%
  mutate_at(vars(factor_vars), funs(as.factor))

```



```{r}
lc %>%
  ggplot(aes(x = calories, y = wtloss)) + 
  geom_point() + theme_minimal()
```

```{r}
lc %>%
  mutate(
    wt_loss_ratio = wtloss / calories
  ) %>%
  ggplot(aes(x = wt_loss_ratio)) + 
  geom_histogram()
```

### Overview of survival

```{r}
require(ggfortify)
surv_plot <- lc %>%
  survfit(Surv(time / 365, status) ~ 1, data = .) %>%
  autoplot() + theme_minimal() + 
  labs(x = "Time (years)",
       y = "Survival probability") + 
  ggtitle("Overview of survival (Kaplan-Meier estimate)")
surv_plot
```

```{r}
survfit1 <- survfit(Surv(time, status) ~ 1, data = lc)
print(survfit1)
```


```{r}
require(ggfortify)
lc %>%
  mutate(dummy_variable = T,
         status = factor(status, levels = c(0, 1), labels = c("censored", "deceased"))) %>%
  survfit(Surv(time / 365, dummy_variable) ~ status, data = .) %>%
  autoplot() + theme_minimal()
```


```{r}
lc %>%
  coxph(Surv(time, status) ~ age + sex + ECOG.phys + Karn.phys + Karn.pt + calories + wtloss, 
        data = .) %>%
  summary()
```

Model with random intercept per center

Random slopes make no sense -> why different effect of age in different centers?

```{r}
fit1 <- coxph(Surv(time, status) ~ age + sex + ECOG.phys + Karn.phys + Karn.pt + calories * wtloss + frailty(center, dist = "gauss"),
        data = lc)
# summary(fit)

fit2 <- coxph(Surv(time, status) ~ age + sex + ECOG.phys + Karn.phys + Karn.pt + calories * wtloss,
        data = lc)

fits <- list(with_random_full = fit1, no_random_full = fit2)

map_df(fits, extractAIC)

extractAIC(fit1)
extractAIC(fit2)

anova(fit1, fit2, test = "Chisq")
```

With manual likelihood-ratio test for chi-square distribution with 1 degree 
of freedom

```{r}
pchisq(2*(logLik(fit1) - logLik(fit2)), df = 1, lower.tail = F)
```


Follow LRT the model with random intercept per center is better. 

Use this random part, now reduce fixed part.

```{r}
fit_full <- coxph(Surv(time, status) ~ age + sex + ECOG.phys + Karn.phys + Karn.pt + calories * wtloss + frailty(center, dist = "gauss"), data = lc)
                  
```

```{r}
fit <- fit_full
drop1(fit, test = "Chisq")
```

```{r}
fit <- coxph(Surv(time, status) ~ sex + ECOG.phys + Karn.phys + Karn.pt + calories * wtloss + frailty(center, dist = "gauss"),
        data = lc)
drop1(fit, test = "Chisq")
```

```{r}
fit <- coxph(Surv(time, status) ~ sex + ECOG.phys + Karn.phys + calories * wtloss + frailty(center, dist = "gauss"),
        data = lc)
drop1(fit, test = "Chisq")
```


```{r}
fit <- coxph(Surv(time, status) ~ sex + ECOG.phys + Karn.phys + calories + wtloss + frailty(center, dist = "gauss"),
        data = lc)
drop1(fit, test = "Chisq")
```

```{r}
fit <- coxph(Surv(time, status) ~ sex + ECOG.phys + Karn.phys + wtloss + frailty(center, dist = "gauss"),
        data = lc)
drop1(fit, test = "Chisq")
```

```{r}
fit <- coxph(Surv(time, status) ~ sex + ECOG.phys + wtloss + frailty(center, dist = "gauss"),
        data = lc)
drop1(fit, test = "Chisq")
```

```{r}
fit <- coxph(Surv(time, status) ~ sex + ECOG.phys + frailty(center, dist = "gauss"),
        data = lc)
drop1(fit, test = "Chisq")
```

```{r}
fit_final <- coxph(Surv(time, status) ~ sex + ECOG.phys + frailty(center, dist = "gauss"), data = lc)
summary(fit_final)
```

```{r}
knitr::kable(extract_RR(fit_final))
```

```{r}
cox.zph(fit_final)
```

```{r}
lc %<>% mutate(lp = fit_final$linear.predictors)
lc %>%
  mutate(lp_quantile = quant(lp, n.tiles = 3, label = "risk group ")) %>%
  survfit(Surv(time / 365, status) ~ lp_quantile, data = .) %>%
  autoplot() + theme_minimal() +
  labs(x = "Time (years)",
       y = "Survival probability") + 
  ggtitle("Overview of survival (Kaplan-Meier estimate)", "strata by terciles of linear predictor")
```

```{r}
lc %>%
  survfit(Surv(time / 365, status) ~ sex, data = .) %>%
  autoplot() + theme_minimal() +
  labs(x = "Time (years)",
       y = "Survival probability") + 
  ggtitle("Overview of survival (Kaplan-Meier estimate)", "strata by sex")
```

```{r}
lc %>%
  survfit(Surv(time / 365, status) ~ ECOG.phys, data = .) %>% 
  autoplot() + theme_minimal() +
  labs(x = "Time (years)",
       y = "Survival probability") +
  ggtitle("Overview of survival (Kaplan-Meier estimate)",
          "strata by ECOG performance score")

```


```{r}
fit_sex <- coxph(Surv(time, status)~ strata(sex) + ECOG.phys + frailty(center, "gauss"), data = lc)
plot(survfit(fit_sex), fun = "cloglog",
     main = "c-loglog plot for final model stratified by sex",
     col = c("blue", "red"))
legend("topleft", c("male", "female"), col = c("blue", "red"), lty = 1)
fit_ecog <- coxph(Surv(time, status)~ sex + strata(ECOG.phys) + frailty(center, "gauss"), data = lc)
plot(survfit(fit_ecog), fun = "cloglog", col = c("green", "blue", "red"),
     main = "c-loglog plot for final model stratified by ECOG performance status")
legend("topleft", legend = c("0", "1", "2-3"), lty = 1, col = c("green", "blue", "red"))
```


## Session information

<!-- Insert the session information into the document -->
```{r session-info}
```
