---
title: "Machine learning challange"
output: html_notebook
---



Setup R

```{r}
library(here)
library(dplyr)
library(ggplot2)
library(magrittr)
library(data.table)
library(purrr)
library(epistats)
```

Read in data

```{r}
aml <- read.csv(here("data", "AMLtraindata.csv"))
train_clinical <- readr::read_rds(here("data", "aml_train_clinical.rds"))
train_clinical_imputed <- readr::read_rds(here("data", "aml_train_clinical_imputed.rds"))
train_proteomic <- readr::read_rds(here("data", "aml_train_proteomic.rds"))

```

```{r}
table(aml$Chemo.Simplest, aml$resp.simple)
prop.table(table(aml$Chemo.Simplest, aml$resp.simple), margin = 1)
fisher.test(table(aml$Chemo.Simplest, aml$resp.simple))
```

### Descriptive analysis

```{r}
require(tableone)
t1_vars <- c("SEX", "Age.at.Dx", "AHD", "PRIOR.MAL", "PRIOR.CHEMO", 
"PRIOR.XRT", "Infection", 
"Chemo.Simplest", "resp.simple")
t1 <- tableone::CreateTableOne(vars = t1_vars, data = aml)
print(t1, quote = F, noSpaces = T)
```

Some distribution checks

```{r}
ks_pvals <- train_proteomic %>%
  map_dbl(~ks.test(.x, y = "pnorm")$p.value)
hist(ks_pvals)
```


## Model

Recode prior malignancy

```{r}
train_clinical_imputed %<>%
  mutate(prior_malignancy = PRIOR.CHEMO + PRIOR.XRT >= 1)
```


Define x and y
y = 1 for treatment resistance


```{r}
y <- as.numeric(aml$resp.simple) - 1
y_fact <- epistats::logical2factor(as.logical(y))
X_known_pf <- as.matrix(dplyr::select(train_clinical_imputed,
                                         Age.at.Dx,
                                         prior_malignancy))
X_clin <- as.matrix(dplyr::select(train_clinical_imputed, -resp.simple, - Chemo.Simplest) %>%
                 mutate_all(as.numeric))
X_prot <- as.matrix(train_proteomic)
prot_means <- colMeans(train_proteomic)
prot_sds <- map_dbl(train_proteomic, sd)
X_prot <- scale(X_prot, center = T, scale =  T)
X <- cbind(X_clin, X_prot)
```

### Logistic regression

Combine 

```{r}
require(rms)
# fit_lr <- glm(resp.simple ~ Age.at.Dx + prior_malignancy, data = train_clinical_imputed,
#                family = "binomial")
# fit_lr2 <- glm(resp.simple ~ poly(Age.at.Dx, 2) + prior_malignancy, data = train_clinical_imputed,
#                family = "binomial")
fit_lr <- lrm(resp.simple ~ Age.at.Dx + prior_malignancy, data = train_clinical_imputed,
              x = T, y = T)
fit_lr2 <-lrm(resp.simple ~ poly(Age.at.Dx, 2) + prior_malignancy, 
              data = train_clinical_imputed,
              x = T, y = T)
# summary(fit_lr2)
# anova(fit_lr2, fit_lr, test = "Chisq")
```

Validate model

```{r}
library(rms)
validate(fit_lr)
# abs(.3356) / 2 + .5
val.prob(logit = fit_lr$linear.predictors, y = as.numeric(fit_lr$y))
```


### SVM

```{r}
library(caret)
ctrl <- trainControl(method = "cv", number = 10, savePredictions = T, classProbs = T)#,
                     # summaryFunction = twoClassSummary)

tune_grid_lin <- data.frame(C = c(.25, .5, 0.75, 0.9, 1, 1.1, 1.25, 1.5))
tune_grid <- expand.grid(sigma = c(.005, .01, .015, 0.2),
                    C = c(.25, .5, 0.75, 0.9, 1, 1.1, 1.25, 1.5)
)

fit <- train(y = y_fact, x = X_clin,
             trControl = ctrl,
                     # tuneGrid = tune_grid_lin,
                     # tuneGrid = tune_grid,
             metric = "ROC",
             # weights = c(1, 1.54),
             # maximize = "logLoss",
             method = "rf")
fit 
confusionMatrix(fit)
```


```{r}
nfolds_out = 5
nreps_out = 10
nruns = nfolds_out * nreps_out

train_folds <- createMultiFolds(train_clinical_imputed$resp.simple,
                          k = nfolds_out, times = nreps_out)

fits  <- vector(mode = "list", length = nruns)
preds <- vector(mode = "list", length = nruns)
accs  <- vector(mode = "numeric", length = nruns)

train_X <- as.matrix(dplyr::select(train_clinical_imputed, -resp.simple, -cyto.cat, -Chemo.Simplest, -prior_malignancy))
train_y <- factor(train_clinical_imputed$resp.simple, labels = c("CR", "RESISTANT"))

for (j in 1:nruns) {
  cat("run ", j, " out of ", nruns, "\n")
  ftrain_X <- train_X[train_folds[[j]],]
  ftrain_y <- train_y[train_folds[[j]]] 
  ftest_X <- train_X[-train_folds[[j]],]
  ftest_y <- train_y[-train_folds[[j]]] 
  
  upsamp_dat <- upSample(ftrain_X, ftrain_y)
  levels(upsamp_dat$Class) <- c("CR", "RESISTANT")
  
  fit <- train(Class ~ .,
               data = upsamp_dat,
               trControl = ctrl,
               method = "rf")
  fits[[j]] <- fit
  
  pred <- predict(fit, ftest_X)
  preds[[j]] <- pred
  
  acc <- mean(pred == ftest_y)
  accs[[j]] <- acc
  
}

```

```{r}
nfolds_out = 5
nreps_out = 10
nruns = nfolds_out * nreps_out

train_folds <- createMultiFolds(train_clinical_imputed$resp.simple,
                          k = nfolds_out, times = nreps_out)

fits_prot  <- vector(mode = "list", length = nruns)
preds_prot <- vector(mode = "list", length = nruns)
accs_prot  <- vector(mode = "numeric", length = nruns)

train_X_prot <- as.matrix(cbind(dplyr::select(train_clinical_imputed, -resp.simple, -cyto.cat, -Chemo.Simplest, -prior_malignancy), train_proteomic))
train_y <- factor(train_clinical_imputed$resp.simple, labels = c("CR", "RESISTANT"))

for (j in 1:nruns) {
  cat("run ", j, " out of ", nruns, "\n")
  ftrain_X_prot <- train_X_prot[train_folds[[j]],]
  ftrain_y <- train_y[train_folds[[j]]] 
  ftest_X_prot <- train_X_prot[-train_folds[[j]],]
  ftest_y <- train_y[-train_folds[[j]]] 
  
  upsamp_dat <- upSample(ftrain_X_prot, ftrain_y)
  levels(upsamp_dat$Class) <- c("CR", "RESISTANT")
  
  fit <- train(Class ~ .,
               data = upsamp_dat,
               trControl = ctrl,
               method = "rf")
  fits_prot[[j]] <- fit
  
  pred <- predict(fit, ftest_X_prot)
  preds_prot[[j]] <- pred
  
  acc <- mean(pred == ftest_y)
  accs_prot[[j]] <- acc
  
}

```

```{r}
var_imps_prot <- map(fits_prot, varImp) %>% map("importance")
var_imps_prot <- do.call(cbind, var_imps_prot)
rowMeans(var_imps_prot) %>% sort(decreasing = T)
```

Get the variable importance form all

```{r}
var_imps <- map(fits, varImp) %>% map("importance")
var_imps <- do.call(cbind, var_imps)
rowMeans(var_imps) %>% sort(decreasing = T)
```

Do logistic regression on 3 most important predictors

```{r}
fit_lr3 <- lrm(resp.simple ~ Age.at.Dx + CD34 + PLT, data = train_clinical_imputed,
               x = T, y = T)
fit_lr3_glm <- glm(resp.simple ~ Age.at.Dx + CD34 + PLT, data = train_clinical_imputed,
               family = "binomial")
fit_lr3_val <- validate(fit_lr3, B = 100)
fit_lr3_val[1, 5] / 2 + .5
val.prob(logit = fit_lr3$linear.predictors, y = as.numeric(fit_lr3$y))
fit_null <- lrm(resp.simple ~ 1, data = train_clinical_imputed)
lrtest(fit_null, fit_lr3)
lrtest(fit_lr, fit_lr3)
```


```{r}
library(pROC)
roc_obj <- roc(response = fit_lr3$y, predictor = fit_lr3$linear.predictors)
plot((roc_obj), print.auc = T)
```


Use shrinkage based on bootstrapping

```{r}

## Use the optimism-corrected slope as the shrinkage factor
shrinkage.factor <- fit_lr3_val["Slope","index.corrected"]

## Show results
res.coef <- data.frame(Original = coef(fit_lr3),
                       Shrunk.boot = c(coef(fit_lr3)[1], coef(fit_lr3)[-1] * shrinkage.factor))
round(res.coef, 3)
```

## Predict on new data

```{r}
library(mice)
betas <- res.coef$Shrunk.boot
pvars <- rownames(res.coef)[-1]

aml_test <- read.csv(here("data", "AMLtestdata.csv"))
aml_test_imputed <- complete(mice::mice(dplyr::select(aml_test, SEX:CD19), m = 1), 1)
X_test <- model.matrix(reformulate(termlabels = pvars), data = aml_test_imputed)

lp_test <- X_test %*% betas
p_test <- 1 / (1 + exp(-lp_test))
class_test <- p_test > .5

```

Export predictions

```{r}
preddat <- data.frame(
  X.Patient_id = aml_test$X.Patient_id,
  p_complete_response = p_test
)
save(preddat, file = here("data", "predicted group 1.Rdata"))
```



Some glm testing for colleges

```{r}
fit_glmnet <- glmnet(X_clin, y)
X_test_2 <- as.matrix(dplyr::select(aml_test_imputed, - Chemo.Simplest) %>%
                 mutate_all(as.numeric))
predict(fit_glmnet, newx = X_test)
```

Check with actual labels

```{r}
aml_full <- read.csv(here("data", "AML_training.csv"))
act_labels <- aml_full %>%
  filter(X.Patient_id %in% aml_test$X.Patient_id) %>%
  select(X.Patient_id, resp.simple) 
```

```{r}
table(class_test, act_labels$resp.simple)
(58 + 8) / (10 + 8 + 58 + 19)
prop.table(table(act_labels$resp.simple))
```

```{r}
y_test <- as.numeric(as.character(factor(act_labels$resp.simple, levels = c("CR", "RESISTANT"), labels = c(1, 0))))
val.prob(logit = lp_test, y = y_test)
```

