---
title: "Survival analysis"
author: "Wouter van Amsterdam"
date: 2018-01-15
output: html_document
---

<!-- The file analysis/chunks.R contains chunks that define default settings
shared across the workflowr files. -->
```{r read-chunk, include=FALSE, cache=FALSE}
knitr::read_chunk("chunks.R")
```

<!-- Update knitr chunk options -->
```{r knitr-opts-chunk, include=FALSE}
```

<!-- Insert the date the file was last updated -->
```{r last-updated, echo=FALSE, results='asis'}
```

<!-- Insert the code version (Git commit SHA1) if Git repository exists and R
 package git2r is installed -->
```{r code-version, echo=FALSE, results='asis'}
```

<!-- Add your analysis here -->

## Introduction

Tutor: Rebecca Stellato

## Survival analysis

Data: events + time + censoring 

Time from well defined starting point, almost always right-skewed

### Formalism

#### Survival function
$F(t)$ is the cumulative probability of dying at time-point t for a participant. 
Monotonically increasing, bounded at 1

$S(t) = 1 - F(t)$

Can be done parametric (exponential or Weibull). 

#### Kaplan-Meier

Usually: Kaplan-Meier, which is unparametric.

For each time interval $j$

* $n_j$: number of patients entering interval
* $d_j$: number of events in interval

Probability of surviving at time $t$ with $t_k < t < t_{k+1}$

$$S(t) = \prod_j^k{\frac{n_j-d_j}{n_j}}$$


#### Testing

Log-rank test


#### Hazard fuction

Hazard rate is instantatinious rate of an event

$$\lambda(t) = -S'(t)/S(t)$$

Culmulative hazard

### Cox proportional hazard regression

Assumption

* Everyone has the same unknown baseline hazard rate $\lambda_0(t)$
* individual hazard rate is by mutiplying the baseline with exponential function

$$\lambda_i(t) = \lambda_0(t)e^{\beta_1x_1+\beta_2x_2...}$$

Then $e^{\beta_j}$ is the **hazard ratio**

Baseline intercept incorporated in $\lambda_0$

$$h_{i\ vs\ j} = e^{B(X_i - X_j)}$$

**assumption** this quantity does not depend on $t$

This is what proportional hazard means

#### Implementation in R

Uses Wald-test

#### Assumptions

Cox proportional hazards is semi-parametric

No assumption on baseline hazard distribution. 

Use partial likelihood (without baseline hazard) by evaluating hazard funcation.

If event happend to subject $i$ at time $Y_i$, the probability 
that this happend is:

$$L_i(\beta) = \frac{\lambda(Y_i|X_i)}{\sum_{j:Y_j\geq Y_i}{\lambda{Y_i|X_j}}}$$



Confidence based on Wald, Score, or likelihood ratio-test

#### Stratified Cox model

$$\lambda_i(t) = \lambda_g(t)e^{\beta_1x_1+\beta_2x_2...}$$

Where $g$ the groups

Then groups no longer in linear predictor.
$\beta_i$ assumed to be equal in both groups.

##### Check proportional hazards assumptions for categorical variables

* Stratify by group
* Plot LML (log minus log = log (cumulative))
* if two (cumulative) hazards are proportional, 
their logs will be parallel and differ by a constant

As long as they don't cross, it is OK (except when you have a lot of data)

If you don't STRATIFY, you'll get out what you put in 
(propartianality), so allways use stratified analysis 
(with `strata` in R).

If this is OK, we can use it in the unstratified model

Otherwise: you can only do stratified analysis. 
You can create sort of survival times based on both models 
and compare.

Or some interaction with time.

#### Checking other assumptions

Continous variables: unable to do visually

`coxzph` for statistical tests

#### Variations for 2 variable problems

* g = group
* x = continous variable

  1. baseline same for g, effect of x same in both group
  2. baseline same for g, effect of x different (interaction)
  3. baseline different for g, effect of x same in groups (strata)
  4. baseline different for g, effect of x different
    * can do 2 models
    * can do with strata `coxph(surv ~ x * strata(g))`

Compare models with likelihood ratio test or AIC

### Predict survival from Cox proportional hazards

* go from hazard to cumulative hazard
* use $H(t) = -\ln(S(t)) \iff S(t) = e^{-H(t)}$
* gives $S_i(t) = S_0(t)^{e^{\beta_1x_1+...}}$

So if baseline survival is $p$, than for patient with hazard ratio $HR$ 
is $p^{HR}$. Baseline assumes all covariate values are 0. 

With R, predictions are given for 'mean' values when not given (also: 
'mean' treatment which is like half treated)

R can give predicted survival curves for any levels of covariates.

## With data

Gene expression analysis for progression-free survival

```{r}
data(nki70, package = "penalized")
str(nki70)
```

```{r}
library(survival)

fit <- coxph(Surv(time, event)~Diam + Age, data = nki70)
fit2 <- coxph(Surv(time, event)~Diam * Age, data = nki70)
summary(fit)
summary(fit2)
broom::tidy(fit)

AIC(fit, fit2)
anova(fit, fit2)
```


## Session information

<!-- Insert the session information into the document -->
```{r session-info}
```
