---
title: "Parial plot example"
author: "Wouter van Amsterdam"
date: 2018-03-10
output: html_document
---

<!-- The file analysis/chunks.R contains chunks that define default settings
shared across the workflowr files. -->
```{r read-chunk, include=FALSE, cache=FALSE}
knitr::read_chunk("chunks.R")
```

<!-- Update knitr chunk options -->
```{r knitr-opts-chunk, include=FALSE}
```

<!-- Insert the date the file was last updated -->
```{r last-updated, echo=FALSE, results='asis'}
```

<!-- Insert the code version (Git commit SHA1) if Git repository exists and R
 package git2r is installed -->
```{r code-version, echo=FALSE, results='asis'}
```

<!-- Add your analysis here -->

# Setup

```{r}
library(dplyr)
library(data.table)
library(magrittr)
library(purrr)
library(here) # for tracking working directory
library(ggplot2)
library(epistats)
library(broom)

```


# Introduction

Partial plots can be used to check assumptions of a model

```{r}
data("mtcars")
```

## Linear model data

Correct linear assumptions

```{r}
set.seed(12345)
nobs = 1000
npred = 2

corr_x1x2 = 0.
x1 = rnorm(nobs)
x2 = corr_x1x2 * x1 + (1-corr_x1x2)*rnorm(nobs)
x = as.matrix(cbind(x1, x2))
betas = c(1, 1)
resid_noise = .3

y = x %*%betas + rnorm(nobs, sd = resid_noise)
```

Show marginal plots

```{r}
plot(y, x1)
plot(y, x2)

```

Seem perfectly linear

Create linear model

```{r}
fit <- lm(y~x1 + x2)
plot(fit)
```

Create partial plots

```{r}
partial_plots.lm(fit)
```


Now if x1 and x2 are dependent


```{r}
require(effects)
set.seed(12345)
nobs = 1000
npred = 2

corr_x1x2 = 0.5
x1 = rnorm(nobs)
x2 = corr_x1x2 * x1 + (1-corr_x1x2)*rnorm(nobs)
x = as.matrix(cbind(x1, x2))
betas = c(1, 1)
resid_noise = .3

y = x %*%betas + rnorm(nobs, sd = resid_noise)

fit <- lm(y~x1+x2)
# part2 <- partial_residuals.lm(fit, term = "x1")

# plot(x1, residuals(fit, type = "partial")[,1])
par(mfrow = c(1,2)); plot(y, x1); plot(y, x2); par(mfrow = c(1,1))
partial_plots.lm(fit)
# plot(Effect(focal.predictors = c("x1", "x2"), mod = fit))
```


```{r}
set.seed(12345)
nobs = 1000
npred = 2

corr_x1x2 = 0.8
x1 = runif(nobs, min = 0, max = 2)
x2 = exp(x1) + .0001*rnorm(nobs)
x = as.matrix(cbind(x1, x2))
betas = c(1, 1)
resid_noise = .00001

y = (x1^2)*(x2) + rnorm(nobs, sd = resid_noise)

par(mfrow = c(1,2))
plot(x1, y)
plot(x2, y)
par(mfrow = c(1,1))

fit <- lm(y~x1+x2)
partial_plots.lm(fit)
```

Now with a different relationship between x1 and x2

```{r}
set.seed(12345)
nobs = 1000
npred = 2

a = 5; b = 2; gamma = 1
x1 = rnorm(nobs)
x2 = a * x1^3 - b * x1^2 + gamma * x1 + .1*rnorm(nobs)
x = as.matrix(cbind(x1, x2))
betas = c(1, 1)
resid_noise = .3

y = x %*% betas + rnorm(nobs, sd = resid_noise)

fit <- lm(y~x1+x2)
coef(fit)
plot(fit, which = 2)
partial_plots.lm(fit)
```

Now with a log relationship between x1 and x2

```{r}
set.seed(12345)
nobs = 1000
npred = 2

x1 = rnorm(nobs)
x2 = log(x1) + .1*rnorm(nobs)
x = as.matrix(cbind(x1, x2))
betas = c(1, 1)
resid_noise = .3

y = x %*% betas + rnorm(nobs, sd = resid_noise)

fit <- lm(y~x1+x2)
coef(fit)
plot(fit, which = 2)
partial_plots.lm(fit)
```

Now with a sine relationship between x1 and x2

```{r}
set.seed(12345)
nobs = 1000
npred = 2

x1 = runif(nobs)
x2 = x1 + sin(x1*2*2*pi) + .2*rnorm(nobs)
x = as.matrix(cbind(x1, x2))
betas = c(1, 1)
resid_noise = .3
plot(x1, x2)

y = x %*% betas + rnorm(nobs, sd = resid_noise)

fit <- lm(y~x1+x2)
coef(fit)
plot(fit, which = 2)
partial_plots.lm(fit)
```

## Logistic regression

Simulate data

```{r}
set.seed(12345)
nobs = 1000
npred = 2

x1 = runif(nobs)
x2 = runif(nobs)
x3 = runif(nobs)
x = as.matrix(cbind(x1, x2))
betas = c(1, 1, 1)
resid_noise = .3
plot(x1, x2)

eta = x1 + x2^2 + x3^3 + rnorm(nobs, sd = resid_noise)
probs = 1 / (1 + exp(-eta))
y = map_dbl(probs, function (p) sample(0:1, prob = c(1-p, p), size = 1))

fit <- glm(y~x1+x2+x3, family = binomial(link = "logit"))
coef(fit)
plot(fit, which = 2)
partial_plots.lm(fit)
```


```{r}
partial_residuals.glm <- function(fit, term, type = "response") {
  formula0  = formula(fit)
  all_vars  = all.vars(formula0)
  response  = all_vars[1]
  all_terms = all_vars[-1]
  new_terms = setdiff(all_terms, term)

  fit_resp <- glm(reformulate(new_terms, response), data = fit$model,
                  family = binomial(link = "logit"))
  resp_logits = log(1 / (1 - residuals(fit_resp, type = "response")))
  fit_term <- lm(reformulate(new_terms, term), data = fit$model)

  return(
    # data.frame(resid_response = resid(fit_resp, type = type),
    #            resid_term     = resid(fit_term, type = type))
    data.frame(resid_response = resp_logits,
               resid_term     = resid(fit_term, type = type))
  )
}


partial_plots.glm <- function(fit, terms = NULL, what = "resid") {
  formula0  = formula(fit)
  all_vars  = all.vars(formula0)
  response  = all_vars[1]
  all_terms = all_vars[-1]

  terms = if (!is.null(terms)) {terms} else {all_terms}

  resid_data = pmap_df(list(terms), function(term) {
    data.frame(term = term,
               partial_residuals.glm(fit, term), stringsAsFactors = F)
  })

  p = ggplot(resid_data, aes(x = resid_term, y = resid_response)) +
    geom_point() + geom_smooth(method = "lm", alpha = 0.15) +
    facet_wrap(~term, scales = "free_x") +
    theme_minimal() +
    labs(x = "Residual of partial model term ~ .",
         y = paste0("Residual of partial model ", response, " ~ ."))

  if (what == "resid") {
    print(p)
    return(resid_data)
  } else if (what == "plot") {
    return(p)
  }
}
partial_plots.glm(fit)
```


```{r}
require(effects)
plot(predictorEffect(mod = fit, predictor = "x2"))
```


```{r}
plot(resid(lm(x1~x2+x3)), residuals(fit, type = "partial")[,1])
plot(x1, residuals(fit, type = "partial")[,1])
```


## Session information

<!-- Insert the session information into the document -->
```{r session-info}
```
