---
title: "Simulation of the Boltzmann distribution"
author: "Wouter van Amsterdam"
date: 2017-12-04
output: html_document
---

<!-- The file analysis/chunks.R contains chunks that define default settings
shared across the workflowr files. -->
```{r read-chunk, include=FALSE, cache=FALSE}
knitr::read_chunk("chunks.R")
```

<!-- Update knitr chunk options -->
```{r knitr-opts-chunk, include=FALSE}
```

<!-- Insert the date the file was last updated -->
```{r last-updated, echo=FALSE, results='asis'}
```

<!-- Insert the code version (Git commit SHA1) if Git repository exists and R
 package git2r is installed -->
```{r code-version, echo=FALSE, results='asis'}
```

<!-- Add your analysis here -->
## Introduction
This simulation is inspired by the proposed simulation from the book 
'Concepts in Thermal Physics' by Blundell and Blundell


The Boltzmann distribution is given by 

$$P(microstate\ r) = \frac{e^{-E_r/k_{b}T}}{\sum_i{e^{-E_i/k_{b}T}}}$$

Where $k_b$ the Boltzmann constant: $k_b = 1.3807*10^{-23}JK^{-1}$

Here the denominator is known as the partition function $Z$.

This distrubtion describes the probability of finding a system 
that is in thermal equilibrium with a large resevoir in a certain microstate,
with an associated energy, which is assumed to be unique for a certain state.

So $E_{r_1} = E_{r_2}$ means that $r_1 = r_2$.

## Simulations
Think of a particular system with particles that each can have discrete energy 
states. The system in is thermal isolation, so no energy can be added or 
withdrawn. Each particle can exchange energy with another particle. 

For visualization, view the particles are laid out on a 2x2 grid.

```{r}
npart = 400
dims  = rep(sqrt(npart), 2)

init_energy = 1

particles_init <- rep(init_energy, npart)
particles <- particles_init

image(matrix(particles_init, nrow = dims[1]))
```

Visualize distribution of initial energy

```{r}
barplot(table(particles))
```

Now randomly pick a particle that gives away an energy quantum to another particle.

```{r}
set.seed(2)
giving_particle <- sample(1:npart, size = 1)
receiving_particle <- sample(1:npart, size = 1)

particles[giving_particle] <- particles[giving_particle]-1
particles[receiving_particle] <- particles[receiving_particle]+1

image(matrix(particles, nrow = dims[1]))

barplot(table(particles))
```

Now repeat this process 10 times, making sure that no particles get negative energy 
(which is physically unacceptable).


```{r}
ntimes = 10

for (i in 1:ntimes) {
  giving_particle <- sample(which(particles > 0), size = 1)
  receiving_particle <- sample(1:npart, size = 1)
  
  particles[giving_particle] <- particles[giving_particle]-1
  particles[receiving_particle] <- particles[receiving_particle]+1
}

image(matrix(particles, nrow = dims[1]))
  
barplot(table(particles))

```


And 100 times

```{r}
ntimes = 100

for (i in 1:ntimes) {
  giving_particle <- sample(which(particles > 0), size = 1)
  receiving_particle <- sample(1:npart, size = 1)
  
  particles[giving_particle] <- particles[giving_particle]-1
  particles[receiving_particle] <- particles[receiving_particle]+1
}

image(matrix(particles, nrow = dims[1]))
  
barplot(table(particles))

```


And 1000 times

```{r}
ntimes = 1000

for (i in 1:ntimes) {
  giving_particle <- sample(which(particles > 0), size = 1)
  receiving_particle <- sample(1:npart, size = 1)
  
  particles[giving_particle] <- particles[giving_particle]-1
  particles[receiving_particle] <- particles[receiving_particle]+1
}

image(matrix(particles, nrow = dims[1]))
  
barplot(table(particles))

```

And a million times (this takes some time to compute)

```{r}
ntimes = 1e6

for (i in 1:ntimes) {
  giving_particle <- sample(which(particles > 0), size = 1)
  receiving_particle <- sample(1:npart, size = 1)
  
  particles[giving_particle] <- particles[giving_particle]-1
  particles[receiving_particle] <- particles[receiving_particle]+1
}

image(matrix(particles, nrow = dims[1]))
  
barplot(table(particles))

```

The distribution now very closely resembles the Boltzmann distribution

## Session information

<!-- Insert the session information into the document -->
```{r session-info}
```
