---
title: "Summary of statistical tests and distributions"
author: "Wouter van Amsterdam"
date: 2017-11-13
output: html_document
---

<!-- The file analysis/chunks.R contains chunks that define default settings
shared across the workflowr files. -->
```{r read-chunk, include=FALSE, cache=FALSE}
knitr::read_chunk("chunks.R")
```

<!-- Update knitr chunk options -->
```{r knitr-opts-chunk, include=FALSE}
```

<!-- Insert the date the file was last updated -->
```{r last-updated, echo=FALSE, results='asis'}
```

<!-- Insert the code version (Git commit SHA1) if Git repository exists and R
 package git2r is installed -->
```{r code-version, echo=FALSE, results='asis'}
```

<!-- Add your analysis here -->
## Statistical tests
### Single Sample T-test
### Two paired samples T-test
See single sample T-test for $d = X_1 - X_2$

### Two unpaired samples T-test, equal variance
#### Other names
#### Definition
##### Test-statistic
t, follows Student's T-distribution with n1+n2-2 degrees of freedom

#### Variables
###### Types
interval, continous

###### Distributions
normal

#### Relevant equations
##### Test-statistic
$$t = \frac{\bar{Y_{1}} - \bar{Y_{2}}}{\bar{{SE}}_{difference}}$$

##### Standard error
$${SE}_{difference} = s_{pooled}*\sqrt{\frac{1}{n_{1}}+\frac{1}{n_{2}}}$$

##### Pooled variance
$$s_{pooled} = \frac{(n_{1}-1)*s_{1}^2 + (n_{2}-1)*s_{2}^2}{n_{1}+n_{2}-2}$$

#### Assumptions
$X_1$, $X_2$ normally distributed, or CTL holds. Equal variance or non-equal variance.

### Two unpaired samples T-test, unequal variance
#### Other names
#### Definition
##### Test-statistic
t, follows Student's T-distribution with df degrees of freedom

#### Variables
##### Dependent
###### Types
interval, continous

###### Distributions
normal

##### Independent
###### Types
interval, continous

###### Distributions
normal

#### Relevant equations
##### Test-statistic
$$t = \frac{\bar{Y_{1}} - \bar{Y_{2}}}{\bar{{SE}}_{difference}}$$

##### Standard error
$${SE}_{difference} = \sqrt{\frac{s_1^2}{n_{1}}+\frac{s_2^2}{n_{2}}}$$

##### Degrees of freedom, Welch-Satterthwaite
$${df} = \frac{(\frac{s_{1}^2}{n_{1}} + \frac{s_{2}^2}{n_{2}})^2}{({s_{1}^2}/{n_{1}})^2/(n_{1}-1) + ({s_{2}^2}/{n_{2}})^2/(n_{2}-1)}$$


#### Assumptions
$X_1$, $X_2$ normally distributed, or CTL holds. Equal variance or non-equal variance.





## Session information

<!-- Insert the session information into the document -->
```{r session-info}
```
